---
title: "Practical 2: Data Visualisation"
author: "Nina van Gerwen (1860852)"
date: "2022-09-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library(ISLR)
library(tidyverse)
```

### Question 1

Aesthetics: aes(x = Hits, y = HmRun)
Geoms:
1) geom_point() for the scatterplot
2) geom_density_2d() for the weird circles
Scales: xlab = "Hits, ylab = "Home runs", labs(title = "...")
Themes: theme_minimal()

As for statistical transformations, I believe geom_density_2d() does this
as it performs a 2D kernel density estimation.

### Question 2

```{r}
set.seed(1234)
student_grade  <- rnorm(32, 7)
student_number <- round(runif(32) * 2e6 + 5e6)
programme      <- sample(c("Science", "Social Science"), 32, replace = TRUE)

gg_students <- data.frame(student_number, student_grade, programme)
```


### Question 3

```{r}
rev_homeruns_plot <- ggplot(data = Hitters, aes(x = HmRun, y = Hits)) +
  geom_point() +
  labs(x = "Home runs", y = "Hits") +
  geom_density_2d() +
  labs(title = "Cool density and scatter plot of baseball data") +
  theme_minimal()

rev_homeruns_plot
```

### Question 4

```{r}
upgraded_homeruns_plot <- rev_homeruns_plot + aes(x = HmRun, y = Hits,
                                                  col = League,
                                                  size = Salary)

upgraded_homeruns_plot
```

### Question 5

There are indeed many geoms.


### Question 6

```{r}
student_hist <- ggplot(data = gg_students, aes(x = student_grade)) +
  geom_histogram(binwidth = 0.15) +
  theme_minimal()

student_hist
```

### Question 7

```{r}
student_dens <- ggplot(data = gg_students, aes(x = student_grade)) +
  geom_density(fill = "light seagreen", col = NA)

student_dens
```
A very pretty colour indeed.


### Question 8

```{r}
upgr_student_dens <- student_dens +
  geom_rug(col = "red", size = 1)

upgr_student_dens
```


### Question 9

```{r}
upgr_student_dens +
  xlim(0, 10) +
  labs(y = "", title = "Density plot of student grades") +
  theme_minimal()
```
I had already removed the border in the first graph.


### Question 10

```{r}
ggplot(data = gg_students, aes(y = student_grade, x = programme,
                                                  fill = programme)) +
  geom_boxplot() 
```

### Question 11

The somewhat thicker horizontal line is the mean of the group, the other two
horizontal lines are the first and third quartile of the
students grade in each programme. The vertical whiskers are the range of
grades in each programme.

### Question 12

```{r}
ggplot(data = gg_students, aes(x = student_grade, fill = programme)) +
  geom_density(col = NA, alpha = .5)

```

### Question 13

```{r}
Hitters %>%
  ggplot(data = ., aes(x = Years)) +
  geom_bar()
```

### Question 14

```{r}
line_plot <- ggplot(data = Smarket[1:200 ,], aes(x = 1:200, y = Volume)) +
  geom_line() +
  labs(x = "Days")

line_plot
```

### Question 15

```{r}
line_plot_2 <- line_plot +
  geom_line(col = "purple", size = 1) +
  geom_point(col = "purple", size = 2)

line_plot_2
```
I really like purple.

### Question 16

```{r}
which.max(Smarket$Volume[1:200])
max(Smarket$Volume[1:200])
```
So the 170th day had the highest trade volume of 2.33.


### Question 17

```{r}
line_plot_2 +
  geom_label(aes(x = 170, y = 2.33083, label = "Peak volume"))
```

### Question 18

```{r}
library(magrittr)
baseball <- Hitters %>%
  filter(., !is.na(Salary)) %>%
  mutate(Cat_Sal = cut(x = Salary, breaks = 3, labels = c("Well-Off", 
                                                          "Rich", 
                                                          "Super Rich")),
         Prop_HmRun = CHmRun/CHits)
```


### Question 19

```{r}
baseball_plot <- ggplot(data = baseball, aes(x = CWalks, y = Prop_HmRun)) +
  geom_point() +
  ylim(0, 0.4) + xlim(0, 1600) +
  labs(x = "Number of walks during career", 
       y = "Proportion of home runs during career hits")

baseball_plot
```

### Question 20

```{r}
baseball_plot + facet_wrap(vars(baseball$Cat_Sal))
```

### Question 21

```{r}
Carseats %>%
  ggplot(data = ., aes(x = Income, y = Sales, col = Urban)) +
  geom_point() +
  facet_wrap(vars(Carseats$US)) +
  geom_smooth()
  
```
Tadaaa! A beautiful plot that shows the relation between community income
and car sales while taking into account whether the area is urban or not and
whether the shop is in the USA. There does not seem to be a relation between
the two variables. Noticeable is however that there seem more car sales in the US.
This could possibly be explained by to the fact that it is a large country with terrible
public transportation. Hence, everyone needs to have a car.
