---
title: "W1: Data Wrangling"
author: "Alex Carriero"
date: "12/09/2022"
output: pdf_document
---
```{r}
# Class Notes: 
# TDGM - True Data Generating Mechanism 
# TDGM exist, but nobody has ever encountered it 

# Recall frequentest definition of CI 
```

```{r}
# Libraries
library(ISLR)
library(tidyverse)
library(haven)
library(readxl)
```

1. Run the following code in R and inspect their data types using the class() function. Try to guess beforehand what their types will be!
```{r}
object_1 <- 1:5
object_2 <- 1L:5L
object_3 <- "-123.456"
object_4 <- as.numeric(object_2)
object_5 <- letters[object_1]
object_6 <- as.factor(rep(object_5, 2))
object_7 <- c(1, 2, 3, "4", "5", "6")
```

2. Convert object_7 back to a vector of numbers using the as.numeric() function
```{r}
object_7 <- as.numeric(object_7)
class(object_7)
```

3. Make a list called objects containing object 1 to 7 using the list() function.
```{r}
my_list <- list(object_1, object_2, object_3, object_4, object_5, object_6, object_7)
my_list[1]
```

4. Make a data frame out of object_1, object_2, and object_5 using the data.frame() function. 
```{r}
my_df <- data.frame(object_1, object_2, object_5)
my_df
```

5. Useful functions for determining the size of a data frame are ncol() and nrow(). Try them out!
```{r}
ncol(my_df)
nrow(my_df)
```


6. Use the function read_csv() to import the file “data/googleplaystore.csv” and store it in a variable called apps.
```{r}
apps <- read_csv("googleplaystore.csv")
```

7. Did any column get a variable type you did not expect?


8. Use the function head() to look at the first few rows of the apps dataset. 
```{r}
head(apps)
```

9. Repeat steps 5, 6, and 7 but now for “data/students.xlsx” (NB: You’ll need a function from the package readxl). Also try out the function tail() and View() (with a capital V).

10. Create a summary of the three columns in the students dataset using the summary() function. What is the range of the grades achieved by the students?

