---
title: "Analysis of Alcohol consumption among students"
author: "Hasbini Laura"
date: "20 septembre 2022"
output: html_document
---

# Introduction

For this assigment we will look at the following variables for one school only
- Sex
- Age
- Address
- Pstatus
- Medu
- Pedu
- studytime
- failures
- romantic
- famrel
- freetime
- goout
- Dalc
- Walc
- health
- Absences
- G3


The practicals always start with the packages we are going to use. Be sure to run these lines in your session to load their functions before you continue. If there are packages that you have not yet installed, first install them with `install.packages()`.

```{r load_packages, message = FALSE, warning = FALSE}
library(ISLR)
library(tidyverse)
library(haven)
library(readxl)
library(reshape2)
library(dplyr)
```


# Loading the data

```{r readcsv, include = params$answers}
mat_data <- read_csv("data/student-mat.csv")
por_data <- read_csv("data/student-por.csv")
```

```{r head, include = params$answers}
summary(mat_data)
```

```{r head, include = params$answers}
summary(por_data)
```

#Tidyng the data

```{r}
mat_data[mat_data == "no"] <- "FALSE"
mat_data[mat_data == "yes"] <- "TRUE"
por_data[por_data == "no"] <- "FALSE"
por_data[por_data == "yes"] <- "TRUE"
```


```{r}
por_data %>% 
  mutate(school = parse_factor(school), sex = parse_factor(sex), address =
  parse_factor(address), famsize = parse_factor(famsize), Pstatus = parse_factor(Pstatus), Mjob = parse_factor(Mjob), Fjob = parse_factor(Fjob), reason = parse_factor(reason), guardian = parse_factor(guardian), famsup = parse_logical(famsup), paid = parse_logical(paid), activities = parse_logical(activities), nursery = parse_logical(nursery), higher = parse_logical(higher), internet = parse_logical(internet), romantic = parse_logical(romantic))
```

```{r}
mat_data %>% 
  mutate(school = parse_factor(school), sex = parse_factor(sex), address =
    parse_factor(address), famsize = parse_factor(famsize), Pstatus =
    parse_factor(Pstatus), Mjob = parse_factor(Mjob), Fjob = parse_factor(Fjob),
    reason = parse_factor(reason), guardian = parse_factor(guardian), famsup =
    parse_logical(famsup), paid = parse_logical(paid), activities =
    parse_logical(activities), nursery = parse_logical(nursery), higher =
    parse_logical(higher), internet = parse_logical(internet), romantic =
    parse_logical(romantic))
```
```{r}
mat_data <- 
  select(mat_data, -school, -famsize, -Mjob, -Fjob, -reason, -guardian, -traveltime, -schoolsup, -famsup, -paid, -activities, -nursery, -higher, -internet, -G1, -G2)

por_data <-
  select(por_data, -school, -famsize, -Mjob, -Fjob, -reason, -guardian, -traveltime, -schoolsup, -famsup, -paid, -activities, -nursery, -higher, -internet,  -G1, -G2)
```

```{r}
summary(por_data)
```

Creating a new variable liked to total consumption
```{r}
mat_data <- mutate(mat_data, tot_alc = rowMeans(mat_data[c("Dalc", "Walc")]))
por_data <- mutate(por_data, tot_alc = rowMeans(por_data[c("Dalc", "Walc")]))
```


## Summary statistics tables



## Interesting plots

```{r}
por_data_alc <- por_data
```

```{r}
por_data %>%
  select(Dalc, Walc, tot_alc, sex) %>%
  melt(id='sex') %>%
  ggplot(aes(x = variable, y = value, color = sex)) +  # ggplot function
  geom_boxplot()
```


```{r}
por_data %>%
  ggplot(aes(x=Walc, fill=sex)) +
  geom_bar(alpha=0.5) +
  theme_minimal() +
  labs(y="", fill="Week-End Consumption")

```

```{r}
test <- por_data %>%
  select(tot_alc, absences) %>%
  group_by(absences) %>%
  summarise_at(vars(tot_alc), list(name = mean))
```


```{r}
por_data %>%
  select(tot_alc, absences) %>%
  group_by(absences) %>%
  summarise_at(vars(tot_alc), list(name = mean)) %>%
  ggplot(aes(x=absences, y=name)) +
  geom_bar(stat="identity") +
  theme_minimal() +
  labs(y="Alcohol consumption", fill="Absences")

```


