---
title: "Assignment1 - RHAT"
output: html_document
date: "September 2022"
---

# NYC Rat Sightings

## Data Description

The NYC Rat Sightings dataset describes mostly the location where there are rat sightings, the dates when it happened and which type of building it happened on. In the original dataset there are also many empty columns or columns with only one value.

We will analyse if sightings depend on the month they happen. Additionally, we will analyse a dataset that describes borough populations so we can analyse if population density compares to number of rat sightings and which boroughs have more density of rat sightings. 

The dataset we will be using can be found on [kaggle](https://www.kaggle.com/datasets/new-york-city/nyc-rat-sightings).

```{r}
library(tidyverse)
```

## Data cleaning

First, we create a function to use for splitting up the different dates.

```{r}
date.split <- function(txtdate, x){
  date <- parse_date(txtdate, format = "%m/%d/%Y %I:%M:%S %p")
  time <- parse_time(txtdate, format = "%m/%d/%Y %I:%M:%S %p")
  year <- as.numeric(format(date, format = "%Y"))
  month <- as.numeric(format(date, format = "%m"))
  day <- as.numeric(format(date, format = "%d"))
  
  if (x == 1) {value <- year}
  else if (x == 2) {value <- month}
  else if (x == 3) {value <- day}
  else if (x == 4) {value <- time}
  
  return(value)
}
```


Secondly, we create a function to use when splitting a specific number from a data point, as used for both the address and the community boards.

```{r}
get.number <- function(string1, string2){
  
  split_1 <- strsplit(string1, split = " ")
  split_2 <- strsplit(string2, split = " ")
  
  unique_values <- Map(setdiff, split_1, split_2)
  to_be_returned <- unlist(unique_values)
  return(to_be_returned)
}
```




```{r}
#Second attempt at a function that could split "Incident_Address":
get.number2 <- function(address){
  if(!identical(address, NA)){
    nums <- str_extract(address, "[0-9]+")
    return(nums)
  }
}
```

The dataset initially looks like this:

```{r}
rat_dataset <-
  read_csv("Rat_Sightings.csv", show_col_types = FALSE)
rat_dataset
```

with many empty features or with one single value. In this next step, we will be taking these features out and separating some interesting data in more features:

```{r}
rat_dataset <-
  read_csv("Rat_Sightings.csv", show_col_types = FALSE) %>% 
  #select("Unique Key", "Created Date", "Closed Date", "Location Type", "Incident Zip", "Incident Address", "Street Name", "Cross Street 1", "Cross Street 2", "Intersection Street 1", "Intersection Street 2", "Address Type", "Status", "Due Date", "Resolution Action Updated Date", "Community Board", "Borough", "Latitude", "Longitude") %>% 
  rename(Unique_Key = "Unique Key", Created_Date = "Created Date", Closed_Date = "Closed Date", Location_Type = "Location Type", Incident_Zip = "Incident Zip", Incident_Address = "Incident Address", Street_Name = "Street Name", Cross_Street_1 = "Cross Street 1", Cross_Street_2 = "Cross Street 2", Intersection_Street_1 = "Intersection Street 1", Intersection_Street_2 = "Intersection Street 2", Address_Type = "Address Type", Status = "Status", Due_Date = "Due Date", RAU_Date = "Resolution Action Updated Date", Community_Board = "Community Board", Borough = "Borough", Latitude = "Latitude", Longitude = "Longitude") %>% 
  mutate(Created_Year = date.split(Created_Date, 1), Created_Month = date.split(Created_Date, 2), Created_Day = date.split(Created_Date, 3)) %>% 
  mutate(Closed_Year = date.split(Closed_Date, 1), Closed_Month = date.split(Closed_Date, 2), Closed_Day = date.split(Closed_Date, 3)) %>%
  mutate(Due_Year = date.split(Due_Date, 1), Due_Month = date.split(Due_Date, 2), Due_Day = date.split(Due_Date, 3), Due_Time = date.split(Due_Date, 4)) %>%
  mutate(RAU_Year = date.split(RAU_Date, 1), RAU_Month = date.split(RAU_Date, 2), RAU_Day = date.split(RAU_Date, 3), RAU_Time = date.split(RAU_Date, 4)) %>% 
  ##These two lines handled the NAs in both columns, specifically to fit the get.number function
  #mutate(Incident_Address = replace_na(Incident_Address, "None")) %>% 
  #mutate(Street_Name = replace_na(Street_Name, "None")) %>% 
  mutate(Address_Number = get.number2(Incident_Address)) %>% 
  mutate(Community_Board_Number = get.number(Community_Board, Borough)) %>% 
  select(Unique_Key, Created_Year, Created_Month, Created_Day, Closed_Year, Closed_Month, Closed_Day, Location_Type, Incident_Zip, Address_Number, Street_Name, Cross_Street_1, Cross_Street_2, Intersection_Street_1, Intersection_Street_2, Address_Type, Status, Due_Year, Due_Month, Due_Day, Due_Time, RAU_Year, RAU_Month, RAU_Day, RAU_Time, Community_Board_Number, Borough, Latitude, Longitude)
```

We also want to make sure that all of the features have the proper type.

```{r}
rat_dataset$Location_Type <- 
  as.factor(rat_dataset$Location_Type)
rat_dataset$Street_Name <-
  as.factor(rat_dataset$Street_Name)
rat_dataset$Cross_Street_1 <-
  as.factor(rat_dataset$Cross_Street_1)
rat_dataset$Cross_Street_2 <-
  as.factor(rat_dataset$Cross_Street_2)
rat_dataset$Address_Type <-
  as.factor(rat_dataset$Address_Type)
rat_dataset$Status <-
  as.factor(rat_dataset$Status)
rat_dataset$Borough <-
  as.factor(rat_dataset$Borough)
```


So, our starting point data is:

```{r}
rat_dataset
```

