---
title: "Practical 3"
author: "Judith Neve"
date: '2022-10-01'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exploratory Data Analysis

## Introduction

```{r}
library(tidyverse)
library(magrittr)
library(mice)
library(DAAG)
```

## The boys data

```{r}
head(boys)
tail(boys)
```

```{r}
!is.unsorted(boys$age)
```

```{r}
summary(boys)
```

### 1. Create a histogram of the variable age using the function geom_histogram()

```{r}
boys %>%
  ggplot(aes(x = age)) +
  geom_histogram()
```

### 2. Create a bar chart of the variable gen using the function geom_bar()

```{r}
boys %>%
  ggplot(aes(x = gen)) +
  geom_bar()
```

## Assessing missing data

```{r}
md.pattern(boys)
```

### 3. Create a missingness indicator for the variables gen, phb and tv

```{r}
boys_missing <- boys %>% 
  mutate(gen_missing = is.na(gen),
         phb_missing = is.na(phb),
         tv_missing = is.na(tv))

head(boys_missing)
```

### 4. Assess whether missingness in the variabels gen, phb and tv is related to someone's age.

```{r}
boys_missing %>%
  group_by(gen_missing, phb_missing, tv_missing) %>% 
  summarise(age = mean(age))
# none missing are on average oldest, all missing are on average youngest
```

### 5. Create a histogram for the variable age, faceted by whether or not someone has a missing value on gen.

```{r}
boys_missing %>% 
  ggplot(aes(x = age)) +
  geom_histogram() +
  facet_wrap(~ gen_missing)
```

### 6. Create a scatterplot with age on the x-axis and bmi on the y-axis, using the function geom_point().

```{r}
boys_missing %>%
  ggplot(aes(x = age, y = bmi)) +
  geom_point()
```

### 7. Ass a colour aesthetic to the previous plot using the missingness indicator of the variable gen.

```{r}
boys_missing %>%
  ggplot(aes(x = age, y = bmi, col = gen_missing)) +
  geom_point()
```

## Visualising the boys data

### 8. Visualise the relationship between reg and age using a boxplot

```{r}
boys %>% 
  ggplot(aes(x = reg, y = age)) +
  geom_boxplot()
```

### 9. Create a density plot of age, splitting the densities by gen using the fill aesthetic

```{r}
boys %>% 
  ggplot(aes(x = age, fill = gen)) +
  geom_density(alpha = 0.2)
```

### 10. Create a diverging bar chart for hgt in the boys dataset, that displays for every age year that year's mean height in deviations from the ovarall average hgt

```{r}
boys %>% 
  mutate(age_cat = factor(floor(age)),
         dev_hgt = hgt - mean(hgt, na.rm = TRUE)) %>% 
  group_by(age_cat) %>% 
  summarise(hgt = mean(dev_hgt, na.rm = TRUE)) %>% 
  ggplot(aes(x = hgt, y = age_cat)) +
  geom_bar(stat = "identity")
```

## Regression visalisation

### 11. Load the data elastic1 and elastic2 and bind the data frames together using the function bind_rows() and add a grouping variable indicating whether an observation comes from elastic1 or from elastic2

```{r}
elastic <- bind_rows(elastic1 %>%
                       mutate(origin = "elastic1"),
                     elastic2 %>% 
                       mutate(origin = "elastic2"))
```

### 12. Create a scatterplot mapping stretch on the x-axis and distance on the y-axis, and map the just created group indicator as the color aesthetic

```{r}
elastic %>% 
  ggplot(aes(x = stretch, y = distance, col = origin)) +
  geom_point()
```

### 13. Recreate the previous plot, but now assess whether the results of the two datasets appear consistent by adding a linear regression line

```{r}
elastic %>% 
  ggplot(aes(x = stretch, y = distance, col = origin)) +
  geom_point() +
  geom_smooth(method = "lm")
# looks consistent
```

### 14. For each of the data sets elastic1 and elastic2, fit a regression model with y = distance on x = stretch using lm()

```{r}
mod1 <- lm(distance ~ stretch, elastic1)
mod2 <- lm(distance ~ stretch, elastic2)
```

### 15. For both of the previously created fitted models, determine the fitted values and the standard errors of the fitted values, and the proportion explained variance

```{r}
predict(mod1, se.fit = FALSE)
summary(mod1)

predict(mod2, se.fit = FALSE)
summary(mod2)
```

### 16. Study the residual versus leverage plots for both models

```{r}
plot(mod1, which = 5)
plot(mod2, which = 5)
```

### 17. Use the elastic2 variable stretch to obtain predictions on the model fitted on elastic1

```{r}
pred <- predict(mod1, newdata = data.frame(stretch = elastic2$stretch))
pred
```

### 18. Now make a scatterplot to investigate similarity between the predicted values and the observed values for elastic2

```{r}
plot(elastic2$distance, pred)
```

