---
title: "Practical 8"
author: "Judith Neve"
date: '2022-10-28'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(magrittr)
library(psych)
library(caret)
library(gbm)
library(xgboost)
library(data.table)
library(ggforce)

set.seed(45)
df <- readRDS("data/train_disease.RDS")
```

## 1. Get an impression of the data by looking at the structure of the data and creating some descriptive statistics.

```{r}
str(df)
```

```{r}
summary(df)
```

## 2. To further explore the data we work with, create some interesting data visualizations that show whether there are interesting patterns in the data.

```{r}
df %>% 
  ggplot(aes(x = Albumin, y = Ratio_Albumin_Globulin, col = Disease)) +
  geom_point()
```

## 3. Shortly reflect on the difference between bagging, random forests, and boosting.

Boosting uses prior information, bagging fits trees using bootstrapped data, random forests fit trees using random set of predictors

## 4. Apply bagging to the training data, to predict the outcome Disease, using the caret library.

```{r}
cvcontrol <- trainControl(method = "repeatedcv", 
                          number = 10,
                          allowParallel = TRUE)
```

```{r}
bagged_tree <- train(Disease ~ .,
                     df,
                     trControl = cvcontrol,
                     importance = TRUE,
                     method = "treebag")
```

## 5. Interpret the variable importance measure using the varImp function on the trained model object.

```{r}
varImp(bagged_tree)
```

## 6. Create training set predictions based on the bagged model, and use the confusionMatrix() function from the caret package to assess it’s performance.

```{r}
preds <- predict(bagged_tree)
confusionMatrix(preds, df$Disease)
```

## 7. Now ask for the output of the bagged model. Explain why the under both approaches differ.

```{r}
bagged_tree
```

## 8. Fit a random forest to the training data to predict the outcome Disease, using the caret library.

```{r}
df_rf <- train(Disease ~ .,
               df,
               method = "rf",
               trControl = cvcontrol,
               importance = TRUE)
```

## 9. Again, interpret the variable importance measure using the varImp function on the trained model object. Do you draw the same conclusions as under the bagged model?

```{r}
varImp(df_rf) # some pretty big differences
```

## 10. Output the model output from the random forest. Are we doing better than with the bagged model?

```{r}
df_rf
```

## 11. Now, fit a boosting model using the caret library to predict disease status.

```{r}
df_boost <- train(Disease ~ .,
               df,
               method = "gbm",
               trControl = cvcontrol)
```

## 12. Again, interpret the variable importance measure. You will have to call for summary() on the model object you just created. Compare the output to the previously obtained variable importance measures.

```{r}
summary(df_boost)
```

## 13. Output the model output from our gradient boosting procedure. Are we doing better than with the bagged and random forest model?

```{r}
df_boost
```

## 14. Download the file shap.R from this Github repository.

```{r}
library(devtools)
source_url("https://github.com/pablo14/shap-values/blob/master/shap.R?raw=TRUE")
```

## 15. Specify your model as follows, and use it to create predictions on the training data.

```{r}
train_x <- model.matrix(Disease ~ ., df)[,-1]
train_y <- as.numeric(df$Disease) - 1
xgboost_train <- xgboost(data = train_x,
                         label = train_y, 
                         max.depth = 10,
                         eta = 1,
                         nthread = 4,
                         nrounds = 4,
                         objective = "binary:logistic",
                         verbose = 2)



pred <- tibble(Disease = predict(xgboost_train, newdata = train_x)) %>%
  mutate(Disease = factor(ifelse(Disease < 0.5, 1, 2),
                          labels = c("Healthy", "Disease")))

table(pred$Disease, df$Disease)
```

## 16. First, calculate the SHAP rank scores for all variables in the data, and create a variable importance plot using these values. Interpret the plot.

```{r}
shap_results <- shap.score.rank(xgboost_train,
                                X_train = train_x,
                                shap_approx = F)

var_importance(shap_results)

```

## 17. Plot the SHAP values for every individual for every feature and interpret them.

```{r}
shap_long <- shap.prep(shap = shap_results,
                       X_train = train_x)

plot.shap.summary(shap_long)
xgb.plot.shap(train_x, features = colnames(train_x), model = xgboost_train, n_col = 3)
```

## 18. Verify which of the models you created in this practical performs best on the test data.

```{r}
test <- readRDS("data/test_disease.RDS")

outcome <- test$Disease
bagged_test <- predict(bagged_tree, newdata = test)
rf_test <- predict(df_rf, newdata = test)
gbm_test <- predict(df_boost, newdata = test)
xgb_test <- predict(xgboost_train, newdata = model.matrix(Disease ~ ., test)[,-1]) %>%
  factor(x = ifelse(. < 0.5, 1, 2), levels = c(1,2), labels = c("Healthy", "Disease"))

list(bagged_test, 
     rf_test, 
     gbm_test, 
     xgb_test) %>%
  map(~ confusionMatrix(.x, test$Disease))
# the random forest performs best
```

