---
title: "Supervised Learning and Visualization Practical 2"
author: "Daniel Anadria"
date: 24 September 2022
output:
  html_document:
    css: Daniel_02_Data_Visualisation_using_ggplot2.css
    
---
<p style="text-align: center;">**Practical 2: Data Visualisation using ggplot2**</p>

```{r, results='hide', message=FALSE, warning=FALSE}
library(ISLR)
library(tidyverse)
library(ggplot2)
library(haven)
# library(readxl)
```

```{r}
homeruns_plot <- 
  ggplot(Hitters, aes(x = Hits, y = HmRun)) +
  geom_point() +
  labs(x = "Hits", y = "Home runs")

homeruns_plot
```

```{r}
homeruns_plot + 
  geom_density_2d() +
  labs(title = "Cool density and scatter plot of baseball data") +
  theme_minimal()
```

1. Name the aesthetics, geoms, scales, and facets of the above visualisation. Also name any statistical transformations or special coordinate systems.

Let's break down the following expression which is a combination of the two chunks above:

```{r, eval = F}
ggplot(Hitters, aes(x = Hits, y = HmRun)) +
  geom_point() +
  labs(x = "Hits", y = "Home runs")+ 
  geom_density_2d() +
  labs(title = "Cool density and scatter plot of baseball data") +
  theme_minimal()
```

The function `ggplot()` initializes a new ggplot by creating a new ggplot object.
Within this function, the arguments are `data` and `mapping`.
For the `data` argument, we specify `Hitters`, which is a baseball dataset from the `ISLR` package. More information about the dataset can be found by running `?ISLR::Hitters`.
The `mapping` argument specifies the aesthetic mapping to use for the plot. Here we set to to `aes()` which constructs aesthetic mappings. Aesthetic mappings describe how variables in the data are mapped to visual properties (aesthetics) of geoms.

The following layer is the geom layer which takes the data with its aesthetic mapping and combines them to a geom (geometric) object. In our case, we specify this layer using `geom_point()` which will plot all the observations as points on in the x-y coordinate system.

The next layer assigns the labels to the x and y axes using `labs()`.

The subsequent layer uses `geom_density_2d()` which calculates and plots the contours of a 2D density estimate.
This is a tranformation of the original data which is useful to see how the points are spatially distributed.

Then we call `labs()` again, but this time was pass the `title` argument which gives our plot a title.

Finally, we use `theme_minimal()` to remove the default gray background.

2. Run the code below to generate data. There will be three vectors in your environment. Put them in a data frame for entering it in a `ggplot()` call using either the `data.frame()` or the `tibble()` function. Give informative names and make sure the types are correct (use the `as.<type>()` functions). Name the result `gg_students`.

```{r}
set.seed(1234)
student_grade  <- rnorm(32, 7)
student_number <- round(runif(32) * 2e6 + 5e6)
programme      <- sample(c("Science", "Social Science"), 32, replace = TRUE)
```

```{r}
gg_students <- data.frame("grade" = student_grade, "number" = as.character(student_number), program = as.factor(programme))  
```

3. Plot the first homeruns_plot again, but map the Hits to the y-axis and the HmRun to the x-axis instead

```{r}
ggplot(Hitters, aes(x = HmRun, y = Hits))+
  geom_point()+
  labs(x = "Home runs", y = "Hits", title = "Scatter Plot of Baseball Data")+
  theme_minimal()
```

4. Recreate the same plot once more, but now also map the variable League to the colour aesthetic and the variable Salary to the size aesthetic.


```{r, warning=F}
ggplot(Hitters, aes(x = HmRun, y = Hits, color = League, size = Salary))+
  geom_point()+
  labs(x = "Home runs", y = "Hits", title = "Scatter Plot of Baseball Data")+
  theme_minimal()
```


5. Look at the many different geoms on the reference website.

6. Use `geom_histogram()` to create a histogram of the grades of the students in the `gg_students` dataset. Play around with the `binwidth` argument of the `geom_histogram()` function.

```{r}
ggplot(gg_students, aes(x = grade))+
  geom_histogram(binwidth = 0.05)+
  labs(x = "Grade", y = "Count")+
  theme_minimal()
```

7. Use `geom_density()` to create a density plot of the grades of the students in the `gg_students` dataset. Add the argument `fill = "light seagreen"` to `geom_density()`.


```{r}
ggplot(gg_students, aes(x = grade))+
  geom_density(fill = "light seagreen")
```

8. Add rug marks to the density plot through `geom_rug()`. You can edit the `colour` and `size` of the rug marks using those arguments within the geom_rug() function.


```{r}
ggplot(gg_students, aes(x = grade))+
  geom_density(fill = "light seagreen")+
  geom_rug(color = 'red', size = 2)
```

9. Increase the data to ink ratio by removing the y axis label, setting the theme to `theme_minimal()`, and removing the border of the density polygon. Also set the limits of the x-axis to go from 0 to 10 using the `xlim()` function, because those are the plausible values for a student grade.

```{r}
ggplot(gg_students, aes(x = grade))+
  geom_density(fill = "light seagreen")+
  geom_rug(color = 'red', size = 2)+
  labs(y = "") +
  xlim(0, 10)+
  theme_minimal()
```

10. Create a boxplot of student grades per programme in the `gg_students` dataset you made earlier: map the `programme` variable to the x position and the `grade` to the y position. For extra visual aid, you can additionally map the `programme` variable to the `fill aesthetic`.

```{r}
ggplot(gg_students, aes(x = program, y = grade, fill = program))+
  geom_boxplot()+
  theme_minimal()
  
```

11. What do each of the horizontal lines in the boxplot mean? What do the vertical lines (whiskers) mean?

The horizontal lines in the center of the box indicate the median, and the borders of each box indicate the interquantile range (IQR). The vertical lines are called whiskers and the upper and lower whiskers represent scores outside the middle 50%. 


12. Comparison of distributions across categories can also be done by adding a `fill` aesthetic to the density plot you made earlier. Try this out. To take care of the overlap, you might want to add some transparency in the `geom_density()` function using the `alpha` argument.

```{r}
ggplot(gg_students, aes(x = grade, fill = program))+
  geom_density(alpha = 0.5)+
  geom_rug(color = 'red', size = 2)+
  labs(y = "")+
  xlim(0, 10)+
  theme_minimal()
```

13. Create a bar plot of the variable `Years` from the `Hitters` dataset. 

```{r}
ggplot(Hitters, aes(x = Years))+
  geom_bar()
```

14. Use `geom_line()` to make a line plot out of the first 200 observations of the variable `Volume` (the number of trades made on each day) of the `Smarket` dataset. You will need to create a `Day` variable using `mutate()` to map to the x-position. This variable can simply be the integers from 1 to 200. Remember, you can select the first 200 rows using `Smarket[1:200, ]`.

```{r}
Smarket <- Smarket[1:200,] %>% mutate(Day = 1:200 )
```

```{r}
ggplot(Smarket, aes(x = Day, y = Volume))+
  geom_line()+
  theme_minimal()
```

15. Give the line a nice colour and increase its size. Also add points of the same colour on top.

```{r}
ggplot(Smarket, aes(x = Day, y = Volume))+
  geom_line(color = "blue", size = 1)+
  geom_point(colour = "blue", size = 1)+
  theme_minimal()
```

16. Use the function `which.max()` to find out which of the first 200 days has the highest trade volume and use the function `max()` to find out how large this volume was.

```{r}
which.max(Smarket$Volume) # the index of the highest volume
max(Smarket$Volume) # the value of the higest volume
```


17. Use `geom_label(aes(x = your_x, y = your_y, label = "Peak volume"))` to add a label to this day. You can use either the values or call the functions. Place the label near the peak!


```{r}
ggplot(Smarket, aes(x = Day, y = Volume))+
  geom_line(color = "blue", size = 1)+
  geom_label(aes(x = 170, y = 2.5, label = "Peak volume"))+
  theme_minimal()
```

18. Create a data frame called baseball based on the `Hitters` dataset. In this data frame, create a factor variable which splits players’ salary range into 3 categories. Tip: use the `filter()` function to remove the missing values, and then use the `cut()` function and assign nice labels to the categories. In addition, create a variable which indicates the proportion of career hits that was a home run.

```{r}
newhitters <- Hitters %>% 
  filter(!is.na(Salary)) %>% 
  mutate(SalaryCat = cut(Salary, breaks = 3, 
         labels = c("Low salary", "Medium salary", "High salary")),
         Proportion_Hits_HomeRun = CHmRun/CHits)
```


19. Create a scatter plot where you map `CWalks` to the x position and the `proportion` you calculated in the previous exercise to the y position. Fix the y axis limits to (0, 0.4) and the x axis to (0, 1600) using ylim() and xlim(). Add nice x and y axis titles using the labs() function. Save the plot as the variable `baseball_plot`.

```{r}
baseball_plot <- 
  ggplot(newhitters, aes(x = CWalks, y = Proportion_Hits_HomeRun))+
  geom_point()+
  ylim(0, 0.4)+
  xlim(0, 1600)+
  labs(x = "Number of walks during player's career", y = "Proportion of home runs")+
  theme_minimal()
  

baseball_plot
```

20. Split up this plot into three parts based on the salary range variable you calculated. Use the `facet_wrap()` function for this; look at the examples in the help file for tips.

```{r}
baseball_plot + facet_wrap(vars(SalaryCat))
```

21. Create an interesting data visualisation based on the `Carseats` data from the `ISLR` package.

```{r}
carseats <- Carseats %>% 
  mutate(PriceDiff = CompPrice - Price) %>% 
  mutate(IncomeCat = cut(Income, breaks = 3, 
         labels = c("Low income", "Medium income", "High income"))) %>% 
  mutate(Advertising = cut(Advertising, breaks = 3, 
         labels = c("Low advertising", "Medium advertising", "High advertising")))
  
```


I compute the price difference between the competitor and our location.
Higher difference values indicate that our location is cheaper than the competitors.
Next, I split Income and Advertising into three categories.


```{r}
ggplot(carseats, aes(x = PriceDiff, y = Sales, color = Advertising))+
  geom_point()+
  facet_wrap(vars(IncomeCat))+
  labs(x = "Price Difference with Competitors",
       y = "Sales", 
       title = "Child Car Seat Sales by Income and Advertising Level")
```


The end.
