---
title: 'Practical 3: Wrangling & Visualisation'
author: "Nina van Gerwen (1860852)"
date: "28th of September, 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploring missingness

### Loading the data and packages

```{r include=FALSE}
library(tidyverse)
library(magrittr)
library(mice)
library(DAAG)

data(boys)

str(boys)

summary(boys)
```


### 1 + 2) Exploring the data through plots

```{r}
boys %>%
  ggplot(data = ., aes(x = age)) +
  geom_histogram(binwidth = .5, col = "black") +
  theme_minimal()

boys %>%
  filter(., !is.na(gen)) %>% ## first remove NA values in order to increase
  ## readability of the graph (as the large majority is NA)
  ggplot(data = ., aes(x = gen)) +
  geom_bar(col = "black") +
  theme_minimal()
```


### 3 + 4) Missingness indicators

```{r}
## Creating the missingness indicators for the three variables
boys <- boys %>%
  mutate(
  miss_gen = as.factor(ifelse(is.na(gen), "Yes", "No")),
  miss_phb = as.factor(ifelse(is.na(phb), "Yes", "No")),
  miss_tv = as.factor(ifelse(is.na(tv), "Yes", "No"))
  )

## Assessing whether missingness is related to age through logistic regression
boys %>% glm(miss_gen ~ age, data = ., family = "binomial") %>% summary
boys %>% glm(miss_phb ~ age, data = ., family = "binomial") %>% summary
boys %>% glm(miss_tv ~ age, data = ., family = "binomial") %>% summary
```

Missingness very much is related to age.


### 5 + 6) More plots for visualization

```{r}
boys %>%
  ggplot(data = ., aes(x = age, fill = miss_gen)) +
  geom_histogram(binwidth = .5) +
  facet_wrap(boys$miss_gen)

boys %>%
  ggplot(data = ., aes(x = age, y = bmi, col = miss_gen)) +
  geom_point()
```

## Visualising the data


### 8 + 9 + 10) Even more plots

```{r}
boys %>%
  ggplot(data = ., aes(y = age, fill = reg)) +
  geom_boxplot() +
  theme_minimal()

boys %>%
  ggplot(data = ., aes(x = age, fill = gen)) +
  geom_density(alpha = .5) +
  theme_minimal()

boys %>%
  ## First, create a categorical age variable
  mutate(cat_age = cut(age, breaks = 22, labels = 0:21),
         ## Also calculate overall mean age
         mean_hgt = mean(hgt, na.rm = TRUE)) %>%
  ## Then group by the categorical age variable
  group_by(cat_age) %>%
  ## Calculate the height difference in each age category compared to mean age
  mutate(hgt_diff = round(mean(hgt, na.rm = TRUE) - mean_hgt)) %>%
  ## select the two variables of interest for the plot
  select(hgt_diff, cat_age) %>% 
  ## we want only distinct values (otherwise geom_bar doesn't work properly)
  distinct(.) %>%
  ## create the plot
  ggplot(data = ., aes(x = cat_age, y = hgt_diff)) +
  geom_bar(stat = "identity", show.legend = FALSE,
           aes(fill = hgt_diff)) +
  coord_flip() +
  labs(x = "Height difference (in cm)", y = "Age (in years)") +
  theme_minimal()

```


I put a lot of effort in this graph, appreciate it!


## Regression visualisation

### 11) Data preparation

```{r}
data(elastic1)
data(elastic2)

elastic1 <- cbind(elastic1, indicator = as.factor("1"))
elastic2 <- cbind(elastic2, indicator = as.factor("2"))

elastic <- bind_rows(elastic1, elastic2)
```


### 12) Scatter plot

```{r}
elastic %>%
  ggplot(data = ., aes(x = stretch, y = distance, col = indicator)) +
  geom_point()
```

### 13) Recreating the scatter plot

```{r}
elastic %>%
  ggplot(data = ., aes(x = stretch, y = distance, col = indicator)) +
  geom_point() +
  geom_smooth(method = lm, formula = y ~ x)
```
The results appear consistent.

### 14) Regression results

```{r}
results1 <- lm(distance ~ stretch, data = elastic1)
results2 <- lm(distance ~ stretch, data = elastic2)
```


### 15) Getting results

```{r}
## Fitted values and standard errors:
predict(results1, se.fit = TRUE)
predict(results2, se.fit = TRUE)

## Proportion explained variance
summary(results1)
summary(results2)
```


### 16) Residual plots

```{r}
plot(results1, which = 5)
plot(results2, which = 5)
```

### 17) Cross-model predictions

```{r}
elastic2$predicted_values <- -119.143 + 6.571 * elastic2$stretch
```


### 18) More plots

```{r}
library(reshape2)
## get the data
elastic2 %>%
  ## select only the useful variables
  select(stretch, predicted_values, observed_values = distance) %>%
  ## melt the data into a long format, with stretch as the indicator
  melt(., id.vars = "stretch") %>%
  ## then, create a scatterplot with stretch on the x-axis
  ## and the distance values on y-axis.
  ## and colour is the factor of whether they are predicted or observed
  ggplot(data = ., aes(x = stretch, y = value, col = variable)) +
  geom_point()
```
From the plot, you can see that the predicted values are similar to the
observed values.