---
title: "excerise_4"
author: "Florian van Leeuwen"
date: "11/6/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Ensemble methods

```{r}
library(tidyverse)
library(magrittr)
library(psych)
library(caret)
library(gbm)
library(xgboost)
library(data.table)
library(ggforce)
```

```{r}
set.seed(45)
train_disease <- readRDS("~/Documents/GitHub/INFOMDA1-2022/Practicals/P8/Florian van Leeuwen/data-3/train_disease.RDS")
```
## 1. Get an impression of the data by looking at the structure of the data and creating some descriptive statistics.

```{r}
str(train_disease)

head(train_disease)

describe(train_disease)

train_disease %>%
  select(-c(Gender, Disease)) %>%
  describeBy(df$Disease, fast = TRUE)
```

## 2. To further explore the data we work with, create some interesting data visualizations that show whether there are interesting patterns in the data.
```{r}
train_disease %>% 
  ggplot(aes(x = Age, y = Total_Bilirubin, color = Disease, shape = Gender)) +
  geom_point() +
  theme_classic()

train_disease %>%
  select(-Gender) %>%
  pivot_longer(where(is.numeric)) %>%
  ggplot(aes(x = value, col = Disease, fill = Disease)) +
  geom_boxplot(alpha = 0.8) +
  facet_wrap(~name, scales = "free") +
  scale_color_brewer(palette = "Paired") +
  scale_fill_brewer(palette = "Paired") +
  theme_minimal()
```

## 3. Shortly reflect on the difference between bagging, random forests, and boosting.
bagging = running the same model on different datasets using sameling with replacement
random forest = running multiple different trees who can use a random selection of predictors. The avaerge output od these trees will be used.
boosting = selecting the mislasifications and adding them to the dataset for the next model etc
```{r}
## Bagging:       fit a regression tree to N bootstrap samples of the training data
##                take the average of all classification trees to base predictions on
##                Note: out-of-bag data can serve as internal validation set.

## Random forest: Similarly to bagging, classification trees are trained on 
##                a bootstrap sample of the data. However, the decision trees
##                are trained using less than all features in the data. 

## Boosting:      We build a decision tree sequentially. Given the current
##                we fit a (small) tree on the residuals of the current model, 
##                rather than on the outcome Y
```

## 4. Apply bagging to the training data, to predict the outcome Disease, using the caret library.
```{r}
cvcontrol <- trainControl(method = "repeatedcv", 
                          number = 10,
                          allowParallel = TRUE)

M1 <- train(Disease ~ .,
                   data = train_disease, 
                   method = 'treebag',
                   trControl = cvcontrol,
                   importance = TRUE)
```

## 5. Interpret the variable importance measure using the varImp function on the trained model object.
```{r}
M1 %>% 
  varImp %>% 
  plot
```

## 6. Create training set predictions based on the bagged model, and use the confusionMatrix() function from the caret package to assess it’s performance.`
```{r}
confusionMatrix(predict(M1, type = "raw"), train_disease$Disease)
```

## 7. Now ask for the output of the bagged model. Explain why the under both approaches differ.
```{r}
M1
```

## 8. Fit a random forest to the training data to predict the outcome Disease, using the caret library.
```{r}
M2 <- train(Disease ~ .,
                   data = train_disease, 
                   method = 'rf',
                   trControl = cvcontrol,
                   importance = TRUE)
```

## 9. Again, interpret the variable importance measure using the varImp function on the trained model object. Do you draw the same conclusions as under the bagged model?
```{r}
M2 %>% 
  varImp %>% 
  plot
```
## 10. Output the model output from the random forest. Are we doing better than with the bagged model?
```{r}
M2

# yes accuracy is lower
```

## 11. Now, fit a boosting model using the caret library to predict disease status.
```{r}
M3 <- train(Disease ~ .,
                   data = train_disease, 
                   method = 'gbm',
                   verbose = F,
                   trControl = cvcontrol)
```

## 12. Again, interpret the variable importance measure. You will have to call for summary() on the model object you just created. Compare the output to the previously obtained variable importance measures.
```{r}
summary(M3)
```

## 13. Output the model output from our gradient boosting procedure. Are we doing better than with the bagged and random forest model?
```{r}
M3

# Yes, our best model is doing slightly better then the previous two models.
# However, this might still be random variation.
```
## 14. 
```{r}
library(devtools)
source_url("https://github.com/pablo14/shap-values/blob/master/shap.R?raw=TRUE")
```
## 15. Specify your model as follows, and use it to create predictions on the training data.
```{r}
train_x <- model.matrix(Disease ~ ., train_disease)[,-1]
train_y <- as.numeric(train_disease$Disease) - 1
xgboost_train <- xgboost(data = train_x,
                         label = train_y, 
                         max.depth = 10,
                         eta = 1,
                         nthread = 4,
                         nrounds = 4,
                         objective = "binary:logistic",
                         verbose = 2)



pred <- tibble(Disease = predict(xgboost_train, newdata = train_x)) %>%
  mutate(Disease = factor(ifelse(Disease < 0.5, 1, 2),
                          labels = c("Healthy", "Disease")))

table(pred$Disease, train_disease$Disease)
```

## 16. First, calculate the SHAP rank scores for all variables in the data, and create a variable importance plot using these values. Interpret the plot.
```{r}
shap_results <- shap.score.rank(xgboost_train,
                                X_train = train_x,
                                shap_approx = F)

var_importance(shap_results)
```

## 17. Plot the SHAP values for every individual for every feature and interpret them.
```{r}
shap_long <- shap.prep(shap = shap_results,
                       X_train = train_x)

plot.shap.summary(shap_long)
```
```{r}
xgb.plot.shap(train_x, features = colnames(train_x), model = xgboost_train, n_col = 3)
```
## 18. Verify which of the models you created in this practical performs best on the test data.

```{r}
test_disease <- readRDS("~/Documents/GitHub/INFOMDA1-2022/Practicals/P8/Florian van Leeuwen/data-3/test_disease.RDS")

# bagging
X1 <- confusionMatrix(predict(M1, newdata = test_disease),test_disease$Disease )

# random forest
X2 <- confusionMatrix(predict(M2, newdata = test_disease),test_disease$Disease )

# Gradien boosting
X3 <- confusionMatrix(predict(M3, newdata = test_disease),test_disease$Disease )

# XG boost
pred <- tibble(Disease = predict(xgboost_train, newdata = model.matrix(Disease ~ ., test_disease)[,-1])) %>%
  mutate(Disease = factor(ifelse(Disease < 0.5, 1, 2),
                          labels = c("Healthy", "Disease")))

X4 <- confusionMatrix(pred$Disease, test_disease$Disease)

data_frame(Model = c("bagging", "Random Forest", "Gradient boosting", "Xgboost"),
           Accuracy = c(X1$overall[1],X2$overall[1],X3$overall[1],X4$overall[1]) )
```

