---
title: "Practical 1"
author: "Jim Regtien"
date: "12-9-2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Exercise 1

```{r}
library(tidyverse)
library(readxl)
object_1 <- 1:5
object_2 <- 1L:5L
object_3 <- "-123.456"
object_4 <- as.numeric(object_2)
object_5 <- letters[object_1]
object_6 <- as.factor(rep(object_5, 2))
object_7 <- c(1, 2, 3, "4", "5", "6")
```

The first object will be a list of ints, same for the second and fourth objects. The third object will be a string/character. Object 5 will be a vector/list of strings, like the seventh object. The 6th object is a factor.

```{r}
print(class(object_1))
print(class(object_2))
print(class(object_3))
print(class(object_4))
print(class(object_5))
print(class(object_6))
print(class(object_7))
```
### Exercise 2
```{r}
as.numeric(object_7 )
```

### Exercise 3

```{r} 
obj_list <- list(object_1, object_2, object_3, object_4, object_5, object_6, object_7)
obj_list
```

### Exercise 4

```{r}
obj_df <- data.frame(var_1 = object_1, var_2 = object_2, var_3 = object_5)
obj_df
```

### Exercise 5

```{r}
print(nrow(obj_df)) 
print(ncol(obj_df))
```

### Exercise 9
```{r, message = FALSE}
apps <- read_csv('googleplaystore.csv')
print(head(apps))
```

The number of reviews is a double, while I would find it more natural for it to be an integer. The price is a chr, while it might be better if it was a double. 

### Exercise 9
```{r, message = FALSE}
students <- read_xlsx('students.xlsx')
print(head(students))
```

The student number should be an integer, as that type of objects requires less memory. 

### Exercise 10
```{r}
summarise(students, mean =  round(mean(grade), 2), median = round(median(grade), 2),
          variance = round(var(grade), 2), min = round(min(grade), 2), 
          max = round(max(grade), 2))
```
The grades range from a 4.8 to a 9.3.

### Exercise 11

```{r}
students %>%
  filter(grade <= 5.5)
```

### Exercise 12

```{r}
students %>%
  filter(grade >= 8) %>%
  filter(programme == "A")
```

### Exercise 13

```{r}
students %>%
  arrange(programme, grade)
```

### Exercise 14 
```{r}
students %>%
  select(student_number, programme)
```

### Exercise 15

```{r}
students_recoded <- mutate(students, prog = recode(as.vector(students$programme), 
                                                   A = "Science",
                                                   B = "Social Science"))
```

### Exercise 16

```{r, message=FALSE}
popular_apps <- (read_csv('googleplaystore.csv') %>%
  mutate(downloads = parse_number(Installs)) %>%
  filter(downloads >= 5e7) %>%
  arrange(desc(Rating)))
```

### Exercise 17

```{r}
popular_apps %>%
  summarise(median = median(Rating), min = min(Rating), max = max(Rating))
```
### Exercise 18
```{r}
mad <- function(x) median(abs(x - median(x)))

```

```{r}
popular_apps %>%
  summarise(median = median(Rating), min = min(Rating), max = max(Rating), mad = mad(Rating))
```

### Exercise 19

```{r, message=FALSE}
popular_apps %>%
  group_by(Category) %>%
  summarise(median = median(Rating), min = min(Rating), max = max(Rating),
            mad = mad(Rating))
```

### Exercise 20
For this exercise, I will study if paid apps are rated higher than non-paid apps (i.e. does 'buyers remorse' play a role when rating apps). As we can see, the freely available apps are rated about 0.1 point lower than the paid apps. This is not a very significant difference. 

```{r, message=FALSE}
apps %>%
  group_by(Type) %>%
  na.omit() %>% #Ommiting the NaNs
  summarise(mean = round(mean(Rating), 1), median = round(median(Rating), 2), sd = round(sd(Rating), 2))
```