---
title: "Practical"
author: "Hsuan Lee"
output: html_document
---
```{r}
library(tidyverse)
library(magrittr)
library(mice)
library(DAAG)
```

## The `boys` data
```{r}
head(boys)
```

To verify if the data is sorted on the variable `age`
```{r}
!is.unsorted(boys$age)
```

```{r}
summary(boys)
```

**1. Create a histogram of the variable age using the function geom_histogram().**
```{r}
boys %>%
  ggplot(aes(x = age)) +
  geom_histogram()
```

**2. Create a bar chart of the variable gen using the function geom_bar() .**
```{r}
boys %>%
  ggplot(aes(x = gen, fill = gen)) +
  geom_bar()
```

## Assessing missing data

```{r}
md.pattern(boys)
```

**3. Create a missingness indicator for the variables gen , phb and tv .**
```{r}
boys_miss <- boys %>%
  mutate(gen_miss = is.na(gen), 
         phb_miss = is.na(phb),
         tv_miss = is.na(tv))
```

**4. Assess whether missingness in the variables gen, phb and tv is related to someones age.**
```{r}
boys_miss %>%
  group_by(gen_miss) %>%
  summarise(mean_age = mean(age))
```

```{r}
boys_miss %>%
  group_by(phb_miss) %>%
  summarise(mean_phb = mean(age))
```

```{r}
boys_miss %>%
  group_by(tv_miss) %>%
  summarise(mean_age = mean(age))
```

**5. Create a histogram for the variable age , faceted by whether or not someone has a missing value on gen .**
```{r}
boys_miss %>%
  ggplot(aes(x = age)) +
  geom_histogram(fill = "dark blue") +
  facet_wrap(~gen)
```

**6. Create a scatterplot with age on the x-axis and bmi on the y-axis, using the function geom_point() .**
```{r}
boys_miss %>%
  ggplot(aes(x = age, y = bmi)) +
  geom_point()
```

**7. Add a colour aesthetic to the previous plot using the missingness indicator of the variable gen .**
```{r}
boys_miss %>%
  ggplot(aes(x = age, y = bmi)) +
  geom_point(aes(col = gen_miss)) +
  scale_color_manual(values  = c("dark blue", "dark grey")) +
  labs(x = "Age", y = "BMI", color = "gen missingness")
```

## Visualizing the boys data

**8. Visualize the relationship between reg (region) and age using a boxplot.**
```{r}
boys_miss %>%
  ggplot() +
  geom_boxplot(aes(x = reg, y = age, fill = reg)) +
  scale_fill_manual(values = c("dark green","yellow", "black", 
                               "orange", "brown", "purple"))
```

**9. Create a density plot of age, splitting the densities by gen using the fill aesthetic.**
```{r}
boys_miss %>%
  ggplot(aes(x = age, fill = gen)) +
  geom_density() +
  scale_fill_manual(values = c("#882E72", "#5289C7", "#CAE0AB", 
                               "#E8601C", "#DC050C", "#D6C1DE", "#4477AA"))
```

**10. Create a diverging bar chart for hgt in the boys data set, that displays for every age year that year’s mean height in deviations from the overall average hgt .**

Hint: You will have to create a categorical age variable for every year, and center the variable hgt such that it reflects that person’s difference from the mean height in the entire data.
```{r}
boys_miss %>%
  mutate(age_cat = cut(age, breaks = c(0:22), label = paste((0:21), 'year old')),
         hgt_dif = hgt - mean(hgt, na.rm = T)) %>%
  group_by(age_cat) %>%
  summarise(mean_hgt = mean(hgt_dif, na.rm = T)) %>%
  ggplot(aes(x = mean_hgt, y = age_cat)) +
  geom_bar(stat = "identity")
```

## Regression visualization

**11. Load the data elastic1 and elastic2 and bind the data frames together using the function bind_rows() and add a grouping variable indicating whether an observation comes from elastic1 or from elastic2.**
```{r}
df <- bind_rows(elastic1 = elastic1, 
                elastic2 = elastic2, 
                .id = "group")
```

**12. Create a scatterplot mapping stretch on the x-axis and distance on the y-axis, and map the just created group indicator as the color aesthetic.**
```{r}
df %>%
  ggplot(aes(x = stretch, y = distance)) +
  geom_point(aes(color = group))
```

**13. Recreate the previous plot, but now assess whether the results of the two data sets appear consistent by adding a linear regression line.**
```{r}
df %>%
  ggplot(aes(x = stretch, y = distance)) +
  geom_point(aes(color = group)) +
  geom_smooth(aes(color = group), method = 'lm')
```

**14. For each of the data sets elastic1 and elastic2, fit a regression model with y = distance on x = stretch using lm(y ~ x, data).**
```{r}
model_ela1 <- lm(distance~ stretch, data = elastic1)
model_ela2 <- lm(distance~ stretch, data = elastic2)
```

**15. For both of the previously created fitted models, determine the fitted values and the standard errors of the fitted values, and the proportion explained variance R2.**

Hint: Check out the predict function in R and notice the se.fit argument, as well as the summary function.
```{r}
summary(model_ela1)
predict(model_ela1, se.fit = T) # se.fit -> standard error of estimates

summary(model_ela2)
predict(model_ela2, se.fit = T)
```

**16. Study the residual versus leverage plots for both models.**
Hint: Use plot(which = 5) on the fitted objects

A residuals vs. leverage plot is a type of diagnostic plot that allows us to identify influential observations in a regression model.
Leverage refers to the extent to which the coefficients in the regression model would change if a particular observation was removed from the dataset.Observations with high leverage have a strong influence on the coefficients in the regression model
```{r}
plot(model_ela1, which = 5)
plot(model_ela2, which = 5)

model_ela1$residuals # 7 has large residuals
```

**17. Use the elastic2 variable stretch to obtain predictions on the model fitted on elastic1.**

Use model_ela1 on data `elastic2` to find the prediction of `distance`, i.e., use the xs from `elastic2`, but use the beta from model_ela1
```{r}
pred_val <-predict(model_ela1, newdata = elastic2)
pred_val
```

**18. Now make a scatterplot to investigate similarity between the predicted values and the observed values for elastic2.**
```{r}
elastic2 %>%
  bind_cols(pred_val = pred_val) %>%
  ggplot(aes(x = stretch, y = distance, color = "#882E72")) +
  geom_point(color = "#882E72") +
  geom_smooth(method = "lm", color = "#882E72", se = F) +
  geom_point(aes(y = pred_val), colour = "#5289C7") +
  geom_smooth(aes(y = pred_val),method = "lm", 
              color = "#5289C7", fill = "#5289C7", se = T) +
  scale_color_manual(name = "Values", values = c("predicted value" = "#5289C7", 
                                                 "observed value" = "#882E72")) +
  theme_minimal()
```













