```{r}
library(ISLR)
library(tidyverse)
```

##Question 1:

Aes: x is the number of hits, y is the number of homeruns
Geoms: Points and 2D density lines
Scale: Continuous
Facets: NVT
Transformations: NVT
Coordinate system: Cartesian

##Question 2:

```{r}
set.seed(1234)
student_grade  <- rnorm(32, 7)
student_number <- round(runif(32) * 2e6 + 5e6)
programme      <- sample(c("Science", "Social Science"), 32, replace = TRUE)

gg_students <-
  tibble(
    Number = as.character(student_number),
    Grade = student_grade,
    Programme = as.factor(programme) 
  )

head(gg_students)
```

##Question 3:

```{r}
homeruns_plot <-
  ggplot(Hitters, aes(x= HmRun, y = Hits)) +
  geom_point() +
  labs(x = "Home runs", y = "Hits")
```

##Question 4:

```{r}
homeruns_plot <-
  ggplot(Hitters, aes(x= HmRun, y = Hits, colour = League, size = Salary)) +
  geom_point() +
  labs(x = "Home runs", y = "Hits")
```

##Question 5:

#NVT

##Question 6:

```{r}
grade_hist <-
  ggplot(gg_students, aes(x = Grade)) +
  geom_histogram(binwidth = 0.7)

grade_hist
```

##Question 7:

```{r}
grade_dens <-
  ggplot(gg_students, aes(x = Grade)) +
  geom_density(fill = "light seagreen")

grade_dens
```

##Question 8:

```{r}
grade_dens <-
  ggplot(gg_students, aes(x = Grade)) +
  geom_density(fill = "light seagreen") +
  geom_rug(colour = "orange", size = 0.1)

grade_dens
```

##Question 9:

```{r}
grade_dens <-
  ggplot(gg_students, aes(x = Grade)) +
  geom_density(fill = "light seagreen", colour = NA) +
  geom_rug(colour = "orange", size = 0.1) +
  labs(x = "Grade", y = "") +
  theme_minimal() + 
  xlim(0,10)

grade_dens

```

##Question 10:

```{r}
grade_box <-
  ggplot(gg_students, aes(x = Programme, y = Grade, fill = Programme)) +
  geom_boxplot()

grade_box
```

##Question 11:

#Horizontal lines are the 25th percentile, median, and 75th percentile, the vertical lines shows what points aren't considered outliers

##Question 12:

```{r}
gg_students %>%  
  ggplot(aes(x = Grade, fill = Programme)) +
  geom_density(colour = NA, alpha = 0.5) +
  geom_rug(colour = "orange", size = 0.5) +
  labs(x = "Grade", y = "") +
  theme_minimal() + 
  xlim(0,10)
```

##Question 13:

```{r}
Hitters %>% 
  ggplot(aes(x = Years)) +
  geom_bar() +
  theme_minimal()
```

##Question 14:

```{r}
head(Smarket)

Smarket[1:200,] %>% 
  mutate(Day = 1:200) %>% 
  ggplot(aes(x = Day, y = Volume)) +
  geom_line() +
  theme_minimal()
```

##Question 15:

```{r}
Smarket[1:200,] %>% 
  mutate(Day = 1:200) %>% 
  ggplot(aes(x = Day, y = Volume)) +
  geom_line(colour = "green", size = 0.5) +
  geom_point(colour = "green", size = 1) +
  theme_minimal()
```

##Question 16:

```{r}
which.max(Smarket[1:200,]$Volume)
max(Smarket[1:200,]$Volume)
```

##Question 17:

```{r}
Smarket[1:200,] %>% 
  mutate(Day = 1:200) %>% 
  ggplot(aes(x = Day, y = Volume)) +
  geom_line(colour = "green", size = 0.5) +
  geom_point(colour = "green", size = 1) +
  geom_label(aes(x = 170, y = 2.4, label = "Peak Volume")) +
  theme_minimal()
```

##Question 18:

```{r}
baseball <-
  Hitters %>% 
  filter(!is.na(Salary)) %>% 
  mutate(Range = cut(Salary, breaks = 3, labels = c("Low", "Middle", "High")), HR_Proportion = CHmRun/CHits)

head(baseball)  
```

##Question 19:

```{r}
baseball_plot <-
  baseball %>% 
  ggplot(aes(x = CWalks, y = HR_Proportion)) +
  geom_point() +
  labs(x = "Number of walks", y = "Homerun proportions") +
  xlim(0, 1600) +
  ylim(0, 0.4) +
  theme_minimal()
  
baseball_plot

```

##Question 20:

```{r}
baseball_plot + facet_wrap(~Range)
```

##Question 21:

```{r}
head(Carseats)

Carseats %>% 
  ggplot(aes(x = Age, y = Sales, colour = Income)) +
  geom_point() +
  facet_wrap(~ShelveLoc) +
  theme_minimal()
```

