---
title: "Exploratory Data Analysis"
author: "Ana Martins"
date: "2022-10-02"
output: html_document
---

## Introduction

```{r}
library(tidyverse)
library(magrittr)
library(mice)
library(DAAG)
```

## The `boys` data

```{r}
head(boys)
```

```{r}
tail(boys)
```

```{r}
!is.unsorted(boys$age)
```
```{r}
summary(boys)
```
1. **Create a histogram of the variable `age` using the function `geom_histogram()`.**

```{r}
boys_histo <-
  boys %>%
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 1, boundary = 0) +
  theme_minimal()
  
boys_histo
```

2. **Create a bar chart of the variable `gen` using the function `geom_bar()`.**

```{r}
boys_bar <-
  boys %>%
  ggplot(aes(x = gen)) +
  geom_bar() +
  theme_minimal()
  
boys_bar
```


## Assessing missing data

```{r}
md.pattern(boys)
```
3. **Create a missingness indicator for the variables `gen`, `phb` and `tv.**

```{r}
boys_missingness <- boys %>%
  mutate(gen_missingness = is.na(gen),
         phb_missingness = is.na(phb),
         tv_missingness  = is.na(tv))
```

4. **Assess whether missingness in the variables `gen`, `phb` and `tv` is related to someones age.**

```{r}
boys_missingness %>% 
  group_by(gen_missingness) %>% 
  summarise(age = mean(age))

boys_missingness %>% 
  group_by(phb_missingness) %>% 
  summarise(age = mean(age))

boys_missingness %>% 
  group_by(tv_missingness) %>% 
  summarise(age = mean(age))
```

The mean age for all of the for both FALSE and TRUE is very similar, so age does seem to have a correlation with the missing data.

5. **Create a histogram for the variable `age`, faceted by whether or not someone has a missing value on `gen`.**

```{r}
boys_missingness %>% 
  ggplot(aes(age)) +
  geom_histogram(binwidth = 1, boundary = 0) +
  facet_wrap(~ gen_missingness) +
  theme_minimal()
```
6. **Create a scatterplot with age on the x-axis and bmi on the y-axis, using the function geom_point().**

```{r}
boys_missingness %>% 
  ggplot(aes(x = age, y = bmi)) +
  geom_point() +
  theme_minimal()
```

7. **Add a colour aesthetic to the previous plot using the missingness indicator of the variable gen.**

```{r}
boys_missingness %>% 
  ggplot(aes(x = age, y = bmi, color = gen_missingness)) +
  geom_point() +
  theme_minimal()
```


## Visualizing the boys data

8. Visualize the relationship between reg (region) and age using a boxplot.

```{r}
boys %>% 
  ggplot(aes(x = reg, y = age)) +
  geom_boxplot() +
  theme_minimal()
```
9. **Create a density plot of age, splitting the densities by gen using the fill aesthetic.**

```{r}
boys %>% 
  ggplot(aes(x = age, fill = gen)) +
  geom_density() +
  theme_minimal()
```
10. **Create a diverging bar chart for `hgt` in the `boys` data set, that displays for every `age` year that year’s mean height in deviations from the overall average `hgt`.**

```{r}
boys %>% 
  select(age, hgt) %>% 
  drop_na() %>% 
  mutate(age = as.factor(age), hgt_dev = hgt - mean(hgt, na.rm = TRUE)) %>% 
  group_by(age) %>% 
  summarise(hgt_dev = mean(hgt_dev, na.rm = TRUE)) %>% 
  ggplot(aes(hgt_dev, age)) +
  geom_bar(stat = "identity") +
  theme_minimal()
```

## Regression visualization

11. *Load the data `elastic1` and `elastic2` and bind the data frames together using the function `bind_rows()` and add a grouping variable indicating whether an observation comes from `elastic1` or from `elastic2`.*

```{r}
library(DAAG)

help(elastic1)
help(elastic2)

elastics <-
  bind_rows("elastic1" = elastic1, "elastic2" = elastic2, .id = "set")
```


12. **Create a scatterplot mapping `stretch` on the x-axis and `distance` on the y-axis, and map the just created group indicator as the color aesthetic.**

```{r}
elastics %>% 
  ggplot(aes(x = stretch, y = distance, color = set)) +
  geom_point() +
  theme_minimal()
```

13. **Recreate the previous plot, but now assess whether the results of the two data sets appear consistent by adding a linear regression line.**

```{r}
elastics %>% 
  ggplot(aes(x = stretch, y = distance, color = set)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()
```

14. **For each of the data sets `elastic1` and `elastic2`, fit a regression model with `y = distance` on `x = stretch` using `lm(y ~ x, data)`.**

```{r}
fit1 <- lm(distance ~ stretch, elastic1)
fit2 <- lm(distance ~ stretch, elastic2)
```

15. **For both of the previously created fitted models, determine the fitted values and the standard errors of the fitted values, and the proportion explained variance R2.**

```{r}
fit1 %>%
  predict(se.fit = TRUE)
fit1 %>% summary() %$% r.squared

fit2 %>%
  predict(se.fit = TRUE)
fit2 %>% summary() %$% r.squared
```

16. **Study the residual versus leverage plots for both models.**

```{r}
fit1 %>% plot(which = 5)
fit2 %>% plot(which = 5)
```

17. **Use the `elastic2` variable stretch to obtain predictions on the model fitted on `elastic1`.*

```{r}
pred <- predict(fit1, newdata = elastic2)
```

18. **Now make a scatterplot to investigate similarity between the predicted values and the observed values for `elastic2`.**

```{r}
pred <- data.frame(distance = pred, stretch  = elastic2$stretch)

ggplot() +
  geom_point(data = elastic2, mapping = aes(x = stretch, y = distance), color = "Green") +
  geom_line(data = elastic2, mapping = aes(x = stretch, y = distance), color = "Green") +
  geom_point(data = pred, mapping = aes(x = stretch, y = distance), color = "Red") +
  geom_line(data = pred, mapping = aes(x = stretch, y = distance), color = "Red") +
  theme_minimal()
```

