---
title: "Practical 7 Supervised learning: Nonlinear Regression Introduction" 
subtitle: "Supervised Learning & Visualization"
author: "Ernst Paul Swens"
date: "`r format(Sys.time(), '%B, %Y')`"
output: 
   html_document:
      toc: true
      number_sections: false
      toc_float:
         collapsed: true
         smooth_scroll: true
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(magrittr)
library(psych)
library(caret)
library(gbm)
library(xgboost)
library(data.table)
library(ggforce)

set.seed(45)
df <- readRDS("data/train_disease.RDS")
```

# Q1

**Get an impression of the data by looking at the structure of the data and creating some descriptive statistics.** 

```{r}
head(df)
summary(df)
```

# Q2

**To further explore the data we work with, create some interesting data visualizations that show whether there are interesting patterns in the data.** 

```{r}
df %>% 
   pivot_longer(c("Age", "Total_Bilirubin", "Direct_Bilirubin", 
                  "Alkaline_Phosphotase", "Alamine_Aminotransferase", 
                  "Aspartate_Aminotransferase", "Ratio_Albumin_Globulin")) %>%
   ggplot(aes(x = value, fill = Disease)) +
   geom_boxplot() + 
   facet_wrap(~name, scales = "free")
```

# Q3

**Shortly reflect on the difference between bagging, random forests, and boosting.** 

```{r}
# bagging = we apply bootstrapping on the original data set. We create multiple
# regression trees based on those different bootstrapped samples. And finally,
# we combine the 

# random forest = similar to bagging, however when creating the trees, at each
# step only a subset of all predictors can be considered. This decorrelates the
# trees.

# boosting = create high bias and low variance trees sequentially. First, fit 
# a small stump, and the misclassified records get more weight during the next
# tree creating. 
```

# Q4

**Apply bagging to the training data, to predict the outcome `Disease`, using the `caret` library.** 

```{r}
cvcontrol <- trainControl(
   method = "repeatedcv", 
   number = 10,
   allowParallel = TRUE
)

bag_train <- train(
   Disease ~ .,
   data = df, 
   method = 'treebag',
   trControl = cvcontrol,
   importance = TRUE
)

print(bag_train)
```

# Q5

**Interpret the variable importance measure using the `varImp` function on the trained model object.**

```{r}
plot(varImp(bag_train))
```

# Q6

**Create training set predictions based on the bagged model, and use the `confusionMatrix()` function from the `caret` package to assess it’s performance.**

```{r}
pred = predict(bag_train)
true = df$Disease

caret::confusionMatrix(pred, true, positive = "Healthy")
```

# Q7

**Now ask for the output of the bagged model. Explain why the under both approaches differ.**

```{r}
bag_train
```

```{r}
# this shows the out-of-bag validation
```

# Q8

**Fit a random forest to the training data to predict the outcome `Disease`, using the `caret` library.**

```{r}
cvcontrol <- trainControl(
   method = "repeatedcv", 
   number = 10,
   allowParallel = TRUE
)

rf_train <- train(
   Disease ~ .,
   data = df, 
   method = 'rf',
   trControl = cvcontrol,
   importance = TRUE,
   tuneGrid = expand.grid(.mtry=c(4,10)) # sets the number of variables 
)
```

# Q9

**Again, interpret the variable importance measure using the varImp function on the trained model object. Do you draw the same conclusions as under the bagged model?**

```{r}
plot(varImp(rf_train))
```

# Q10

**Output the model output from the random forest. Are we doing better than with the bagged model?**

```{r}
rf_train
```

# Q11

**Now, fit a boosting model using the `caret` library to predict disease status.**

```{r}
gbm_train <- train(Disease ~ .,
                   data = df,
                   method = "gbm",
                   verbose = FALSE,
                   trControl = cvcontrol)
```

# Q12

**Again, interpret the variable importance measure. You will have to call for `summary()` on the model object you just created. Compare the output to the previously obtained variable importance measures.**

```{r}
plot(varImp(gbm_train)) # no? this gives the same result as the output
```

# Q13

**Output the model output from our gradient boosting procedure. Are we doing better than with the bagged and random forest model?**

```{r}
gbm_train
```

# Q14

**Download the file shap.R from this Github repository.**

```{r}
devtools::source_url("https://github.com/pablo14/shap-values/blob/master/shap.R?raw=TRUE")
```

# Q15

**Specify your model as follows, and use it to create predictions on the training data.**

```{r}
train_x <- model.matrix(Disease ~ ., df)[,-1]
train_y <- as.numeric(df$Disease) - 1

xgboost_train <- xgboost(data = train_x,
                         label = train_y, 
                         max.depth = 10,
                         eta = 1,
                         nthread = 4,
                         nrounds = 4,
                         objective = "binary:logistic",
                         verbose = 2)

pred <- tibble(Disease = predict(xgboost_train, newdata = train_x)) %>%
  mutate(Disease = factor(ifelse(Disease < 0.5, 1, 2),
                          labels = c("Healthy", "Disease")))

table(pred$Disease, df$Disease)
```

# Q16

**First, calculate the SHAP rank scores for all variables in the data, and create a variable importance plot using these values. Interpret the plot.**

```{r}
shap_results <- shap.score.rank(
   xgboost_train,
   X_train = train_x,
   shap_approx = F
)

var_importance(shap_results)
```

# Q17

**Plot the `SHAP` values for every individual for every feature and interpret them.**

```{r}
shap_long <- shap.prep(
   shap = shap_results,
   X_train = train_x
)

plot.shap.summary(shap_long)

xgb.plot.shap(train_x, features = colnames(train_x), model = xgboost_train, n_col = 3)
```

```{r}
# The first plot for a particular feature value, how it will impact the model 
# output. For example, being a woman is associated with a lower probability 
# of being diseased.

# The second plots are simply marginal plots of the SHAP values
```

# Q18

**Verify which of the models you created in this practical performs best on the test data.**

```{r}
data_test <- readRDS("Data/test_disease.RDS")

bag_test <- predict(bag_train, newdata = data_test)
rf_test  <- predict(rf_train, newdata = data_test)
gbm_test <- predict(gbm_train, newdata = data_test)
xgb_test <- predict(xgboost_train, newdata = model.matrix(Disease ~ ., data_test)[,-1]) %>% 
   factor(
      x = ifelse(. < 0.5, 1, 2), 
      levels = c(1,2), 
      labels = c("Healthy", "Disease"))

list(
   bag_test, 
   rf_test, 
   gbm_test, 
   xgb_test
) %>% 
   map(~ confusionMatrix(.x, data_test$Disease))

# Based on accuracy (0.72, 0.76, 0.73, 0.69) the random forest obtains the best
# result
```






