---
title: "Excersie 4"
author: "Florian van Leeuwen"
date: "10/5/2022"
output: pdf_document
---

# Supervised learning: Regression 1
```{r}
library(ISLR)
library(MASS)
library(tidyverse)
```
## 1. Create a linear model object called lm_ses using the formula medv ~ lstat and the Boston dataset.
```{r}
lm_ses <- lm(medv ~ lstat, Boston)
```

## 2. Use the function coef() to extract the intercept and slope from the lm_ses object. Interpret the slope coefficient.
```{r}
coef(lm_ses)
```

## 3. Use summary() to get a summary of the lm_ses object. What do you see? You can use the help file ?summary.lm.
```{r}
summary(lm_ses)
```

## 4. Save the predicted y values to a variable called y_pred
```{r}
y_pred <- predict(lm_ses)
```

## 5. Create a scatter plot with y_pred mapped to the x position and the true y value (Boston$medv) mapped to the y value. What do you see? What would this plot look like if the fit were perfect?
```{r}
plot(y_pred, Boston$medv)
```

## 6. Use the seq() function to generate a sequence of 1000 equally spaced values from 0 to 40. Store this vector in a data frame with (data.frame() or tibble()) as its column name lstat. Name the data frame pred_dat.
```{r}
pred_dat <- data.frame(lstat = seq(0,40,length.out = 1000))
```

## 7. Use the newly created data frame as the newdata argument to a predict() call for lm_ses. Store it in a variable named y_pred_new.
```{r}
y_pred_new <- predict(lm_ses, newdata = pred_dat)
```

## 8. Create a scatter plot from the Boston dataset with lstat mapped to the x position and medv mapped to the y position. Store the plot in an object called p_scatter.
```{r}
p_scatter <- ggplot(Boston, aes(x = lstat, y = medv)) +
  geom_point()
```

## 9. Add the vector y_pred_new to the pred_dat data frame with the name medv.
```{r}
pred_dat["medv"] <- y_pred_new
```

## 10. Add a geom_line() to p_scatter, with pred_dat as the data argument. What does this line represent?
```{r}
p_scatter + geom_line(data = pred_dat)
```

## 11. The interval argument can be used to generate confidence or prediction intervals. Create a new object called y_pred_95 using predict() (again with the pred_dat data) with the interval argument set to “confidence”. What is in this object?
```{r}
y_pred_95 <-  as.data.frame(predict(lm_ses, newdata = pred_dat, interval = "confidence"))
```

## 12. Create a data frame with 4 columns: medv, lstat, lower, and upper.
```{r}
gg_pred <- tibble(
  lstat = pred_dat$lstat,
  medv  = y_pred_95[, 1],
  lower = y_pred_95[, 2],
  upper = y_pred_95[, 3]
)

```

## 13. Add a geom_ribbon() to the plot with the data frame you just made. The ribbon geom requires three aesthetics: x (lstat, already mapped), ymin (lower), and ymax (upper). Add the ribbon below the geom_line() and the geom_points() of before to make sure those remain visible. Give it a nice colour and clean up the plot, too!

```{r}
# Create the plot
Boston %>% 
  ggplot(aes(x = lstat, y = medv)) + 
  geom_ribbon(aes(ymin = lower, ymax = upper), data = gg_pred, fill = "#00008b44") +
  geom_point(colour = "#883321") + 
  geom_line(data = pred_dat, colour = "#00008b", size = 1) +
  theme_minimal() + 
  labs(x    = "Proportion of low SES households",
       y    = "Median house value",
       title = "Boston house prices")


```

## 14. 
It is the 95% CI of the estimation

## 15. Do the same thing, but now with the prediction interval instead of the confidence interval.
```{r}

y_pred_95_2 <-  as.data.frame(predict(lm_ses, newdata = pred_dat, interval = "prediction"))

gg_pred2 <- tibble(
  lstat = pred_dat$lstat,
  medv  = y_pred_95_2[, 1],
  lower = y_pred_95_2[, 2],
  upper = y_pred_95_2[, 3]
)


Boston %>% 
  ggplot(aes(x = lstat, y = medv)) + 
  geom_ribbon(aes(ymin = lower, ymax = upper), data = gg_pred2, fill = "#00008b44") +
  geom_point(colour = "#883321") + 
  geom_line(data = pred_dat, colour = "#00008b", size = 1) +
  theme_minimal() + 
  labs(x    = "Proportion of low SES households",
       y    = "Median house value",
       title = "Boston house prices")


```

## 16. Write a function called mse() that takes in two vectors: true y values and predicted y values, and which outputs the mean square error
```{r}

mse <- function(y_true, y_pred) {
  mean((y_true - y_pred)^2)
}
  # your function here
```

## 17. Make sure your mse() function works correctly by running the following code.
```{r}
mse(1:10, 10:1)
```

## 18. Calculate the mean square error of the lm_ses model. Use the medv column as y_true and use the predict() method to generate y_pred.
```{r}

mse(Boston$medv, predict(lm_ses))
```

## 19. The Boston dataset has 506 observations. Use c() and rep() to create a vector with 253 times the word “train”, 152 times the word “validation”, and 101 times the word “test”. Call this vector splits.
```{r}
splits <- c(rep("train", 253), rep("validation", 152), rep("test", 101))
```

## 20 Use the function sample() to randomly order this vector and add it to the Boston dataset using mutate(). Assign the newly created dataset to a variable called boston_master.
```{r}
boston_master <- Boston %>% 
  mutate(splits = sample(splits))
```

## 21. Now use filter() to create a training, validation, and test set from the boston_master data. Call these datasets boston_train, boston_valid, and boston_test.
```{r}
boston_train <- boston_master %>%
  filter(splits == "train")

boston_valid <- boston_master %>%
  filter(splits == "validation")

boston_test <- boston_master %>%
  filter(splits == "test")
```

## 22. Train a linear regression model called model_1 using the training dataset. Use the formula medv ~ lstat like in the first lm() exercise. Use summary() to check that this object is as you expect.
```{r}
model_1 <- lm(medv ~ lstat, boston_train)
summary(model_1)
```

## 23. Calculate the MSE with this object. Save this value as model_1_mse_train.
```{r}
model_1_mse_train <- mse(boston_train$medv, predict(model_1))
```

## 24. Now calculate the MSE on the validation set and assign it to variable model_1_mse_valid. Hint: use the newdata argument in predict().
```{r}
model_1_mse_valed <- mse(boston_valid$medv, predict(model_1, newdata = boston_valid))
```

## 25. Create a second model model_2 for the train data which includes age and tax as predictors. Calculate the train and validation MSE.

```{r}
model_2 <- lm(medv ~ lstat + age + tax, boston_train)
model_1_mse_valed <- mse(boston_train$medv, predict(model_2))
model_1_mse_valed <- mse(boston_valid$medv, predict(model_2, newdata = boston_valid))
```

## 26. Compare model 1 and model 2 in terms of their training and validation MSE. Which would you choose and why?

```{r}
print(c(model_1_mse_train,model_1_mse_valed,model_1_mse_valed,model_1_mse_valed))
```

## 27. Calculate the test MSE for the model of your choice in the previous question. What does this number tell you?
```{r}
sqrt(mse(boston_test$medv, predict(model_2, newdata = boston_test)))
```

## 28. Create a function that performs k-fold cross-validation for linear models.
```{r}
cross_val <- function(formula, dataset, k){
  n_samples <- nrow(dataset)
  select_vec <- rep(1:k, length.out = n_samples)
  data_split <- dataset %>% mutate(folds = sample(select_vec))
  
  mses <- rep(0, k)
  for (i in 1:k) {
    # split the data in train and validation set
    data_train <- data_split %>% filter(folds != i)
    data_valid <- data_split %>% filter(folds == i)
    
    # calculate the model on this data
    model_i <- lm(formula = formula, data = data_train)
    
    # Extract the y column name from the formula
    y_column_name <- as.character(formula)[2]
    
    # calculate the mean square error and assign it to mses
    mses[i] <- mse(y_true = data_valid[[y_column_name]],
                   y_pred = predict(model_i, newdata = data_valid))
  }
    # now we have a vector of k mse values. All we need is to
  # return the mean mse!
  mean(mses)
}


```
## Use your function to perform 9-fold cross validation with a linear model with as its formula medv ~ lstat + age + tax. Compare it to a model with as formulat medv ~ lstat + I(lstat^2) + age + tax

```{r}
cross_val(formula = medv ~ lstat + age + tax, dataset = Boston, k = 9)
cross_val(formula = medv ~ lstat + I(lstat^2) + age + tax, dataset = Boston, k = 9)
```

