---
title: "SLV practical 3"
author: "Amalia Tsakali"
date: "2022-10-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(magrittr)
library(mice)
library(DAAG)
```
##the boys dataset
```{r}
head(boys)
tail(boys)
!is.unsorted(boys$age)
summary(boys)
```

1. Create a histogram of the variable age using the function geom_histogram().
```{r 1}
boys %>% ggplot(aes(x=age))+geom_histogram()
```
2. Create a bar chart of the variable gen using the function geom_bar().
```{r 2}
boys %>% ggplot(aes(x=gen))+geom_bar()
```
##Assessing missing data
```{r}
md.pattern(boys)
```
3. Create a missingness indicator for the variables gen, phb and tv.
```{r 3}
boys_mising <- boys %>%
  mutate(gen_mis = is.na(gen),
         phb_mis = is.na(phb),
         tv_mis  = is.na(tv))
```
4. Assess whether missingness in the variables gen, phb and tv is related to someones age
```{r 4}
boys_mising %>% group_by(gen_mis) %>% summarise(mean(age))
boys_mising %>% group_by(phb_mis) %>% summarise(mean(age))
boys_mising %>% group_by(tv_mis) %>% summarise(mean(age))
```
5. Create a histogram for the variable age, faceted by whether or not someone has a missing value on gen.
```{r 5}
boys_mising %>%ggplot(aes(x=age))+geom_histogram(fill="green")+facet_wrap(~gen_mis)
```
6. Create a scatterplot with age on the x-axis and bmi on the y-axis, using the function geom_point().
```{r 6}
boys_mising%>%ggplot(aes(x=age, y=bmi))+geom_point()
```
7. Add a colour aesthetic to the previous plot using the missingness indicator of the variable gen.
```{r 7}
boys_mising%>%ggplot(aes(x=age, y=bmi))+geom_point(aes(colour=gen_mis))
```
##Visualizing the boys data
8. Visualize the relationship between reg (region) and age using a boxplot.
```{r 8}
boys_mising%>%ggplot(aes(x=reg, y=age))+geom_boxplot()
```
9. Create a density plot of age, splitting the densities by gen using the fill aesthetic.
```{r 9}
boys_mising%>%ggplot(aes(x=age))+geom_density(aes(fill=gen), alpha=0.5)+scale_fill_brewer() 
```
10. Create a diverging bar chart for hgt in the boys data set, that displays for every age year that year’s mean height in deviations from the overall average hgt.
```{r 10}
boys %>% mutate(age_round=cut(age, 0:22, labels = paste0(0:21, " years")),
         height_deviation = hgt - mean(hgt, na.rm = TRUE)) %>% group_by(age_round) %>% summarize(height_deviation = mean(height_deviation, na.rm = TRUE)) %>%
  mutate(color = ifelse(height_deviation > 0, "Above average", "Below average")) %>%
  ggplot(aes(x = height_deviation, y = age_round, fill = color)) +
  geom_bar(stat = "identity") +
  theme_minimal() 


```
##Regression visualization
11. Load the data elastic1 and elastic2 and bind the data frames together using the function bind_rows() and add a grouping variable indicating whether an observation comes from elastic1 or from elastic2.
```{r 11}
elastic <- bind_rows("Elastic1" = elastic1,
                     "Elastic2" = elastic2,
                     .id = "Set")

```
12. Create a scatterplot mapping stretch on the x-axis and distance on the y-axis, and map the just created group indicator as the color aesthetic.
```{r 12}
elastic %>% ggplot(aes(x=stretch, y=distance, colour=Set))+geom_point()+theme_minimal() 
```

13. Recreate the previous plot, but now assess whether the results of the two data sets appear consistent by adding a linear regression line.

```{r}
elastic %>% ggplot(aes(x=stretch, y=distance, colour=Set))+geom_point()+geom_smooth(method = "lm")+theme_minimal()
```

14. For each of the data sets elastic1 and elastic2, fit a regression model with y = distance on x = stretch using lm(y ~ x, data).
```{r 14}
fit1<-lm(distance ~ stretch, elastic1)
fit2<-lm(distance ~ stretch, elastic2)
```

15. For both of the previously created fitted models, determine the fitted values and the standard errors of the fitted values, and the proportion explained variance R2.
```{r 15}
predict(fit1, se.fit = TRUE)
predict(fit2, se.fit = TRUE)
fit1 %>% summary()
fit2 %>% summary()
```
16. Study the residual versus leverage plots for both models
```{r 16}
fit1 %>% plot(which = 5)
fit2 %>% plot(which = 5)
```
17. Use the elastic2 variable stretch to obtain predictions on the model fitted on elastic1.
```{r 17}
prediction1 <- predict(fit1, newdata = elastic2)
```

18. Now make a scatterplot to investigate similarity between the predicted values and the observed values for elastic2.
```{r 18}
prediction_dat <- 
  data.frame(distance = prediction1, 
             stretch  = elastic2$stretch) %>%
  bind_rows(Predicted = .,
            Observed  = elastic2, 
            .id = "Predicted")

prediction_dat %>%
  ggplot(aes(stretch, distance, col = Predicted)) +
  geom_point() + 
  geom_smooth(method = "lm") +
  theme_minimal() 
```

