---
title: "Practical 3 Exploratory Data Analysis" 
subtitle: "Supervised Learning & Visualization"
author: "Ernst Paul Swens"
date: "`r format(Sys.time(), '%B, %Y')`"
output: 
   html_document:
      toc: true
      number_sections: false
      toc_float:
         collapsed: true
         smooth_scroll: true
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(magrittr)
library(mice)
library(DAAG)
```

# Q1

**Create a histogram of the variable age using the function `geom_histogram()`.** 

```{r}
ggplot(boys, aes(x = age)) +
   geom_histogram(bins = 30) + 
   theme_minimal()
```


# Q2

**Create a bar chart of the variable gen using the function `geom_bar()`** 

```{r}
ggplot(boys, aes(x = gen)) +
   geom_bar() + 
   theme_minimal()
```


# Q3

**Create a missingness indicator for the variables `gen`, `phb` and `tv`.** 

```{r}
boys.2 <- boys %>% 
   mutate(
      rgen = as.factor(ifelse(is.na(gen), 1, 0)),
      rphb = as.factor(ifelse(is.na(phb), 1, 0)),
      rtv = as.factor(ifelse(is.na(tv), 1, 0))
   )
```


# Q4

**Assess whether missingness in the variables `gen`, `phb` and `tv` is related to someones `age`.** 

```{r}
boys.2 %>%
   summarise(
      gen = cor(as.numeric(rgen), age),
      phb = cor(as.numeric(rphb), age),
      tv = cor(as.numeric(rtv), age)
   )
```

# Q5

**Create a histogram for the variable `age`, faceted by whether or not someone has a missing value on gen.** 

```{r}
ggplot(boys.2, aes(x = age, fill = rgen)) + 
   geom_density(alpha = 0.3) +
   theme_minimal()

ggplot(boys.2, aes(x = age, fill = rphb)) + 
   geom_density(alpha = 0.3) +
   theme_minimal()

ggplot(boys.2, aes(x = age, fill = rtv)) + 
   geom_density(alpha = 0.3) +
   theme_minimal() 
```


# Q6

**Create a scatterplot with age on the x-axis and bmi on the y-axis, using the function `geom_point()`.**

```{r}
boys.2 %>% ggplot(aes(x = age, y = bmi)) + 
   geom_point() +
   theme_minimal()
```


# Q7

**Add a colour aesthetic to the previous plot using the missingness indicator of the variable gen.**

```{r}
boys.2 %>% ggplot(aes(x = age, y = bmi, color = rgen)) + 
   geom_point() +
   theme_minimal()
```


# Q8 

**Visualize the relationship between reg (region) and age using a boxplot.**

```{r}
boys.2 %>% ggplot(aes(x = reg, y = age)) +
   geom_boxplot() + 
   theme_minimal()
```


# Q9

**Create a density plot of age, splitting the densities by gen using the fill aesthetic.**

```{r}
boys.2 %>% ggplot(aes(x = age, fill = gen)) + 
   geom_density(alpha = 0.2) + 
   theme_minimal()
```

# Q10

**Create a diverging bar chart for hgt in the boys data set, that displays for every age year that year’s mean height in deviations from the overall average hgt.**

```{r}
boys.2 %>% mutate(age = round(age),
                  hgt = hgt - mean(boys.2$hgt, na.rm = TRUE)) %>%
   group_by(age) %>%
   summarise(hgt = mean(hgt, na.rm = TRUE)) %>%
   mutate(group = ifelse(hgt < 1, 0, 1)) %>%
   ggplot(aes(x = age, y = hgt, fill = as.factor(group))) + 
   geom_bar(stat = "identity") + 
   theme_minimal()
```

# Q11

**Load the data elastic1 and elastic2 and bind the data frames together using the function bind_rows() and add a grouping variable indicating whether an observation comes from elastic1 or from elastic2.**

```{r}
elastic <- bind_rows(
   "Elastic1" = elastic1,
   "Elastic2" = elastic2,
   .id = "Set"
)
```


# Q12

**Create a scatterplot mapping stretch on the x-axis and distance on the y-axis, and map the just created group indicator as the color aesthetic.**

```{r}
elastic %>%
  ggplot(aes(x = stretch, y = distance, col = Set)) +
  geom_point() +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() 
```


# Q13

**Recreate the previous plot, but now assess whether the results of the two data sets appear consistent by adding a linear regression line.**

```{r}
elastic %>%
   ggplot(aes(x = stretch, y = distance, col = Set)) +
   geom_point() +
   geom_smooth(method = "lm", se = FALSE) +
   scale_color_brewer(palette = "Set1") +
   theme_minimal()
```

# Q14

**For each of the data sets `elastic1` and `elastic2`, fit a regression model with y = distance on x = stretch using `lm(y ~ x, data)`.**

```{r}
lm.1 <- lm(distance ~ stretch, elastic1)
lm.2 <- lm(distance ~ stretch, elastic2)
```

# Q15

**For both of the previously created fitted models, determine the fitted values and the standard errors of the fitted values, and the proportion explained variance $R^2$.**

```{r}
predict(lm.1, se.fit = TRUE)
predict(lm.2, se.fit = TRUE)

lm.1 %>% summary() %$% r.squared
lm.2 %>% summary() %$% r.squared
```

# Q16

**Study the residual versus leverage plots for both models.**

```{r message=FALSE, warning=FALSE}
plot(lm.1, which = 5)
lm.1$residuals

plot(lm.2, which = 5)
lm.2$residuals
```

# Q17

**Use the `elastic2` variable stretch to obtain predictions on the model fitted on `elastic1`.**

```{r}
pred <- predict(lm.1, newdata = elastic2)
```

# Q18

**# 18. Now make a scatterplot to investigate similarity between the predicted values and the observed values for `elastic2`.**

```{r fig.height=7, fig.width=7}
data.frame(pred = pred,
           obs = elastic2$distance) %>%
   ggplot(aes(x = obs, y = pred)) + 
   geom_point() +
   geom_abline(intercept = 0, slope = 1) +
   theme_minimal()
```

