---
title: "SLV Practical 3"
author: "Shannon Dickson"
date: "`r format(Sys.Date(), '%B %d %Y')`"
output: 
   bookdown::html_document2:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: false
    theme: paper
---

<style type="text/css">
  
body{ /* Normal  */
  font-size: 12px;
  }
td {  /* Table  */
  font-size: 12px;
}
h1.title {
  font-size: 18px;
  color: DarkBlue;
}
h1 { /* Header 1 */
  font-size: 18px;
}
h2 { /* Header 2 */
  font-size: 18px;
}
h3 { /* Header 3 */
  font-size: 18px;
}
code.r{ /* Code block */
  font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
  font-size: 14px;
}
</style>

---

```{r setup, echo = FALSE}
library(knitr)
## Define an asis engine that will evaluate inline code within an asis block:
knit_engines$set(asis = function(options) {
  if(options$echo && options$eval) knit_child(text = options$code)
}
)

knitr::opts_chunk$set(echo = params$answers, 
                      message = FALSE, 
                      warning = FALSE)

hook_output <- knitr::knit_hooks$get("output")

# set a new output hook to truncate text output
knitr::knit_hooks$set(output = function(x, options) {
  if (!is.null(n <- options$out.lines)) {
    x <- xfun::split_lines(x)
    if (length(x) > n) {
      # truncate the output
      x <- c(head(x, n), "....\n")
    }
    x <- paste(x, collapse = "\n")
  }
  hook_output(x, options)
})
```


# Preparation

```{r}
library(tidyverse)
library(magrittr)
library(mice)
library(DAAG)
```

# Data

Look at the start:

```{r}
head(boys)
```
Look at the end:

```{r}
tail(boys)
```
This checks if the data is sorted by age, since age looks to be in ascending order. 

```{r}
!is.unsorted(boys$age)
```

Look at a summary:

```{r}
summary(boys)
```

# Exercises

**1. Create a histogram of the variable age using the function geom_histogram().**

* More younger boys 
* More boys between 10 -20
* Dip around 5 years old 

```{r}
boys %>% 
  ggplot(aes(x = age)) +
  geom_histogram(fill = "darkseagreen") +
  labs(x = "Age (years)") +
  theme_bw() 
```
**2. Create a bar chart of the variable gen using the function geom_bar().**

* Lots of NA values, especially relative to the generations

```{r}
boys %>% 
  ggplot(aes(x = gen)) +
  geom_bar(fill = "darkseagreen") +
  theme_bw()
```
## Assessing missing data

* 233 cases with no missing values
* 525 observations with missing value(s)
* 1622 missing values in total 
* Most missing values in gen, phb, tv 

```{r}
md.pattern(boys)
```

**3.  Create a missingness indicator for the variables gen, phb and tv.**

The code below creates a data frame with variables detailing the missing values per row. 

```{r}
boys_missings <- boys %>%
        mutate(gen_na = is.na(gen),
               phb_na = is.na(phb),
               tv_na  = is.na(tv))
```

**4. Assess whether missingness in the variables gen, phb and tv is related to someones age.**

I will look at the mean as a measure of relation. 

* Age is much lower when there are missing values. 

```{r}
boys_missings %>% 
  group_by(gen_na) %>% 
  summarise(age = mean(age))

boys_missings %>% 
  group_by(phb_na) %>% 
  summarise(age = mean(age))

boys_missings %>% 
  group_by(tv_na) %>% 
  summarise(age = mean(age))
```

**5. Create a histogram for the variable age, faceted by whether or not someone has a missing value on gen.**

* Boys with observed values are at least 7 years old
* Boys with missing values are commonly between 0 = 5 years. 

```{r}
boys_missings %>%
  ggplot(aes(age)) +
  geom_histogram(fill = "darkseagreen") + 
  facet_wrap(~gen_na) +
  theme_minimal()
```

**6. Create a scatterplot with age on the x-axis and bmi on the y-axis, using the function geom_point().**

```{r}
boys_missings %>% 
  ggplot(aes(age, bmi)) +
  geom_point(colour = "darkorange", alpha = 0.5) +
  theme_bw()
```

**7. Add a colour aesthetic to the previous plot using the missingness indicator of the variable gen.**

* More missing values on gen for younger boys (yellow colours to the left)
* No real relationship between bmi and gen

```{r}
boys_missings %>% 
  ggplot(aes(age, bmi, colour = gen_na, alpha = 0.5)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw()
```

# Visualising the boys data

**8. Visualize the relationship between reg (region) and age using a boxplot.**

* Boys who are younger do not have an observed value for religion

```{r}
boys %>% 
  ggplot(aes(reg, age)) +
  geom_boxplot(fill = "darkseagreen") +
  theme_bw()
```

**9. Create a density plot of age, splitting the densities by gen using the fill aesthetic.**

```{r}
boys %>% 
  ggplot(aes(age, fill = gen)) +
  geom_density(alpha = 0.5) +
  scale_fill_brewer() +
  theme_bw()
```

**10. Create a diverging bar chart for hgt in the boys data set, that displays for every age year that year’s mean height in deviations from the overall average hgt.**

* Boys reach average height around 7 years old 

```{r}
# First make a category for age, center height 
boys %>%
  mutate(Age = cut(age, 0:22, labels = paste0(0:21, " years")),
         Height = hgt - mean(hgt, na.rm = TRUE)) %>%
  group_by(Age) %>%
  summarize(Height = mean(Height, na.rm = TRUE)) %>%
  mutate(color = ifelse(Height > 0, "Above average", "Below average")) %>% 
# Now plot
  ggplot(aes(x = Height, y = Age, fill = color)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Dark2") +
  theme_minimal() +
  theme(legend.title = element_blank())
```

# Regression Visualisation

**11. Load the data elastic1 and elastic2 and bind the data frames together using the function bind_rows() and add a grouping variable indicating whether an observation comes from elastic1 or from elastic2.**

```{r}
elastic <- bind_rows("Elastic1" = elastic1,
                     "Elastic2" = elastic2,
                     .id = "Set")
```

**12. Create a scatterplot mapping stretch on the x-axis and distance on the y-axis, and map the just created group indicator as the color aesthetic.**

```{r}
elastic %>% 
  ggplot(aes(stretch, distance, colour = Set)) +
  geom_jitter() +
  scale_color_brewer(palette = "Dark2") +
  theme_bw()
```

**13. Recreate the previous plot, but now assess whether the results of the two data sets appear consistent by adding a linear regression line**

* Elastic1 and elastic 2 are consistent with each other (same regression line)
* Elastic1 has a shorter range 

```{r}
elastic %>% 
  ggplot(aes(stretch, distance, colour = Set)) +
  geom_point() +
  geom_smooth(method = lm) +
  scale_color_brewer(palette = "Dark2") +
  theme_bw()
```

**14. For each of the data sets elastic1 and elastic2, fit a regression model with y = distance on x = stretch using lm(y ~ x, data).**

```{r}
m1 <- lm(distance ~ stretch, elastic1)

m2 <- lm(distance ~ stretch, elastic2)
```

**15. For both of the previously created fitted models, determine the fitted values and the standard errors of the fitted values, and the proportion explained variance R2.**

```{r}
# Fitted values
predict(m1, se = TRUE)
predict(m1, se = TRUE)

# R^2
summary(m1)
summary(m2)
```

**16. Study the residual versus leverage plots for both models.**

* In model 1, case 7 is the most influential point.
* In model 2, case 3 is the most influential point. 

```{r}
plot(m1, which = 5)
plot(m2, which = 5)
```

**17. Use the elastic2 variable stretch to obtain predictions on the model fitted on elastic1.**

```{r}
pred<- predict(m1, newdata = elastic2)
```

**18. Now make a scatterplot to investigate similarity between the predicted values and the observed values for elastic2.**

First step is to create a dataframe for the plot. 

```{r}
pred_data <- data.frame(distance = pred, 
                       stretch  = elastic2$stretch) %>%
  bind_rows(Predicted = .,
            Observed  = elastic2, 
            .id = "Predicted")
```

The plot shows that the predicted and observed values are very similar. 

```{r}
pred_data %>% 
      ggplot(aes(stretch, distance, colour = Predicted)) +
      geom_point(alpha = 0.7) +
      geom_smooth(method = lm) +
      scale_color_brewer(palette = "Dark2") +
      theme_bw()
```
# End of practical

