---
title: "Practical 1 - Simona Cernat"

---



```{r load_packages, message = FALSE, warning = FALSE}
library(ISLR)
library(tidyverse)
library(haven)
library(readxl)
```





---

1. __Run the following code in `R` and inspect their data types using the `class()` function. Try to guess beforehand what their types will be!__

---

```{r 1}
object_1 <- 1:5
object_2 <- 1L:5L
object_3 <- "-123.456"
object_4 <- as.numeric(object_2)
object_5 <- letters[object_1]
object_6 <- as.factor(rep(object_5, 2))
object_7 <- c(1, 2, 3, "4", "5", "6")

```

object 1 and 2 - integers; object 3- string, 4- numeric, 5 - chraracter, 6 - factor, 7 - vector with many types of objects

---
```{r}
class(object_1)
class(object_2)
class(object_3)
class(object_4)
class(object_5)
class(object_6)
```

2. __Convert `object_7` back to a vector of numbers using the `as.numeric()` function__

---

```{r 2}
object_7 = as.numeric(object_7)
object_7
```



3. __Make a list called `objects` containing object 1 to 7 using the `list()` function.__

---

```{r 3}
objects = list(object_1, object_2, object_3, object_4, object_5, object_6, 
                object_7)
```


4. __Make a data frame out of `object_1`, `object_2`, and `object_5` using the `data.frame()` function__

---

```{r 4}
df = data.frame(object_1, object_2, Var3 = object_5)
 df
```



5. __Useful functions for determining the size of a data frame are `ncol()` and `nrow()`. Try them out!__

---

```{r 5}
ncol(df)
nrow(df)
```



6. __Use the function `read_csv()` to import the file "data/googleplaystore.csv" and store it in a variable called `apps`.__

---

```{r readcsv}


 apps <- read_csv("./googleplaystore.csv")
head(apps) 
```



---

7. __Did any column get a variable type you did not expect?__

---

```{r ok}
# Some columns have the wrong data types such as Size, Price which should be integers and  Last update (should be date) 
```

---

8. __Use the function `head()` to look at the first few rows of the `apps` dataset__

---

```{r head}
head(apps)
```


---

9. __Repeat steps 5, 6, and 7 but now for "data/students.xlsx" (NB: You'll need a function from the package `readxl`). Also try out the function `tail()` and `View()` (with a capital V).__

---

```{r readxl}
students <- read_excel("./students.xlsx")
tail(students)
view(students)
```

```{r}
# Here the data types are correct.
```
---

10. __Create a summary of the three columns in the students dataset using the `summary()` function. What is the range of the grades achieved by the students?__

---

```{r summary}
summary(students)
```




11. __Look at the help pages for `filter()` (especially the examples) and show the students with a grade lower than 5.5__

---

```{r filter1}
filter(students, grade < 5.5)
```

---

12. __Show only the students with a grade higher than 8 from programme A__

---


```{r filter2}
filter(students, grade > 8, programme == "A")
```



13. __Sort the students dataset such that the students from programme A are on top of the data frame and within the programmes the highest grades come first.__

---

```{r arrange}
arrange(students, programme, -grade)
```




14. __Show only the `student_number` and `programme` columns from the students dataset__ 

---

```{r select}
select(students, student_number, programme)

```


15. __Use `mutate()` and `recode()` to change the codes in the programme column of the students dataset to their names. Store the result in a variable called `students_recoded`__

---

```{r recode}
students_recoded <- mutate(students, 
                           programme = recode(programme, 
                                              "A" = "Science", "B" = "Social Science")
)

students_recoded
```



16. __Create a data processing pipeline that (a) loads the apps dataset, (b) parses the number of installs as 'Downloads' variable using `mutate` and `parse_number()`, (c) shows only apps with more than 500 000 000 downloads, (d) orders them by rating (best on top), and (e) shows only the relevant columns (you can choose which are relevant, but select at least the `Rating` and `Category` variables). Save the result under the name `popular_apps`.__



```{r bestapps, message = FALSE,}
popular_apps <- read_csv("./googleplaystore.csv") %>%
  mutate(Downloads = parse_number(Installs)) %>%
  filter(Downloads > 500000000) %>%
  arrange(-Rating ) %>%
  select(App, Rating, Category, Installs)

popular_apps
```



17. __Show the median, minimum, and maximum for the popular apps dataset you made in the previous assignment.__



```{r rating}
popular_apps %>% 
  summarise(
    med = median(Rating),
    min = min(Rating), 
    max = max(Rating)
  )

```




18. __Add the median absolute deviation to the summaries you made before__




```{r rating2}
mad <- function(x) {
  median(abs(x - median(x)))
}

popular_apps %>% 
  summarise(
    med = median(Rating),
    min = min(Rating), 
    max = max(Rating),
    mad = mad(Rating)
  )
```


19. __Create a grouped summary of the ratings per category in the popular apps dataset.__

---

```{r group2}
popular_apps %>% 
  group_by(Category) %>%
  summarise(
    med = median(Rating),
    mad = mad(Rating),
    min= min(Rating),
    max = max(Rating)
  )
```


20. __Create an interesting summary based on the Google play store apps dataset. An example could be "do games get higher ratings than communication apps?"__



```{r final, message = FALSE}
popular_apps %>% 
  filter(Category == "SOCIAL"| Category == "FAMILY" ) %>%
  group_by(Category) %>%
  summarise(
    average = mean(Rating),
   
  )
```

