---
title: "SLV Practical 7"
author: "Alex Carriero"
output: html_document
---

# Introduction
```{r, warning = F, message = F}
library(MASS)
library(splines)
library(ISLR)
library(tidyverse)
```

```{r}
# seed 
set.seed(45)
```

# Prediction Plot
```{r}
Boston %>% 
  ggplot(aes(x = lstat, y = medv)) +
  geom_point() +
  theme_minimal()
```

## 1. Create a function called pred_plot() that takes as input an lm object, which outputs the above plot but with a prediction line generated from the model object using the predict() method.

```{r}
pred_plot <- function(model) {
  # First create predictions for all values of lstat
  x_pred <- seq(min(Boston$lstat), max(Boston$lstat), length.out = 500)
  y_pred <- predict(model, newdata = tibble(lstat = x_pred))
  
  # Then create a ggplot object with a line based on those predictions
  Boston %>%
    ggplot(aes(x = lstat, y = medv)) +
    geom_point() +
    geom_line(data = tibble(lstat = x_pred, medv = y_pred), size = 1, col = "blue") +
    theme_minimal()
}
```

## 2. Create a linear regression object called lin_mod which models medv as a function of lstat. Check if your prediction plot works by running pred_plot(lin_mod). Do you see anything out of the ordinary with the predictions?

```{r}
lin_mod <- lm(medv ~ lstat, data = Boston)
pred_plot(lin_mod)

# note: the predicted median housing value is below 0 for high values.
```

# Polynomial Regression 

## 3. Create another linear model pn3_mod, where you add the second and third-degree polynomial terms I(lstat^2) and I(lstat^3) to the formula. Create a pred_plot() with this model.

The function poly() can automatically generate a matrix which contains columns with polynomial basis function outputs.

```{r}
pn3_mod <- lm(medv ~ lstat + I(lstat^2) + I(lstat^3), data = Boston)
pred_plot(pn3_mod)
```

## 4. Play around with the poly() function. What output does it generate with the arguments degree = 3 and raw = TRUE?

```{r}
poly(1:5, degree = 3, raw = TRUE)
```


## 5. Use the poly() function directly in the model formula to create a 3rd-degree polynomial regression predicting medv using lstat. Compare the prediction plot to the previous prediction plot you made. What happens if you change the poly() function to raw = FALSE?
```{r}
pn3_mod2 <- lm(medv ~ poly(lstat, 3, raw = TRUE), data = Boston) # this is cool
pred_plot(pn3_mod2)

# This does not change the fitted values but you can see whether a certain order in the polynomial significantly improves the regression over the lower orders.
```

# Piecewise Regression 

## 6. Create a model called pw2_mod with one predictor: I(lstat <= median(lstat)). Create a pred_plot with this model. Use the coefficients in coef(pw2_mod) to find out what the predicted value for a low-lstat neighbourhood is.
```{r}
pw2_mod <- lm(medv ~ I(lstat <= median(lstat)), data = Boston)
pred_plot(pw2_mod)
```

```{r}
coef(pw2_mod)
# the predicted value for low-lstat neighbourhoods is 16.68 + 11.71 = 28.39
```

## 7. Use the cut() function in the formula to generate a piecewise regression model called pw5_mod that contains 5 equally spaced sections. Again, plot the result using pred_plot.
```{r}
pw5_mod <- lm(medv ~ cut(lstat, 5), data = Boston)
pred_plot(pw5_mod)
```
```{r}
table(cut(Boston$lstat, 5))
```


## 8. Optional: Create a piecewise regression model pwq_mod where the sections are not equally spaced, but have equal amounts of training data. Hint: use the quantile() function.
```{r}
brks <- c(-Inf, quantile(Boston$lstat, probs = c(.2, .4, .6, .8)), Inf)
pwq_mod <- lm(medv ~ cut(lstat, brks), data = Boston)
pred_plot(pwq_mod)
```
```{r}
table(cut(Boston$lstat, brks))
```

# Piecewise Polynomial regression 
## 9. This function does not have comments. Copy - paste the function and add comments to each line. To figure out what each line does, you can first create “fake” vec and knots variables, for example vec <- 1:20 and knots <- 2 and try out the lines separately.
```{r}
piecewise_cubic_basis <- function(vec, knots = 1) {
  # If there is only one section, just return the 3rd order polynomial
  if (knots == 0) return(poly(vec, degree = 3, raw = TRUE))
  
  # cut the vector
  cut_vec <- cut(vec, breaks = knots + 1)
  
  # initialise a matrix for the piecewise polynomial
  out <- matrix(nrow = length(vec), ncol = 0)
  
  # loop over the levels of the cut vector
  for (lvl in levels(cut_vec)) {
    
    # temporary vector
    tmp <- vec
    
    # set all values to 0 except the current section
    tmp[cut_vec != lvl] <- 0
    
    # add the polynomial based on this vector to the matrix
    out <- cbind(out, poly(tmp, degree = 3, raw = TRUE))
    
  }
  
  # return the piecewise polynomial matrix
  out
}
```


## 10. Create piecewise cubic models with 1, 2, and 3 knots (pc1_mod - pc3_mod) using this piecewise cubic basis function. Compare them using the pred_plot() function.
```{r}
pc1_mod <- lm(medv ~ piecewise_cubic_basis(lstat, 1), data = Boston)
pc2_mod <- lm(medv ~ piecewise_cubic_basis(lstat, 2), data = Boston)
pc3_mod <- lm(medv ~ piecewise_cubic_basis(lstat, 3), data = Boston)

pred_plot(pc1_mod)
pred_plot(pc2_mod)
pred_plot(pc3_mod)
```
# Splines 

## 11. Create a data frame called boston_tpb with the columns medv and lstat from the Boston dataset.

```{r}
boston_tpb <- Boston %>% as_tibble %>% select(medv, lstat)
```

## 12. Now use mutate to add squared and cubed versions of the lstat variable to this dataset.

```{r}
boston_tpb <- boston_tpb %>% mutate(lstat2 = lstat^2, lstat3 = lstat^3)
```


## 13. Use mutate to add a column lstat_tpb to this dataset which is 0 below the median and has value (lstat - median(lstat))^3 above the median. Tip: you may want to use ifelse() within your mutate() call.
```{r}
boston_tpb <- boston_tpb %>% 
  mutate(lstat_tpb = ifelse(lstat >  median(lstat), (lstat - median(lstat))^3, 0))
```

## 14. Create a linear model tpb_mod using the lm() function. How many predictors are in the model? How many degrees of freedom does this model have?
```{r}
tpb_mod <- lm(medv ~ lstat + lstat2 + lstat3 + lstat_tpb, data = boston_tpb)
summary(tpb_mod)

# this model has 5 predictors and also 5 degrees of freedom
# See ISLR page 273
```

## 15. Create a cubic spline model bs1_mod with a knot at the median using the bs() function. Compare its predictions to those of the tpb_mod using the predict() function on both models.
```{r}
bs1_mod <- lm(medv ~ bs(lstat, knots = median(lstat)), data = Boston)
summary(bs1_mod)
```

```{r}
# Comparing the predictions from the two models: negligible absolute difference
mean(abs(predict(bs1_mod) - predict(tpb_mod)))
```

## 16. Create a prediction plot from the bs1_mod object using the plot_pred() function.
```{r}
pred_plot(bs1_mod)
```
Note that this line fits very well, but at the right end of the plot, the curve slopes up. Theoretically, this is unexpected – always pay attention to which predictions you are making and whether that behaviour is in line with your expectations.

The last extension we will look at is the natural spline. This works in the same way as the cubic spline, with the additional constraint that the function is required to be linear at the boundaries. The ns() function from the splines package is for generating the basis representation for a natural spline.


## 17. Create a natural cubic spline model (ns3_mod) with 3 degrees of freedom using the ns() function. Plot it, and compare it to the bs1_mod.
```{r}
ns3_mod <- lm(medv ~ ns(lstat, df = 3), data = Boston)
pred_plot(ns3_mod)
```
## 18. Plot lin_mod, pn3_mod, pw5_mod, pc3_mod, bs1_mod, and ns3_mod and give them nice titles by adding + ggtitle("My title") to the plot. You may use the function plot_grid() from the package cowplot to put your plots in a grid.
```{r}
library(cowplot)
plot_grid(
  pred_plot(lin_mod) + ggtitle("Linear regression"),
  pred_plot(pn3_mod) + ggtitle("Polynomial"),
  pred_plot(pw5_mod) + ggtitle("Piecewise constant"),
  pred_plot(pc3_mod) + ggtitle("Piecewise cubic"),
  pred_plot(bs1_mod) + ggtitle("Cubic spline"),
  pred_plot(ns3_mod) + ggtitle("Natural spline")
)
```

# Programming Assignment 

## 19. Use 12-fold cross validation to determine which of the 6 methods (lin, pn3, pw5, pc3, bs1, and ns3) has the lowest out-of-sample MSE.
```{r}
# first create an mse function
mse <- function(y_true, y_pred) mean((y_true - y_pred)^2)

# add a 12 split column to the boston dataset so we can cross-validate
boston_cv <- Boston %>% mutate(split = sample(rep(1:12, length.out = nrow(Boston))))

# prepare an output matrix with 12 slots per method for mse values
output_matrix <- matrix(nrow = 12, ncol = 6) 
colnames(output_matrix) <- c("lin", "pn3", "pw5", "pc3", "bs1", "ns3")

# loop over the splits, run each method, and return the mse values
for (i in 1:12) {
  train <- boston_cv %>% filter(split != i)
  test  <- boston_cv %>% filter(split == i)
  
  brks <- c(-Inf, 7, 15, 22, Inf)
  
  lin_mod <- lm(medv ~ lstat,                            data = train)
  pn3_mod <- lm(medv ~ poly(lstat, 3),                   data = train)
  pw5_mod <- lm(medv ~ cut(lstat, brks),                 data = train)
  pc3_mod <- lm(medv ~ piecewise_cubic_basis(lstat, 3),  data = train)
  bs1_mod <- lm(medv ~ bs(lstat, knots = median(lstat)), data = train)
  ns3_mod <- lm(medv ~ ns(lstat, df = 3),                data = train)
  
  output_matrix[i, ] <- c(
    mse(test$medv, predict(lin_mod, newdata = test)),
    mse(test$medv, predict(pn3_mod, newdata = test)),
    mse(test$medv, predict(pw5_mod, newdata = test)),
    mse(test$medv, predict(pc3_mod, newdata = test)),
    mse(test$medv, predict(bs1_mod, newdata = test)),
    mse(test$medv, predict(ns3_mod, newdata = test))
  )
}

# this is the comparison of the methods
colMeans(output_matrix)
```

```{r}
# we can show it graphically too
tibble(names = as_factor(colnames(output_matrix)), 
       mse   = colMeans(output_matrix)) %>% 
  ggplot(aes(x = names, y = mse, fill = names)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  scale_fill_viridis_d(guide = "none") +
  labs(
    x     = "Method", 
    y     = "Mean squared error", 
    title = "Comparing regression method prediction performance"
  )
```

