---
title: "P7_Classification evaluation_Kyuri"
author: "Kyuri Park"
date: "`r format(Sys.time(), '%B %d, %Y')`"
urlcolor: blue
output: 
  html_document:
    toc: true
    toc_depth: 1
    toc_float: true
    theme: paper
    highlight: tango
    df_print: paged
---
<style type="text/css">
@import url('https://fonts.googleapis.com/css2?family=Lato:wght@300;400&display=swap');

body{ /* Normal  */
  font-size: 13px;
  font-family: 'Lato', sans-serif;
  }
h1.title {
  font-size: 25px;
  color: DarkBlue;
  margin-bottom:5px;
}

h1 { /* Header 1 */
  font-size: 20px;
  font-weight: bold;
}
h2 { /* Header 2 */
  font-size: 15px;
  line-height: 1.6;
}
h3 { /* Header 3 */
  font-size: 14px;
  line-height: 1.6;
}

pre { /* Code block - determines code spacing between lines */
  font-size: 13px;
}
</style>
<hr>

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
               warning = FALSE,
               comment = NA)

```

# Introduction

First, load the packages:

```{r packages}
library(MASS)
library(ISLR)
library(tidyverse)

library(pROC)

library(rpart)
library(rpart.plot)
library(randomForest)
```

Set the seed for reproducibility.
```{r}
set.seed(123)
```


# Confusion matrix, continued

## 1. Create a logistic regression model `lr_mod` for this data using the formula `response ~ .` and create a confusion matrix based on a .5 cutoff probability.

```{r}
cardio <- read.csv("data/cardiovascular_treatment.csv")

# check data
glimpse(cardio) # data type is not correct

# clean data
cardio <- cardio %>% 
  mutate(severity = as.factor(severity),
         gender = as.factor(gender),
         dose = as.factor(dose),
         response = as.factor(response))
```


```{r}
# fit lr 
lr_mod <- glm(response ~ ., family = "binomial", data = cardio)
# predicted probability
prob_lr <- predict(lr_mod, type = "response")
# prediction (cutoff = 0.5)
pred_lr <- ifelse(prob_lr > 0.5, 1, 0)

# confusion matrix
cm_lr <- table(true = cardio$response, prediction = pred_lr); cm_lr
```



#### Confusion matrix metrics

## 2. Calculate the accuracy, true positive rate (sensitivity), the true negative rate (specificity), the false positive rate, the positive predictive value, and the negative predictive value. You can use the [confusion matrix table on wikipedia](https://en.wikipedia.org/w/index.php?title=Sensitivity_and_specificity&oldid=862159646#Confusion_matrix). What can you say about the model performance? Which metrics are most relevant if this model were to be used in the real world?

Given the *accuracy = 0.7*, the model predicts about 70% correctly. 
***Interpretation***
```{r}
# check model performance
TN <- cm_lr[1,1]
TP <- cm_lr[2,2]
FN <- cm_lr[2,1]
FP <- cm_lr[1,2]

data.frame(
ACC = (TP + TN) / sum(cm_lr), # accuracy
TPR = TP / (TN + FN), # sensitivity (TPR)
TNR = TN / (TN + FP), # specificity (TNR)
FPR = FP / (FP + TN), # false positive rate
PPV = TP / (TP + FP), # positive predictive value (precision)
NPV = TN / (TN + FN) # negative predictive value  
) %>% round(2)
```

## 3. Create an LDA model `lda_mod` for the same prediction problem. Compare its performance to the LR model.

The performance is almost identical.
```{r}
lda_mod <- lda(response ~., data = cardio)
pred_lda <- predict(lda_mod)$class

cm_lda <- table(true = cardio$response, predicted = pred_lda); cm_lda

# check model performance
TN <- cm_lda[1,1]
TP <- cm_lda[2,2]
FN <- cm_lda[2,1]
FP <- cm_lda[1,2]

data.frame(
ACC = (TP + TN) / sum(cm_lda), # accuracy
TPR = TP / (TN + FN), # sensitivity (TPR)
TNR = TN / (TN + FP), # specificity (TNR)
FPR = FP / (FP + TN), # false positive rate
PPV = TP / (TP + FP), # positive predictive value (precision)
NPV = TN / (TN + FN) # negative predictive value  
) %>% round(2)
```


## 4. Compare the classification performance of `lr_mod` and `lda_mod` for the new patients in the `data/new_patients.csv`.

Again, the performance is almost identical.
```{r}
# load the new patients data
new_patients <- read.csv("data/new_patients.csv") %>% 
  mutate(severity = as.factor(severity),
         gender   = as.factor(gender),
         dose     = as.factor(dose),
         response = as.factor(response))

# predict based on new patients data
## logistic regression
prob_lr_new <- predict(lr_mod, newdata = new_patients, type = "response")
pred_lr_new <- ifelse(prob_lr_new > .5, 1, 0)

## LDA
pred_lda_new <- predict(lda_mod, newdata = new_patients)$class

# confusion matrix
## lr
cmat_lr_new <- table(true = new_patients$response, pred = pred_lr_new); cmat_lr_new
## lda
cmat_lda_new <- table(true = new_patients$response, pred = pred_lda_new); cmat_lda_new

```

---

#### Brier score


## Calculate the out-of-sample brier score for the `lr_mod` and give an interpretation of this number.

```{r}
brier_score <- mean((prob_lr_new - (as.numeric(new_patients$response) - 1)) ^ 2); brier_score
```
The lower the Brier score is for a set of predictions, the better the predictions are calibrated. In this case, the brier score is `r brier_score %>%  round(2)`, which is not too high.

---

#### ROC curve

## 5. Create two LR models: `lr1_mod` with `severity`, `age`, and `bb_score` as predictors, and `lr2_mod` with the formula `response ~ age + I(age^2) + gender + bb_score * prior_cvd * dose`. Save the predicted probabilities on the training data.
```{r}
lr1_mod <- glm(response ~ severity + bb_score + age, 
               family = "binomial", data = cardio)
prob_lr1 <- predict(lr1_mod, type = "response")

lr2_mod <- glm(response ~ age + I(age^2) + gender + bb_score * prior_cvd * dose, 
               family = "binomial", data = cardio)
prob_lr2 <- predict(lr2_mod, type = "response")
```

## 6. Use the function `roc()` from the `pROC` package to create two ROC objects with the predicted probabilities: `roc_lr1` and `roc_lr2`. Use the `ggroc()` method on these objects to create an ROC curve plot for each. Which model performs better? Why?
Model 2 is better as both sensitivity and speificity are higher across the whole range.
```{r}
roc_lr1 <- roc(cardio$response, prob_lr1)
roc_lr2 <- roc(cardio$response, prob_lr2)

ggroc(list(roc_lr1, roc_lr2), legacy.axes = TRUE, aes=c("color")) +
  scale_color_discrete(labels = c("roc_lr1", "roc_lr2")) + 
  theme_classic() + labs(color = "models") 

```

## 7. Print the `roc_lr1` and `roc_lr2` objects. Which AUC value is higher? How does this relate to the plots you made before? What is the minimum AUC value and what would a "perfect" AUC value be and how would it look in a plot?
AUC of `roc_lr2` is higher (0.74) than AUC of `roc_lr1` (0.63). 
Perfect AUC would be 1, which has specificity of 1 as well as sensitivity of 1 (see the plot below).
```{r}
roc_lr1; roc_lr2

# perfect AUC 
ggplot(data.frame(x = c(0, 0, 1), y = c(0, 1, 1)), 
       aes(x = x, y = y)) +
  geom_line() +
  labs(y = "sensitivity", 
       x = "1 - specificity", 
       title = "Perfect model") +
  theme_classic()
```


# Iris dataset

```{r iris}

# fit lda model, i.e. calculate model parameters
lda_iris <- lda(Species ~ ., data = iris)

# use those parameters to compute the first linear discriminant
first_ld <- -c(as.matrix(iris[, -5]) %*% lda_iris$scaling[,1])

# plot
tibble(
  ld = first_ld,
  Species = iris$Species
) %>% 
  ggplot(aes(x = ld, fill = Species)) +
  geom_histogram(binwidth = .5, position = "identity", alpha = .9) +
  scale_fill_viridis_d() +
  theme_minimal() +
  labs(
    x = "Discriminant function",
    y = "Frequency", 
    main = "Fisher's linear discriminant function on Iris species"
  ) + 
  theme(legend.position = "top")

```


## 8. Explore the iris dataset using summaries and plots.
```{r}
summary(iris)

library(GGally)
ggpairs(iris)
```

## 9. Fit an additional LDA model, but this time with only `Sepal.Length` and `Sepal.Width` as predictors. Call this model `lda_iris_sepal`.

```{r}
lda_iris_sepal <- lda(Species ~ Sepal.Length + Sepal.Width, data = iris)
```

## 10. Create a confusion matrix of the `lda_iris` and `lda_iris_sepal` models. (NB: we did not split the dataset into training and test set, so use the training dataset to generate the predictions.). Which performs better in terms of accuracy?
cm_iris performs better!
```{r}
cm_iris <- table(true = iris$Species, predicted = predict(lda_iris)$class)
cm_sepal <- table(true = iris$Species, predicted = predict(lda_iris_sepal)$class)

cm_iris; cm_sepal
```


# Classification trees

Classification trees in `R` can be fit using the `rpart()` function.


## 11. Use `rpart()` to create a classification tree for the `Species` of `iris`. Call this model `iris_tree_mod`. Plot this model using `rpart.plot()`.
```{r}
iris_tree_mod <- rpart(Species ~., data = iris)
rpart.plot(iris_tree_mod)
```


## 12. How would an iris with 2.7 cm long and 1.5 cm wide petals be classified?

Will be classified as `versicolor` according to the tree.


## 13. Create a scatterplot where you map `Petal.Length` to the x position and `Petal.Width` to the y position. Then, manually add a vertical and a horizontal line (using `geom_segment`) at the locations of the splits from the classification tree. Interpret this plot.
```{r}
iris %>% 
ggplot(aes(x = Petal.Length, y = Petal.Width, color= Species)) +
  geom_point() + 
  geom_segment(aes(x = 2.5, xend = 2.5, y = -Inf, yend = Inf),
               color = "gray") +
  geom_segment(aes(x = 2.5, xend = Inf, y = 1.75, yend = 1.75), 
               color = "gray") +
  theme_minimal()
```

```{r}
# check the control parameters
?rpart.control
```


## 14. Create a classification tree model where the splits continue until all the observations have been classified. Call this model `iris_tree_full_mod`. Plot this model using `rpart.plot()`. Do you expect this model to perform better or worse on new Irises?
```{r}
iris_tree_full_mod <- rpart(Species ~., data = iris, control = rpart.control(minsplit =1, cp=0 ))
rpart.plot(iris_tree_full_mod)
```



# Final assignment: Random forest for classification

## 15. Use the function `randomForest()` to create a random forest model on the iris dataset. Use the function `importance()` on this model and create a bar plot of variable importance. Does this agree with your expectations? How well does the random forest model perform compared to the `lda_iris` model?


```{r}
# randomforest model
rf_mod <- randomForest(Species ~ ., data = iris)
# variable importance
var_imp <- importance(rf_mod)

var_imp %>% as.data.frame() %>% 
  transmute(vars = rownames(.), importance = .[,1]) %>% 
  ggplot(aes(x = vars, y = importance, fill = vars)) +
  geom_col() +
  scale_fill_viridis_d() +
  theme_minimal() 
```

