---
title: "P8_Nonlinear Regression_Kyuri"
author: "Kyuri Park"
date: "`r format(Sys.time(), '%B %d, %Y')`"
urlcolor: blue
output: 
  html_document:
    toc: true
    toc_depth: 1
    toc_float: true
    theme: paper
    highlight: tango
    df_print: paged
---
<style type="text/css">
@import url('https://fonts.googleapis.com/css2?family=Lato:wght@300;400&display=swap');

body{ /* Normal  */
  font-size: 13px;
  font-family: 'Lato', sans-serif;
  }
h1.title {
  font-size: 25px;
  color: DarkBlue;
  margin-bottom:5px;
}

h1 { /* Header 1 */
  font-size: 20px;
  font-weight: bold;
}
h2 { /* Header 2 */
  font-size: 15px;
  line-height: 1.6;
}
h3 { /* Header 3 */
  font-size: 14px;
  line-height: 1.6;
}

pre { /* Code block - determines code spacing between lines */
  font-size: 13px;
}
</style>
<hr>

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
               warning = FALSE,
               comment = NA)

```

# Introduction

First, load the packages:

```{r packages, warning = FALSE, message = FALSE}
library(MASS)
library(splines)
library(ISLR)
library(tidyverse)
library(magrittr)
```


First, we specify a seed as usual.

```{r seed}
set.seed(123)
```


# Prediction plot

```{r}
data(Boston)
```


## 1. Create a function called `pred_plot()` that takes as input an `lm` object, which outputs the above plot but with a prediction line generated from the model object using the `predict()` method.

```{r}
pred_plot <- function(lm.obj) {
  # first create predictions for all values of lstat
  x_pred <- seq(min(Boston$lstat), max(Boston$lstat), length.out = 500)
  y_pred <- predict(lm.obj, newdata = data.frame(lstat = x_pred))
  
  # then create a ggplot object with a line based on those predictions
  Boston %>%
    ggplot(aes(x = lstat, y = medv)) +
    geom_point() +
    geom_line(data = tibble(lstat = x_pred, medv = y_pred), size = 1, col = "skyblue") +
    theme_minimal()
}
```


## 2. Create a linear regression object called `lin_mod` which models `medv` as a function of `lstat`. Check if your prediction plot works by running `pred_plot(lin_mod)`. Do you see anything out of the ordinary with the predictions?

```{r}
lin_mod <- lm(medv ~ lstat, data = Boston)
pred_plot(lin_mod)
```

# Polynomial regression

## 3. Create another linear model `pn3_mod`, where you add the second and third-degree polynomial terms `I(lstat^2)` and `I(lstat^3)` to the formula. Create a `pred_plot()` with this model.

```{r}
pn3_mod <- lm(medv ~ lstat + I(lstat^2) + I(lstat^3), data = Boston)
pred_plot(pn3_mod)

```



## 4. Play around with the poly() function. What output does it generate with the arguments `degree = 3` and `raw = TRUE`?
```{r}
poly(1:10, degree = 3, raw = TRUE)
poly(1:10, degree = 3, raw = FALSE)
?poly
```

## 5. Use the poly() function directly in the model formula to create a 3rd-degree polynomial regression predicting `medv` using `lstat`. Compare the prediction plot to the previous prediction plot you made. What happens if you change the poly() function to `raw = FALSE`?

```{r}
# raw = TRUE
pn3_mod2_true <- lm(medv ~ poly(lstat, 3, raw = TRUE), data = Boston)
pred_plot(pn3_mod2_true)

# raw = FALSE
pn3_mod2_false <- lm(medv ~ poly(lstat, 3, raw = FALSE), data = Boston)
pred_plot(pn3_mod2_false)
```

# Piecewise regression


## 6. Create a model called `pw2_mod` with one predictor: `I(lstat <= median(lstat))`. Create a pred_plot with this model. Use the coefficients in `coef(pw2_mod)` to find out what the predicted value for a low-lstat neighbourhood is.

```{r}
pw2_mod <- lm(medv ~ I(lstat <= median(lstat)), data = Boston)
pred_plot(pw2_mod)

coef(pw2_mod)
```

## 7. Use the `cut()` function in the formula to generate a piecewise regression model called `pw5_mod` that contains 5 equally spaced sections. Again, plot the result using `pred_plot`.

```{r}
pw5_mod <- lm(medv ~ cut(lstat, 5), data = Boston)
pred_plot(pw5_mod)
```


## 8. Optional: Create a piecewise regression model `pwq_mod` where the sections are not equally spaced, but have equal amounts of training data. Hint: use the `quantile()` function.

```{r}
brks <- c(-Inf, quantile(Boston$lstat, probs = c(.2, .4, .6, .8)), Inf)
pwq_mod <- lm(medv ~ cut(lstat, brks), data = Boston)
pred_plot(pwq_mod)
```

# Piecewise polynomial regression




9. This function does not have comments. Copy - paste the function and add comments to each line. To figure out what each line does, you can first create "fake" `vec` and `knots` variables, for example `vec <- 1:20` and `knots <- 2` and try out the lines separately.

```{r}
piecewise_cubic_basis <- function(vec, knots = 1) {
  # if knots is 0 then just return the 3rd order polynomial
  if (knots == 0) return(poly(vec, degree = 3, raw = TRUE))
  # cut the vector in the number of knots + 1
  cut_vec <- cut(vec, breaks = knots + 1)
  # matrix storage 
  out <- matrix(nrow = length(vec), ncol = 0)
  # loop over the levels of the cut vector
  for (lvl in levels(cut_vec)) {
    # temporary storage
    tmp <- vec
    # set all values to zero except for the current looping value
    tmp[cut_vec != lvl] <- 0
    # append the polynomial based on this vector to the matrix
    out <- cbind(out, poly(tmp, degree = 3, raw = TRUE))
  }
  
  out
}
```


## 10. Create piecewise cubic models with 1, 2, and 3 knots (`pc1_mod` - `pc3_mod`) using this piecewise cubic basis function. Compare them using the `pred_plot()` function.

```{r}
pc1_mod <- lm(medv ~ piecewise_cubic_basis(lstat, 1), data = Boston)
pc2_mod <- lm(medv ~ piecewise_cubic_basis(lstat, 2), data = Boston)
pc3_mod <- lm(medv ~ piecewise_cubic_basis(lstat, 3), data = Boston)

pred_plot(pc1_mod)
pred_plot(pc1_mod)
pred_plot(pc1_mod)
```


# Splines

## 11. Create a data frame called `boston_tpb` with the columns `medv` and `lstat` from the `Boston` dataset.
```{r}
boston_tpb <- Boston %>% select(medv, lstat)
```

## 12. Now use `mutate` to add squared and cubed versions of the `lstat` variable to this dataset.
```{r}
boston_tpb %<>% 
  mutate(lstat2 = (lstat)^2, lstat3 = (lstat)^3)
```


## 13. Use `mutate` to add a column `lstat_tpb` to this dataset which is 0 below the median and has value `(lstat - median(lstat))^3` above the median. Tip: you may want to use `ifelse()` within your `mutate()` call.

```{r}
boston_tpb %<>%
  mutate(lstat_tpb = ifelse(lstat > median(lstat), (lstat - median(lstat))^3, 0))
```

## 14. Create a linear model `tpb_mod` using the `lm()` function. How many predictors are in the model? How many degrees of freedom does this model have?

```{r}
tpb_mod <- lm(medv ~ ., data = boston_tpb)
summary(tpb_mod)
nrow(boston_tpb)
```


## 15. Create a cubic spline model `bs1_mod` with a knot at the median using the `bs()` function. Compare its predictions to those of the `tpb_mod` using the `predict()` function on both models.

```{r}
bs1_mod <- lm(medv ~ bs(lstat, knots = median(lstat)), data =Boston)

# Comparing the predictions from the two models: negligible absolute difference
mean(abs(predict(bs1_mod) - predict(tpb_mod)))
```


## 16. Create a prediction plot from the `bs1_mod` object using the `plot_pred()` function.
```{r}
pred_plot(bs1_mod)
```

## 17. Create a natural cubic spline model (`ns3_mod`) with 3 degrees of freedom using the `ns()` function. Plot it, and compare it to the `bs1_mod`.
```{r}
ns3_mod <- lm(medv ~ ns(lstat, df = 3), data = Boston)
pred_plot(ns3_mod)
```



## 18. Plot `lin_mod`, `pn3_mod`, `pw5_mod`, `pc3_mod`, `bs1_mod`, and `ns3_mod` and give them nice titles by adding `+ ggtitle("My title")` to the plot. You may use the function `plot_grid()` from the package `cowplot` to put your plots in a grid.
```{r}
library(cowplot)

plot_grid(
  pred_plot(lin_mod) + ggtitle("Linear regression"),
  pred_plot(pn3_mod) + ggtitle("Polynomial"),
  pred_plot(pw5_mod) + ggtitle("Piecewise constant"),
  pred_plot(pc3_mod) + ggtitle("Piecewise cubic"),
  pred_plot(bs1_mod) + ggtitle("Cubic spline"),
  pred_plot(ns3_mod) + ggtitle("Natural spline")
)
```


