---
title: "Practical 1"
author: "Emilia Löscher"
date: '2022-09-14'
output: html_document
---

Loading the needed packages
```{r, message=FALSE, warning= FALSE}
library(ISLR)
library(tidyverse)
library(haven)
library(readxl)
```

Objects
```{r}
object_1 <- 1:5
object_2 <- 1L:5L
object_3 <- "-123.456"
object_4 <- as.numeric(object_2)
object_5 <- letters[object_1]
object_6 <- as.factor(rep(object_5, 2))
object_7 <- c(1, 2, 3, "4", "5", "6")
```


# 1
## Run the following code in R and inspect their data types using the class() function. Try to guess beforehand what their types will be!

```{r}
#object 1 - guess integer
class(object_1)
#correct: integer

#object 2 - guess integer
class(object_2)
#correct: integer

#object 3 - guess character
class(object_3)
#correct: character

#object 4 - guess numeric
class(object_4)
#correct: numeric

#object 5 - guess character
class(object_5)
#correct: character

#object 6 - guess factor
class(object_6)
# correct: factor 

#object 7 - guess character
class(object_7)
# correct: character
```

# 2
## Convert object_7 back to a vector of numbers using the as.numeric() function
```{r}
object_7 <- as.numeric(object_7)
class(object_7)
```
Object 7 is numeric now.

# 3 
## Make a list called objects containing object 1 to 7 using the list() function.
```{r}
obj_list <- list(object_1, object_2, object_3, object_4, object_5, object_6, object_7)
```

# 4
## Make a data frame out of object_1, object_2, and object_5 using the data.frame() function
```{r}
obj_df <- data.frame(object_1,object_2, object_5)
```


# 5
## Useful functions for determining the size of a data frame are ncol() and nrow(). Try them out.
```{r}
nrow(obj_df)
#The data frame has 5 rows

ncol(obj_df)
#The data frame has 3 colums 
```

# 6
## Use the function read_csv() to import the file “data/googleplaystore.csv” and store it in a variable called apps.
```{r}
apps <- read_csv("googleplaystore.csv")

apply(apps,2,class)
```


# 7
## Did any column get a variable type you did not expect?

I expected the "Rating" and "Reviews" to be numeric and "Installs" and "Type" to be factor.


# 8
## Use the function head() to look at the first few rows of the apps dataset
```{r}
head(apps)
```

# 9 
## Repeat steps 5, 6, and 7 but now for “data/students.xlsx” (NB: You’ll need a function from the package readxl). Also try out the function tail() and View() (with a capital V).

```{r}
students <- read_xlsx("students.xlsx")

nrow(students)
#The data frame has 37 rows 

ncol(students)
#The data frame has 3 columns

apply(students,2,class)

#I expected "grade "to be numeric and "programme" to be a factor.

head(students)

tail(students)

View(students)
```

# 10
## Create a summary of the three columns in the students dataset using the summary() function. What is the range of the grades achieved by the students?

The range can be seen, either from the Min and Max in the summary:
```{r}
summary(students)
```
Or by applying the range function to the grade column:
```{r}
range(students$grade)
```
The range in numbers (difference between minimum and maximum) can be found by calculating:
```{r}
diff(range(students$grade))
```

Note that mean, median, and quantiles for the student_numbers does not really make sense in this context. 


# 11
## Look at the help pages for filter() (especially the examples) and show the students with a grade lower than 5.5

```{r}
students %>% filter(grade < 5.5)
```

Without pipelines
```{r}
filter(students, grade < 5.5)
```

# 12
## Show only the students with a grade higher than 8 from programme A
```{r}
students %>% filter(grade > 8, programme== "A")

#Without pipelines
filter(students, grade > 8, programme== "A")
```

# 13
## Sort the students dataset such that the students from programme A are on top of the data frame and within the programmes the highest grades come first.
```{r}
arrange(students, programme, desc(grade))

#Alternatively:
arrange(students, programme, -grade)
```

# 14
## Show only the student_number and programme columns from the students dataset
```{r}
select(students, student_number, programme)
```

# 15
## Adding a column regarding pass or fail
```{r}
students <- mutate(students, pass = grade > 5.5)
students
```

Use mutate() and recode() to change the codes in the programme column of the students dataset to their names. Store the result in a variable called students_recoded

Data set with new column "students_recoded"
```{r}
mutate(students, students_recoded = recode(students$programme,"A" ="Science", "B" = "Social Science"))

#Only variable
students_recoded <- recode(students$programme,"A" ="Science", "B" = "Social Science")

```
# 16
## Create a data processing pipeline that 

```{r, warning=FALSE, message=FALSE}
##(a) loads the apps dataset,
popular_apps <- 
  read_csv("C:/Users/emsul/OneDrive/Documents/UU/INFOMDATA1/SLV_practical1/data/googleplaystore.csv") %>% 
##(b) parses the number of installs as ‘Downloads’ variable using mutate and parse_number(), 
  mutate(Downloads = parse_number(Installs)) %>% 
##(c) shows only apps with more than 500 000 000 downloads, 
  filter(Downloads > 500000000) %>% 
##(d) orders them by rating (best on top), 
  arrange(-Rating) %>% 
##(e) shows only the relevant columns (you can choose which are relevant, but select at least the Rating and Category variables). Save the result under the name popular_apps.
  select(App, Category, Rating, Downloads, Genres) %>% 
  distinct(App, .keep_all = TRUE)

popular_apps
```

# 17
## Show the median, minimum, and maximum for the popular apps dataset you made in the previous assignment.
```{r}
popular_apps %>%  
  summarize(  mean = mean(Rating), 
              variance = var(Rating), 
              min = min(Rating), 
              max = max(Rating))
```
For the other variables such a summary does not make sense as they are not numeric or all the same (Downloads = 500 000 000 (+))


# 18
Median Absolute Deviation
```{r}
mad <- function(x) {
  median(abs(x - median(x)))
}

```
## Add the median absolute deviation to the summaries you made before

```{r}
popular_apps %>%  
  summarize(  mean = mean(Rating), 
              variance = var(Rating), 
              min = min(Rating), 
              max = max(Rating),
              mad = mad(Rating))
```


# 19
## Create a grouped summary of the ratings per category in the popular apps dataset.

```{r}
popular_apps %>%  
  group_by(Category) %>% 
  summarize(  n = length(Rating), 
              mean = mean(Rating), 
              variance = var(Rating, na.rm = TRUE), 
              min = min(Rating), 
              max = max(Rating),
              mad = mad(Rating))
nrow(popular_apps)
```

# 20
## Create an interesting summary based on the Google play store apps dataset. An example could be “do games get higher ratings than communication apps?”

Out of the 20 apps that have more than 500 million downloads, 6 belong to the category "Communication", 3 to "Social" and 2 each to "Travel" and "Video Players".
There are 16 other categories which each only have 1 app that has more than 500 million downloads.

This underscores that a limited number of communication apps are more widely used than any many specific gaming apps or apps of other categories. 

The "Photography" app and the "Game" app have the highest rating with 4.5.
The "Books" app and the "News and Magazines" app have the lowest rating with 3.9.
The six communication apps have an average rating of 4.18.

