---
title: "P3_Kyuri"
author: "Kyuri Park"
date: "`r format(Sys.time(), '%B %d, %Y')`"
#mainfont: Courier New
#fontsize: 11pt
urlcolor: blue
#geometry: margin = 1 in
output: 
  # pdf_document:
  #   latex_engine: xelatex
  html_document:
    toc: true
    toc_depth: 1
    toc_float: true
    theme: paper
    highlight: tango
    df_print: paged
---

<style type="text/css">
@import url('https://fonts.googleapis.com/css2?family=Lato:wght@300;400&display=swap');

body{ /* Normal  */
  font-size: 13px;
  font-family: 'Lato', sans-serif;
  }
h1.title {
  font-size: 25px;
  color: DarkBlue;
  margin-bottom:5px;
}

h1 { /* Header 1 */
  font-size: 20px;
  font-weight: bold;
}
h2 { /* Header 2 */
  font-size: 15px;
  line-height: 1.6;
}
h3 { /* Header 3 */
  font-size: 14px;
  line-height: 1.6;
}

pre { /* Code block - determines code spacing between lines */
  font-size: 13px;
}

</style>
<hr>
```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
               warning = FALSE,
               comment = NA,
               out.width = "70%")
```

```{r load_packages}
library(tidyverse)
library(magrittr)
library(mice)
library(DAAG)
```

## 1. Create a histogram of the variable `age` using the function `geom_histogram()`.

```{r}
boys %>%  ggplot(aes(x=age)) +
  geom_histogram() 
```


## 2. Create a bar chart of the variable `gen` using the function `geom_bar()`.

```{r}
boys %>% ggplot(aes(x=gen)) +
  geom_bar()
```


## 3. Create a missingness indicator for the variables `gen`, `phb` and `tv`.

```{r}
boys <- boys %>% 
  mutate(gen_ind = is.na(gen),
         phb_ind = is.na(phb),
         tv_ind = is.na(tv))
```


## 4. Assess whether missingness in the variables `gen`, `phb` and `tv` is related to someones age.
As seen below, it is observed that the average age is largely different for those who have missing values on `gen`, `phb` and `tv`.

```{r}
vars <- c("gen_ind", "phb_ind", "tv_ind")

summaries <- list()
for (i in vars){
  summaries[[i]] <- boys %>%
    group_by(value = .data[[i]]) %>%
    summarize(avg.age = mean(age))}

pander::pander(summaries)
```


## 5. Create a histogram for the variable `age`, faceted by whether or not someone has a missing value on `gen`.
```{r}
boys %>% ggplot(aes(x=age)) +
  geom_histogram()+
  facet_wrap(~gen_ind)
```

## 6. Create a scatterplot with `age` on the x-axis and `bmi` on the y-axis, using the function `geom_point()`.
```{r}
boys %>% ggplot(aes(x=age, y=bmi)) +
  geom_point()
```


## 7. Add a colour aesthetic to the previous plot using the missingness indicator of the variable `gen`.
```{r}
boys %>% ggplot(aes(x=age, y=bmi, color=gen_ind)) +
  geom_point()
```


## 8. Visualize the relationship between `reg` (region) and `age` using a boxplot.
```{r}
boys %>% ggplot(aes(x = reg, y = age)) +
  geom_boxplot() 
```


## 9. Create a density plot of age, splitting the densities by `gen` using the `fill` aesthetic.
```{r}
boys %>% ggplot(aes(x=age, fill = gen)) +
  geom_density(alpha= 0.5) +
  scale_fill_brewer(palette = "Pastel1")
```


## 10. Create a diverging bar chart for `hgt` in the `boys` data set, that displays for every `age` year that year’s mean height in deviations from the overall average `hgt`.
```{r}
boys %>% 
  mutate(age_cat = as.factor(round(age)),
         hgt_cent = scale(hgt, scale=FALSE)) %>%
  filter(!is.na(hgt_cent)) %>% 
  group_by(age_cat) %>% 
  summarize(hgt_cent = mean(hgt_cent)) %>% 
  ggplot(aes(x = age_cat, y = hgt_cent)) +
  geom_col()

# what does this plot represent? And how the summarzie works here? Felt like mutate() should've been used
boys %>% 
  mutate(age_cat = as.factor(round(age)),
         hgt_cent = scale(hgt, scale=FALSE)) %>%
  filter(!is.na(hgt_cent)) %>% 
  ## these two lines...
  # group_by(age_cat) %>% 
  # summarize(hgt_cent = mean(hgt_cent)) %>% 
  ggplot(aes(x = age_cat, y = hgt_cent)) +
  geom_col()

```


## 11. Load the data `elastic1` and `elastic2` and bind the data frames together using the function `bind_rows()` and add a grouping variable indicating whether an observation comes from `elastic1` or from `elastic2`.
```{r}
elastic <- bind_rows(list("elastic1" = elastic1, "elastic2" = elastic2), .id = "group") %>% remove_rownames()
elastic

```


## 12. Create a scatterplot mapping `stretch` on the x-axis and `distance` on the y-axis, and map the just created group indicator as the color aesthetic.
```{r}
elastic %>% 
  ggplot(aes(x = stretch, y = distance, color = group)) +
  geom_point()
```


## 13. Recreate the previous plot, but now assess whether the results of the two data sets appear consistent by adding a linear regression line.
```{r}
elastic %>% 
  ggplot(aes(x = stretch, y = distance, color = group)) +
  geom_point() +
  geom_smooth(method = "lm")
```


## 14. For each of the data sets `elastic1` and `elastic2`, fit a regression model with `y = distance` on `x = stretch` using `lm(y ~ x, data)`.
```{r}
elastic %>% 
  split(.$group) %>%
  map(~ lm(distance ~ stretch, data = .x)) %>%
  map(summary) 
```

## 15. For both of the previously created fitted models, determine the fitted values and the standard errors of the fitted values, and the proportion explained variance $R^2$.
```{r}
res <- elastic %>%
  split(.$group) %>%
  map(~ lm(distance ~ stretch, data = .x)) %>%
  map(~ list(fitted_values = predict(., se.fit=T)$fit, 
             se_fitted = predict(., se.fit=T)$se.fit, 
             R2 = summary(.)$r.squared)) 

pander::pander(res)
```

## 16. Study the residual versus leverage plots for both models.
```{r, results='hide', fig.keep='all'}
elastic %>%
  split(.$group) %>%
  map(~ lm(distance ~ stretch, data = .x)) %>%
  map(~plot(., which=5))
```

## 17. Use the `elastic2` variable `stretch` to obtain predictions on the model fitted on `elastic1`.
```{r}
fit <- lm(distance ~ stretch, data = elastic1)
pred <- predict(fit, newdata = elastic2)
```


## 18. Now make a scatterplot to investigate similarity between the predicted values and the observed values for `elastic2`.
```{r}
plot(pred, elastic2$distance)
```

