---
title: "Timo_van_Veghel_Practical_4"
author: "Timo"
date: "2022-10-05"
output: html_document
---

```{r}
library(ISLR)
library(MASS)
library(tidyverse)
```

##Question 1:

```{r}
lm_ses <- lm(medv ~ lstat, Boston)
lm_ses
```

##Question 2:

```{r}
coef(lm_ses)
```
##Question 3:

```{r}
summary(lm_ses)
```

##Question 4:

```{r}
y_pred <- predict(lm_ses)
```

##Question 5:

```{r}
tibble(pred = y_pred, obs = Boston$medv) %>% 
  ggplot(aes(x = pred, y = obs)) +
  geom_point() +
  theme_minimal()
```

##Question 6:

```{r}
pred_dat <-
  tibble(lstat = seq(from = 0, to = 40, length.out = 1000))
```

##Question 7:

```{r}
y_pred_new <- predict(lm_ses, newdata = pred_dat)

y_pred_new

```

##Question 8:

```{r}
p_scatter <-
  Boston %>% 
  ggplot(aes(x = lstat, y = medv)) +
  geom_point() +
  theme_minimal()

p_scatter
```

##Question 9:

```{r}
pred_dat <- pred_dat %>% mutate(medv = y_pred_new)
```

##Question 10:

```{r}
p_scatter + geom_line(data = pred_dat)
```

##Question 11:

```{r}
y_pred_95 <- predict(lm_ses, newdata = pred_dat, interval = "confidence")

```

##Question 12:

```{r}
gg_pred <- tibble(
  lstat = pred_dat$lstat,
  medv  = y_pred_95[, 1],
  lower = y_pred_95[, 2],
  upper = y_pred_95[, 3]
)
```

##Question 13:

```{r}
p_scatter <-
  Boston %>% 
  ggplot(aes(x = lstat, y = medv)) +
  geom_point(colour = "dark gray") +
  geom_line(data = pred_dat, colour = "maroon") +
  geom_ribbon(data = gg_pred, aes(x = lstat, ymin = lower, ymax = upper), colour = "red", fill = "orange", alpha = 0.5) +
  theme_minimal()

p_scatter
```

##Question 14:

The ribbon represents which datapoints fall within the confidence interval of our prediction line

##Question 15:

```{r}
y_pred_95 <- predict(lm_ses, newdata = pred_dat, interval = "prediction")

gg_pred <- tibble(
  lstat = pred_dat$lstat,
  medv = y_pred_95[, 1],
  l95 = y_pred_95[, 2],
  u95 = y_pred_95[, 3]
)

Boston %>% 
  ggplot(aes(x = lstat, y = medv)) +
  geom_ribbon(aes(ymin = l95, ymax =u95), data = gg_pred, fill = "#00008b44") +
  geom_point(colour = "#883321") +
  geom_line(data = pred_dat, colour = "#00008b", size = 1) +
  theme_minimal() +
  labs(x = "Proportion of low SES households", y = "Median house value", title = "Boston house prices")
```

##Question 16:

```{r}
mse <- function(y_true, y_pred) {
  mean((y_true - y_pred)^2)
}
```

##Question 17:

```{r}
mse(1:10, 10:1)
```

##Question 18:

```{r}
mse(Boston$medv, predict(lm_ses))
```

##Question 19:

```{r}
splits <- c(rep("train", 253), rep("validation", 152), rep("tes", 101))
```

##Question 20:

```{r}
boston_master <- Boston %>% mutate(splits = sample(splits))
```

##Question 21:

```{r}
boston_train <- boston_master %>% filter(splits == "train")

boston_valid <- boston_master %>% filter(splits == "validation")

boston_test <- boston_master %>% filter(splits == "test")
```

##Question 22:

```{r}
model_1 <- lm(medv ~ lstat, data = boston_train)

summary(model_1)
```

##Question 23:

```{r}
model_1_mse_train <- mse(boston_train$medv, predict(model_1))
```

##Question 24:

```{r}
model_1_mse_valid <- mse(boston_valid$medv, predict(model_1, newdata = boston_valid))
```

##Question 25:

```{r}
model_2 <- lm(medv ~ lstat + age + tax, data = boston_train)
model_2_mse_train <- mse(boston_train$medv, predict(model_2))
model_2_mse_valid <- mse(boston_valid$medv, predict(model_2, newdata = boston_valid))
```

##Question 26:

Probably model 2

##Question 27:

```{r}
model_2_mse_test <- mse(boston_test$medv, predict(model_2, newdata = boston_test))

sqrt(model_2_mse_test)
```

##Question 28:

```{r}
mse <- function(y_true, y_pred) mean((y_true - y_pred)^2)

cv_lm <- function(formula, dataset, k) {
  stopifnot(is_formula(formula))
  stopifnot(is.data.frame(dataset))
  stopifnot(is.integer(as.integer(k)))
  
  n_samples  <- nrow(dataset)
  select_vec <- rep(1:k, length.out = n_samples)
  data_split <- dataset %>% mutate(folds = sample(select_vec))
  
  mses <- rep(0, k)
  
  for (i in 1:k) {
    # split the data in train and validation set
    data_train <- data_split %>% filter(folds != i)
    data_valid <- data_split %>% filter(folds == i)
    
    model_i <- lm(formula = formula, data = data_train)
    
    y_column_name <- as.character(formula)[2]
    
    mses[i] <- mse(y_true = data_valid[[y_column_name]],
                   y_pred = predict(model_i, newdata = data_valid))
  }
  mean(mses)
}
```

##Question 29:

```{r}
cv_lm(formula = medv ~ lstat + age + tax, dataset = Boston, k = 9)

cv_lm(formula = medv ~ lstat + I(lstat^2) + age + tax, dataset = Boston, k = 9)
```

