---
title: "Practical 7"
author: "Simranjit"
date: "2022-10-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Nonlinear Regression

##Introduction
```{r}
library(MASS)
library(splines)
library(ISLR)
library(tidyverse)
library(cowplot)
```
```{r}
set.seed(45)
```
##Prediction plot
```{r}
Boston %>%
ggplot(aes(x = lstat, y = medv)) +
geom_point() +
theme_minimal()
```
1.Create a function called pred_plot() that takes as input an lm object, which outputs the above plot but with a prediction line generated from the model object using the predict() method.
```{r}
pred_plot <- function(model) {
  x_pred <- seq(min(Boston$lstat), max(Boston$lstat), length.out = 500)
  y_pred <- predict(model, newdata = tibble(lstat = x_pred))
    Boston %>%
    ggplot(aes(x = lstat, y = medv)) +
    geom_point() +
    geom_line(data = tibble(lstat = x_pred, medv = y_pred), size = 1, col = "blue") +
    theme_minimal()
}

```

2.Create a linear regression object called lin_mod which models medv as a function of lstat. Check if your prediction plot works by running pred_plot(lin_mod). Do you see anything out of the ordinary with the predictions?
```{r}
lin_mod <- lm(medv ~ lstat, data = Boston)
pred_plot(lin_mod)
# the predicted housing value is lesser than zero for higher values.
```
##Polynomial regression
3. Create another linear model pn3_mod, where you add the second and third-degree polynomial terms I(lstat^2) and I(lstat^3) to the formula. Create a pred_plot() with this model.

```{r}
pn3_mod <- lm(medv ~ lstat + I(lstat^2) + I(lstat^3), data = Boston)
pred_plot(pn3_mod)
```
4. Play around with the poly() function. What output does it generate with the arguments degree = 3 and raw = TRUE?
```{r}
poly(1:5, degree = 3, raw = TRUE)


```
5. Use the poly() function directly in the model formula to create a 3rd-degree polynomial regression predicting medv using lstat. Compare the prediction plot to the previous prediction plot you made. What happens if you change the poly() function to raw = FALSE?
```{r}
pn3_mod2 <- lm(medv ~ poly(lstat, 3, raw = TRUE), data = Boston)
pred_plot(pn3_mod2)
summary(pn3_mod2)
```
##Piecewise regression

6. Create a model called pw2_mod with one predictor: I(lstat <= median(lstat)). Create a pred_plot with this model. Use the coefficients in coef(pw2_mod) to find out what the predicted value for a low-lstat neighbourhood is.
```{r}
pw2_mod <- lm(medv ~ I(lstat <= median(lstat)), data = Boston)
pred_plot(pw2_mod)
coef(pw2_mod)
```
 7.Use the cut() function in the formula to generate a piecewise regression model called pw5_mod that contains 5 equally spaced sections. Again, plot the result using pred_plot.
```{r}
pw5_mod <- lm(medv ~ cut(lstat, 5), data = Boston)
pred_plot(pw5_mod)
table(cut(Boston$lstat, 5))
```
8.Create a piecewise regression model pwq_mod where the sections are not equally spaced, but have equal amounts of training data. Hint: use the quantile() function.
```{r}
brks <- c(-Inf, quantile(Boston$lstat, probs = c(.2, .4, .6, .8)), Inf)
pwq_mod <- lm(medv ~ cut(lstat, brks), data = Boston)
pred_plot(pwq_mod)

table(cut(Boston$lstat, brks))

```
##Piecewise polynomial regression
9. This function does not have comments. Copy - paste the function and add comments to each line. To figure out what each line does, you can first create “fake” vec and knots variables, for example vec <- 1:20 and knots <- 2 and try out the lines separately.
```{r}
piecewise_cubic_basis <- function(vec, knots = 1) {
  
  if (knots == 0) return(poly(vec, degree = 3, raw = TRUE))
  
  cut_vec <- cut(vec, breaks = knots + 1)
  
  out <- matrix(nrow = length(vec), ncol = 0)
  
 
  for (lvl in levels(cut_vec)) {
    
   
    tmp <- vec
    
   
    tmp[cut_vec != lvl] <- 0
    
    
    out <- cbind(out, poly(tmp, degree = 3, raw = TRUE))
    
  }
  
  
  out
}



```
10. Create piecewise cubic models with 1, 2, and 3 knots (pc1_mod - pc3_mod) using this piecewise cubic basis function. Compare them using the pred_plot() function.
```{r}
pc1_mod <- lm(medv ~ piecewise_cubic_basis(lstat, 1), data = Boston)
pc2_mod <- lm(medv ~ piecewise_cubic_basis(lstat, 2), data = Boston)
pc3_mod <- lm(medv ~ piecewise_cubic_basis(lstat, 3), data = Boston)

pred_plot(pc1_mod)

pred_plot(pc2_mod)
pred_plot(pc3_mod)
```
##Splines
11.Create a data frame called boston_tpb with the columns medv and lstat from the Boston dataset.
```{r}
boston_tpb <- Boston %>% as_tibble %>% select(medv, lstat)
```
12.Now use mutate to add squared and cubed versions of the lstat variable to this dataset.
```{r}
boston_tpb <- boston_tpb %>% mutate(lstat2 = lstat^2, lstat3 = lstat^3)
```
13.Use mutate to add a column lstat_tpb to this dataset which is 0 below the median and has value (lstat - median(lstat))^3 above the median. Tip: you may want to use ifelse() within your mutate() call.
```{r}
boston_tpb <- boston_tpb %>% 
  mutate(lstat_tpb = ifelse(lstat >  median(lstat), (lstat - median(lstat))^3, 0))
```
14.Create a linear model tpb_mod using the lm() function. How many predictors are in the model? How many degrees of freedom does this model have?
```{r}
tpb_mod <- lm(medv ~ lstat + lstat2 + lstat3 + lstat_tpb, data = boston_tpb)
summary(tpb_mod)
```
15. Create a cubic spline model bs1_mod with a knot at the median using the bs() function. Compare its predictions to those of the tpb_mod using the predict() function on both models.
```{r}
bs1_mod <- lm(medv ~ bs(lstat, knots = median(lstat)), data = Boston)
summary(bs1_mod)
mean(abs(predict(bs1_mod) - predict(tpb_mod)))
```
16.Create a prediction plot from the bs1_mod object using the plot_pred() function.
```{r}
pred_plot(bs1_mod)
```
17.Create a natural cubic spline model (ns3_mod) with 3 degrees of freedom using the ns() function. Plot it, and compare it to the bs1_mod.
```{r}
ns3_mod <- lm(medv ~ ns(lstat, df = 3), data = Boston)
pred_plot(ns3_mod)
```
18.Plot lin_mod, pn3_mod, pw5_mod, pc3_mod, bs1_mod, and ns3_mod and give them nice titles by adding + ggtitle("My title") to the plot. You may use the function plot_grid() from the package cowplot to put your plots in a grid.
```{r}
plot_grid(
  pred_plot(lin_mod) + ggtitle("Linear regression"),
  pred_plot(pn3_mod) + ggtitle("Polynomial"),
  pred_plot(pw5_mod) + ggtitle("Piecewise constant"),
  pred_plot(pc3_mod) + ggtitle("Piecewise cubic"),
  pred_plot(bs1_mod) + ggtitle("Cubic spline"),
  pred_plot(ns3_mod) + ggtitle("Natural spline")
)  
```
