---
title: "Practical 3"
author: "Florian van Leeuwen"
date: "9/26/2022"
output: html_document
---

## Practical 3

```{r, packages}
library(tidyverse)
library(magrittr)
library(mice)
library(DAAG)
library(stargazer)
```

```{r}
# inspect data
head(boys)
tail(boys)
!is.unsorted(boys$age)
summary(boys)
```
## 1. Create a histogram of the variable age using the function geom_histogram().
```{r}
ggplot(boys, aes(x = age)) +
  geom_histogram(fill = "#fee6ce") +
  theme_classic()
```

## 2. Create a bar chart of the variable gen using the function geom_bar().
```{r}
ggplot(boys, aes(x = gen)) + 
  geom_bar(fill = "#addd8e") +
  theme_classic()
```

## 3. Create a missingness indicator for the variables gen, phb and tv.
```{r}
md.pattern(boys)

boys <- boys %>% 
  mutate(gen_mis = is.na(gen),
         phb_mis = is.na(phb),
         tv_mis = is.na(tv))
```

## 4. Assess whether missingness in the variables gen, phb and tv is related to someones age.
```{r}
boys %>% 
  group_by(gen_mis) %>% 
  summarise(age = mean(age))

boys %>% 
  group_by(phb_mis) %>% 
  summarise(age = mean(age))

boys %>% 
  group_by(tv_mis) %>% 
  summarise(age = mean(age))
```

## 5. Create a histogram for the variable age, faceted by whether or not someone has a missing value on gen.
```{r}
ggplot(boys, aes(x = age)) +
  geom_histogram(fill = "darkblue") +
  facet_grid(cols = vars(gen_mis)) +
  labs(title = "Distribution of age by if someone has a missing value on gen") +
  theme_minimal()

ggplot(boys, aes(x = age)) +
  geom_histogram(fill = "darkblue") +
  facet_wrap(~gen_mis) +
  labs(title = "Distribution of age by if someone has a missing value on gen") +
  theme_minimal()
```

## 6. Create a scatterplot with age on the x-axis and bmi on the y-axis, using the function geom_point().
```{r}
boys %>% 
  ggplot(aes(x = age, y = bmi)) +
  geom_point(color = "darkblue") +
  theme_minimal()
```

## 7. Add a colour aesthetic to the previous plot using the missingness indicator of the variable gen.
```{r}
boys %>% 
  ggplot(aes(x = age, y = bmi, color = gen_mis, shape = gen_mis)) +
  geom_point() +
  theme_classic()
```

## 8. Visualize the relationship between reg (region) and age using a boxplot.
```{r}
boys %>% 
  ggplot(aes(x = age)) +
  geom_boxplot(color = "darkblue") +
  facet_grid(rows = vars(reg)) +
  theme_minimal()
```
## 9. Create a density plot of age, splitting the densities by gen using the fill aesthetic.
```{r}
boys %>% 
  ggplot(aes(x = age, fill = gen)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  labs(title = "Distribution of age by gen")
```
## 10. Create a diverging bar chart for hgt in the boys data set, that displays for every age year that year’s mean height in deviations from the overall average hgt.
```{r}
boys %>% 
  mutate(Age = as.factor(ceiling(age)),
         hgt_dif =hgt - mean(hgt, na.rm = TRUE)) %>%
  group_by(Age) %>% 
  summarize(Height_diff = mean(hgt_dif, na.rm = TRUE)) %>% 
  mutate(color = ifelse(Height_diff > 0, "Above average", "Below average")) %>%
  ggplot(aes(x = Height_diff, y = Age, fill = color)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
  theme(legend.title = element_blank()) +
  labs(x = "Height difference from the mean")
  

```

## 11. Load the data elastic1 and elastic2 and bind the data frames together using the function bind_rows() and add a grouping variable indicating whether an observation comes from elastic1 or from elastic2
```{r}
head(elastic1)
head(elastic2)
elastic <- bind_rows("Elastic1" = elastic1,
                     "Elastic2" = elastic2,
                     .id = "Set")
```
## 12. Create a scatterplot mapping stretch on the x-axis and distance on the y-axis, and map the just created group indicator as the color aesthetic.
```{r}
elastic %>%
  ggplot(aes(x = stretch, y = distance, color = Set, shape = Set)) +
  geom_point() +
  theme_classic()
```
## 13. Recreate the previous plot, but now assess whether the results of the two data sets appear consistent by adding a linear regression line.
```{r}
elastic %>%
  ggplot(aes(x = stretch, y = distance, color = Set, shape = Set)) +
  geom_point() +
  geom_smooth(method='lm', se = FALSE) +
  theme_classic() 
```

## 14. For each of the data sets elastic1 and elastic2, fit a regression model with y = distance on x = stretch using lm(y ~ x, data).
```{r}
model1 <- lm(distance ~ stretch, elastic1)
model2 <- lm(distance ~ stretch, elastic2)
```

## 15. For both of the previously created fitted models, determine the fitted values and the standard errors of the fitted values, and the proportion explained variance  
```{r}
model1 %>% predict(se.fit = TRUE)
predict(model2, se.fit = TRUE)

stargazer(model1, model2, title="Results", align=TRUE, type="text")
```

## 16. Study the residual versus leverage plots for both models.
```{r}
plot(model1, which = 5)
plot(model2, which = 5)
```

## 17. Use the elastic2 variable stretch to obtain predictions on the model fitted on elastic1.
```{r}
pred <- predict(model1, newdata = elastic2)
```

## 18. Now make a scatterplot to investigate similarity between the predicted values and the observed values for elastic2.
```{r}
pred_dat <- 
  data.frame(distance = pred, 
             stretch  = elastic2$stretch) %>%
  bind_rows(Predicted = .,
            Observed  = elastic2, 
            .id = "Predicted")

pred_dat %>% ggplot(aes(x = distance, y = stretch, color = Predicted)) +
                      geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Predicted and observed distances")
```

