---
title: "SLV Practical 4"
author: "Shannon Dickson"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
# Packages
library(ISLR)
library(MASS)
library(tidyverse)
```

```{r}
# Data
Boston <- MASS::Boston
```

**1. Create a linear model object called lm_ses using the formula medv ~ lstat and the Boston dataset.**

```{r}
lm_ses <- lm(formula = medv ~ lstat, data = Boston)
```

**2. Use the function coef() to extract the intercept and slope from the lm_ses object. Interpret the slope coefficient.**

```{r}
coef(lm_ses)
```

**3. Use summary() to get a summary of the lm_ses object. What do you see? You can use the help file ?summary.lm.**

```{r}
summary(lm_ses)
```

**4. Save the predicted y values to a variable called y_pred**

```{r}
y_pred <- predict(lm_ses)
```

**5. Create a scatter plot with y_pred mapped to the x position and the true y value (Boston$medv) mapped to the y value. What do you see? What would this plot look like if the fit were perfect?**

```{r}
# bind in a tibble
tibble(pred = y_pred, 
       obs  = Boston$medv) %>% 
# plot
  ggplot(aes(x = pred, y = obs)) +
  geom_point() +
  labs(x = "predicted values",
       y = "observed values") +
  geom_abline(slope = 1) +
    theme_bw()
```

**6. Use the seq() function to generate a sequence of 1000 equally spaced values from 0 to 40. Store this vector in a data frame with (data.frame() or tibble()) as its column name lstat. Name the data frame pred_dat.**

```{r}
pred_dat <- tibble(lstat = seq(0, 40, length.out = 1000))
```

**7. Use the newly created data frame as the newdata argument to a predict() call for lm_ses. Store it in a variable named y_pred_new.**

```{r}
y_pred_new <- predict(lm_ses, newdata = pred_dat) 
```

**8.Create a scatter plot from the Boston dataset with lstat mapped to the x position and medv mapped to the y position. Store the plot in an object called p_scatter.**

```{r}
p_scatter <- 
  Boston %>% 
  ggplot(aes(x = lstat, y = medv)) +
  geom_point() +
  theme_bw()

p_scatter
```

**9. Add the vector y_pred_new to the pred_dat data frame with the name medv.**

```{r}
pred_dat$medv <- y_pred_new
```

**10. Add a geom_line() to p_scatter, with pred_dat as the data argument. What does this line represent?**

```{r}
p_scatter + 
  geom_line(data = pred_dat)
```

**11. The interval argument can be used to generate confidence or prediction intervals. Create a new object called y_pred_95 using predict() (again with the pred_dat data) with the interval argument set to “confidence”. What is in this object?**

```{r}
y_pred_95 <- predict(lm_ses, newdata = pred_dat, interval = "confidence")

# a matrix with coefficients and their upper/lower condifence limits
head(y_pred_95)
```

**12. Create a data frame with 4 columns: medv, lstat, lower, and upper.**

```{r}
plot_data <- tibble(
  lstat = pred_dat$lstat,
  medv  = y_pred_95[, 1],
  lower = y_pred_95[, 2],
  upper = y_pred_95[, 3])

plot_data
```
**13. Add a geom_ribbon() to the plot with the data frame you just made. The ribbon geom requires three aesthetics: x (lstat, already mapped), ymin (lower), and ymax (upper). Add the ribbon below the geom_line() and the geom_points() of before to make sure those remain visible. Give it a nice colour and clean up the plot, too!**

```{r}
Boston %>% 
  ggplot(aes(lstat, medv)) + 
  geom_ribbon(aes(ymin = lower, ymax = upper), data = plot_data, fill = "#00008b44") +
  geom_point(colour = "darkseagreen4") + 
  geom_line(data = pred_dat, colour = "darkorange", size = 1) +
  labs(title = "Boston house prices",
       x = "Proportion of low SES households",
       y = "Median house value") +
  theme_minimal()
```

**14. Explain in your own words what the ribbon represents.**

The ribbon is the 95% confidence interval of the linear regression line, representing the uncertainty in our estimates (coefficients).

**15. Do the same thing, but now with the prediction interval instead of the confidence interval.**

```{r}
# get the prediction interval with predict
y_pred_int <- predict(lm_ses, newdata = pred_dat, interval = "prediction")
```

```{r}
# cbind to a tibble
plot_data <- tibble(
  lstat = pred_dat$lstat,
  medv  = y_pred_int[, 1],
  lwr   = y_pred_int[, 2],
  upr   = y_pred_int[, 3])
```

```{r}
# Create the plot
Boston %>% 
  ggplot(aes(lstat, medv)) + 
  geom_ribbon(aes(ymin = lwr, ymax = upr), data = plot_data, fill = "lightgrey") +
  geom_point(colour = "darkseagreen4") + 
  geom_line(data = pred_dat, colour = "darkorange", size = 1) +
  labs(title = "Boston house prices",
       x = "Proportion of low SES households",
       y = "Median house value") +
    theme_minimal() 
```

**16. Write a function called mse() that takes in two vectors: true y values and predicted y values, and which outputs the mean square error.**

```{r}
mse <- function(y_true, y_pred) {
  mean((y_true - y_pred)^2)
}
```

**17. Make sure your mse() function works correctly by running the following code.**

```{r}
mse(1:10, 10:1)
```

**18. Calculate the mean square error of the lm_ses model. Use the medv column as y_true and use the predict() method to generate y_pred.**

```{r}
mse(Boston$medv, predict(lm_ses))
```

**19. The Boston dataset has 506 observations. Use c() and rep() to create a vector with 253 times the word “train”, 152 times the word “validation”, and 101 times the word “test”. Call this vector splits.**

```{r}
splits <- c(rep("train", 253), rep("validation", 152), rep("test", 101))
```

**20. Use the function sample() to randomly order this vector and add it to the Boston dataset using mutate(). Assign the newly created dataset to a variable called boston_master.**

```{r}
boston_master <- Boston %>% mutate(splits = sample(splits))
```

**21. Now use filter() to create a training, validation, and test set from the boston_master data. Call these datasets boston_train, boston_valid, and boston_test.**

```{r}
boston_train <- boston_master %>%
  filter(splits == "train")

boston_valid <- boston_master %>%
  filter(splits == "validation")

boston_test  <- boston_master %>%
  filter(splits == "test")
```

**22. Train a linear regression model called model_1 using the training dataset. Use the formula medv ~ lstat like in the first lm() exercise. Use summary() to check that this object is as you expect.**

```{r}
model_1 <- lm(medv ~ lstat, data = boston_train)

summary(model_1)
```

**23. Calculate the MSE with this object. Save this value as model_1_mse_train.**
```{r}
model_1_mse_train <- mse(y_true = boston_train$medv, y_pred = predict(model_1))
```

**24. Now calculate the MSE on the validation set and assign it to variable model_1_mse_valid. Hint: use the newdata argument in predict().**

```{r}
model_1_mse_valid <- mse(y_true = boston_valid$medv, y_pred = predict(model_1, newdata = boston_valid))
```

**25. Create a second model model_2 for the train data which includes age and tax as predictors. Calculate the train and validation MSE.**

```{r}
model_2 <- lm(medv ~ lstat + age + tax, data = boston_train)

model_2_mse_train <- mse(y_true = boston_train$medv, y_pred = predict(model_2))

model_2_mse_valid <- mse(y_true = boston_valid$medv, y_pred = predict(model_2, newdata = boston_valid))
```

**26. Compare model 1 and model 2 in terms of their training and validation MSE. Which would you choose and why?**


**27. Calculate the test MSE for the model of your choice in the previous question. What does this number tell you?**

MSE TEST: amount of error expected when predicting the median of a new (unseen) town in Boston 

```{r}
model_2_mse_test <- mse(y_true = boston_test$medv, y_pred = predict(model_2, newdata = boston_test)) %>% 
  sqrt()
```

**28. Create a function that performs k-fold cross-validation for linear models.**

```{r}
cv_lm <- function(formula, dataset, k) {
  # We can do some error checking before starting the function
  stopifnot(is_formula(formula))       # formula must be a formula
  stopifnot(is.data.frame(dataset))    # dataset must be data frame
  stopifnot(is.integer(as.integer(k))) # k must be convertible to int
  
  # first, add a selection column to the dataset as before
  n_samples  <- nrow(dataset)
  select_vec <- rep(1:k, length.out = n_samples)
  data_split <- dataset %>% mutate(folds = sample(select_vec))
  
  # initialise an output vector of k mse values, which we 
  # will fill by using a _for loop_ going over each fold
  mses <- rep(0, k)
  
  # start the for loop
  for (i in 1:k) {
    # split the data in train and validation set
    data_train <- data_split %>% filter(folds != i)
    data_valid <- data_split %>% filter(folds == i)
    
    # calculate the model on this data
    model_i <- lm(formula = formula, data = data_train)
    
    # Extract the y column name from the formula
    y_column_name <- as.character(formula)[2]
    
    # calculate the mean square error and assign it to mses
    mses[i] <- mse(y_true = data_valid[[y_column_name]],
                   y_pred = predict(model_i, newdata = data_valid))
  }
  
  # now we have a vector of k mse values. All we need is to
  # return the mean mse!
  mean(mses)
}
```

**29. Use your function to perform 9-fold cross validation with a linear model with as its formula medv ~ lstat + age + tax. Compare it to a model with as formulat medv ~ lstat + I(lstat^2) + age + tax.**

```{r}
cv_lm(formula = medv ~ lstat + age + tax, dataset = Boston, k = 9)
```

```{r}
cv_lm(formula = medv ~ lstat + I(lstat^2) + age + tax, dataset = Boston, k = 9)
```

