---
title: "A.Carriero Practical 3"
author: "Alex Carriero"
date: "26/09/2022"
output: pdf_document
---

```{r}
# Load Libraries
library(tidyverse)
library(magrittr)
library(mice)
library(DAAG)
```

```{r}
# Use the boys data set from mice

!is.unsorted(boys$age) # check if data are sorted by age
summary(boys)          # summary statistics 
```


## 1. Create a histogram of the variable age using the function geom_histogram().
```{r}
boys %>% 
  ggplot(aes(x=age)) + 
  geom_histogram(fill = "wheat", color = "wheat3", binwidth=1, alpha = 0.7)+ 
  theme_minimal()
```

## 2. Create a bar chart of the variable gen using the function geom_bar().
```{r}
boys %>% 
  ggplot(aes(x= gen)) + 
  geom_bar(fill = "skyblue", color = "grey", alpha = 0.6) + 
  theme_minimal()
```

```{r, fig.height = 5, fig.width = 10}
# Assessing the missing data pattern 
md.pattern(boys)
```
## 3. Create a missingness indicator for the variables gen, phb and tv
```{r}
# create new indicator variables ( 1 = missing, 0 = not missing)
boys_m <- boys %>% 
            mutate(gen_m = as.factor(ifelse(is.na(gen),1,0)), 
                   phb_m = as.factor(ifelse(is.na(phb),1,0)), 
                   tv_m  = as.factor(ifelse(is.na(tv), 1,0)))
```

## 4. Assess whether missingness in the variables gen, phb and tv is related to someones age.
```{r}
# summarize age by missingness in gen 
boys_m %>% 
  group_by(gen_m) %>% 
  summarize(mean = mean(age), 
            min = min(age),
            med = median(age),
            max = max(age),
            sd  = sd(age))

# summarize age by missingness in phb
boys_m %>% 
  group_by(phb_m) %>% 
  summarize(mean = mean(age), 
            min = min(age),
            med = median(age),
            max = max(age),
            sd  = sd(age))

# summarize age by missingness in tv
boys_m %>% 
  group_by(tv_m) %>% 
  summarize(mean = mean(age), 
            min = min(age),
            med = median(age),
            max = max(age),
            sd  = sd(age))

# Results: we see that the mean/median age among people with missing values for gen, 
#          phb and tv is much younger than people without missing values. 

# much larger sd for age among people with missing values. 
```


## 5. Create a histogram for the variable age, faceted by whether or not someone has a missing value on gen.
```{r}
boys_m %>% 
    ggplot(aes(x = age))+ 
    geom_histogram( fill = "wheat", color = "wheat3", aes = 0.5) +
    theme_minimal() + 
    facet_wrap(~gen_m, labeller = labeller(gen_m = c("0" = "Not Missing", 
                                                     "1" = "Missing")))
```
## 6. Create a scatterplot with age on the x-axis and bmi on the y-axis, using the function geom_point() and 
## 7. Add a colour aesthetic to the previous plot using the missingness indicator of the variable gen.
```{r}
boys_m %>% 
    ggplot(aes(x = age, y = bmi, color = gen_m)) + 
    geom_point(alpha = 0.6) + 
    scale_color_manual(name = "Gen", labels = c('Not Missing', 'Missing'), 
                       values = c('darkolivegreen', 'darkolivegreen2')) +
    theme_minimal() 
```
## 8. Visualize the relationship between reg (region) and age using a boxplot.
```{r}
boys_m %>%
  ggplot(aes(x= reg, y = age)) + 
  geom_boxplot(fill = "aliceblue")
```

## 9. Create a density plot of age, splitting the densities by gen using the fill aesthetic.
```{r}
boys_m %>% 
  ggplot(aes(x = age, fill = gen)) + 
  geom_density(alpha = 0.7) + 
  theme_minimal() +
  scale_fill_brewer() 
```

## 10. Create a diverging bar chart for hgt in the boys data set, that displays for every age year that year’s mean height in deviations from the overall average hgt.
```{r}
boys %>%
  mutate(Age = as.factor(ceiling(age)),
         Height = hgt - mean(hgt, na.rm = TRUE)) %>%
  group_by(Age) %>%
  summarize(Height = mean(Height, na.rm = TRUE)) %>%
  mutate(color = ifelse(Height > 0, "Above average", "Below average")) %>%
  ggplot(aes(x = Height, y = Age, fill = color)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
  theme(legend.title = element_blank())
```

## 11. Load the data elastic1 and elastic2 and bind the data frames together using the function bind_rows() and add a grouping variable indicating whether an observation comes from elastic1 or from elastic2.
```{r}
elastic <- bind_rows("Elastic1" = elastic1,
                     "Elastic2" = elastic2,
                     .id = "Set")
```

## 12. Create a scatterplot mapping stretch on the x-axis and distance on the y-axis, and map the just created group indicator as the color aesthetic.
```{r}
elastic %>% 
  ggplot(aes(x= stretch, y = distance, color = Set)) + 
  geom_point() + 
  theme_minimal()
```

## 13. Recreate the previous plot, but now assess whether the results of the two data sets appear consistent by adding a linear regression line.
```{r}
elastic %>% 
  ggplot(aes(x = stretch, y = distance, col = Set)) +
  geom_point() +
  geom_smooth(method = "lm", aes(fill = Set)) + 
  theme_minimal()
```

## 14. For each of the data sets elastic1 and elastic2, fit a regression model with y = distance on x = stretch using lm(y ~ x, data).
```{r}
fit1 <- elastic1 %$% 
            lm(distance~stretch)

fit2 <- elastic2 %$% 
            lm(distance~stretch)
```

## 15. For both of the previously created fitted models, determine the fitted values and the standard errors of the fitted values, and the proportion explained variance $R^2$. 
```{r}
# predicted values 
fit1 %>% predict(se.fit = TRUE)
fit2 %>% predict(se.fit = TRUE)

# summary 
fit1 %>% summary()
fit2 %>% summary()

# directly r-squared 
fit1 %>% summary() %$% r.squared
fit2 %>% summary() %$% r.squared
```

## 16. Study the residual versus leverage plots for both models.
```{r}
fit1 %>% plot(which = 5)
fit2 %>% plot(which = 5)
```

## 17. Use the elastic2 variable stretch to obtain predictions on the model fitted on elastic1.
```{r}
pred <- predict(fit1, newdata = elastic2)
# pred
```

## 18. Now make a scatterplot to investigate similarity between the predicted values and the observed values for elastic2.
```{r}
x <- as.data.frame(cbind(elastic2,pred)) %>% 
        pivot_longer(cols = c("distance", "pred"), names_to = "predicted", values_to = "distance") %>% 
        mutate(predicted = recode(predicted, "distance" = "obs"))

x %>%
  ggplot(aes(x= stretch, y = distance, col = predicted, fill = predicted)) + 
  geom_point() + 
  geom_smooth(method = "lm")+ 
  theme_minimal()
```
```{r}
# finally determine correlation between observed and predicted values 
cor(elastic2$distance, pred)
```

