---
title: "P2 - Alex Carriero"
author: "Alex Carriero"
date: "19/09/2022"
output: pdf_document
---

```{r, warning=F, message=F}
# Libraries
library(ISLR)
library(tidyverse)
```

```{r, fig.height=5, fig.width=10}
# Intro to GGplot using Hitters dataset
homeruns_plot <- 
  ggplot(Hitters, aes(x = Hits, y = HmRun)) +
  geom_point() +
  labs(x = "Hits", y = "Home runs")

homeruns_plot
```
```{r, fig.height=5, fig.width=10} 
# layered plots 
homeruns_plot + 
  geom_density_2d() +
  labs(title = "Cool density and scatter plot of baseball data") +
  theme_minimal()
```
1. Name the aesthetics, geoms, scales, and facets of the above visualisation. Also name any statistical transformations or special coordinate systems.

- Aesthetics: x = Hits, y = Home Runs
- Geoms: 
    - geom_point() used to create scatter plot
    - geom_density_2d() used to perform a 2D kernel density estimation and display the results with contours. 
- Scales: none 
- Facets: none

2. Run the code below to generate data. There will be three vectors in your environment. Put them in a data frame for entering it in a ggplot() call using either the data.frame() or the tibble() function. Give informative names and make sure the types are correct (use the as.<type>() functions). Name the result gg_students. 

```{r} 
# Generate data
set.seed(1234)
student_grade  <- rnorm(32, 7)
student_number <- round(runif(32) * 2e6 + 5e6)
programme      <- sample(c("Science", "Social Science"), 32, replace = TRUE)

# Store in  data frame 

gg_students <- as.data.frame(cbind(student_number, student_grade, programme))
gg_students <- gg_students %>% 
                mutate(student_number = as.numeric(student_number),
                       student_grade = as.numeric(student_grade), 
                       programme = as.factor(programme))

# gg_students
```

\newpage

3. Plot the first homeruns_plot again, but map the Hits to the y-axis and the HmRun to the x-axis instead.
```{r, fig.height=4, fig.width=10}
ggplot(Hitters, aes(x= HmRun, y = Hits)) + 
  geom_point()
```

4. Recreate the same plot once more, but now also map the variable League to the colour aesthetic and the variable Salary to the size aesthetic.

```{r, results="hide"}
summary(Hitters$League)
summary(Hitters$Salary)  # there are 59 missing salary values

Hitters <- Hitters%>%
           na.omit(Hitters$Salary) 
```

```{r, fig.height=4, fig.width=10}
new_plot <- ggplot(Hitters, aes(x= HmRun, y = Hits)) + 
            geom_point(aes(color = League, size = Salary))

new_plot
```

\newpage

5. Look at the many different geoms on the reference website.\
\
There are two types of geoms:

 - geoms which perform a transformation of the data beforehand, such as geom_density_2d() which calculates contour lines from x and y positions.
- geoms which do not transform data beforehand, but use the aesthetic mapping directly, such as geom_point().


6. Use geom_histogram() to create a histogram of the grades of the students in the gg_students dataset. Play around with the binwidth argument of the geom_histogram() function.
```{r}
gg_students %>% 
  ggplot(aes(x= student_grade))+ 
  geom_histogram(binwidth=0.4)
```

7. Use geom_density() to create a density plot of the grades of the students in the gg_students dataset. Add the argument fill = "light seagreen" to geom_density().
```{r, fig.height=3, fig.width=10}
d <- gg_students %>% 
      ggplot(aes(x= student_grade))+ 
      geom_density(fill = "light seagreen", alpha= 0.5)
d
```

8. Add rug marks to the density plot through geom_rug(). You can edit the colour and size of the rug marks using those arguments within the geom_rug() function.
```{r, fig.height=3, fig.width=10}
d + 
  geom_rug(color = "black", size = 0.5)
```

9. Increase the data to ink ratio by removing the y axis label, setting the theme to theme_minimal(), and removing the border of the density polygon. Also set the limits of the x-axis to go from 0 to 10 using the xlim() function, because those are the plausible values for a student grade.
```{r, fig.height=3, fig.width=10}
gg_students %>%
  ggplot(aes(x= student_grade))+ 
  geom_density(fill = "light seagreen", alpha= 0.5, color = NA)+
  geom_rug(color = "black", size = 0.5)+
  theme_minimal()+
  theme(axis.text.y=element_blank()) + 
  xlim(0,10)
```

10. Create a boxplot of student grades per programme in the gg_students dataset you made earlier: map the programme variable to the x position and the grade to the y position. For extra visual aid, you can additionally map the programme variable to the fill aesthetic.
```{r}

```

11. What do each of the horizontal lines in the boxplot mean? What do the vertical lines (whiskers) mean?


12. Comparison of distributions across categories can also be done by adding a fill aesthetic to the density plot you made earlier. Try this out. To take care of the overlap, you might want to add some transparency in the geom_density() function using the alpha argument.

13. Create a bar plot of the variable Years from the Hitters dataset.
