---
title: "Example answers"
author: "Laura HASBINI"
date: "28-9-2018"
mainfont: Arial
fontsize: 12pt
urlcolor: blue
output: 
  pdf_document:
    latex_engine: xelatex
---

# Data Visualisation using ggplot2

First, load the packages:

```{r load_packages}
library(ISLR)
library(tidyverse)
```

## 1. What is ggplot

Example of the dataset Hitters found in the library ISLR. This dataset contains the names of several players other variables linked to their sport. 
```{r objects}
head(Hitters)
```

Below, we can the histogram of salary in thousands of dollars.
```{r}
hist(Hitters$Salary, main = "Histogram of Hitters salary", xlab = "Salary in thousands of dollars")
```

Other type of graphs can be done, such as barplot. We can see below the barplot of the number of members in each league. 

```{r}
barplot(table(Hitters$League))
```

Regular 2d plots can be made with the function plot. We can observe in the graphique below, the number of home runs plotted regarding the number of hits, each point represent one player.
```{r}
plot(x=Hitters$Hits, y=Hitters$HmRun, xlab = 'Hits', ylab = "Home runs")
```

Another wa to plot suchs graphs is to use ggplot. With this functions, plot are considered as object. The plotting can thus be decomposed in several steps : 
- The selection of the data
- The aesthetic of the plot
- Additional components
- The labels, title, ...

This is particularly helpful to add additional features to plots already created, as can be seen below. 

```{r}
homeruns_plot <- ggplot(Hitters, aes(x=Hits, y=HmRun)) + geom_point() + labs(x = "Hits", y = "Home runs")

homeruns_plot
```

```{r}
homeruns_plot + 
  geom_density_2d() +
  labs(title = "Cool density and scatter plot of baseball data") +
  theme_minimal()
```

###QUESTION 1 : 
"Name the aesthetics, geoms, scales and facets of the above visualisation. Also name any statistical transformations or special coordinate systems"

```{r}
# The above graph represent the number of home runs as a function of the number of hits. Consequently the number of hits are mapped to x-position and the number of home runs to y-position.
# The geoms used are points and contour lines
# Both x and y axis are continuously scaled
# Finally the plot doesn't have any facets, statistical transformations and the coordinate system is cartesian.
```

## Aesthetics and data preparation
Before doing any plot, one must pay attention to the type of the data. Numbers should be "numerics", categories should be "factors" and names/identifiers should be "character".

###QUESTION 2 
"Run the code below to generate data. There will be three vectors in your environment. Put them in a data frame for entering it in a ggplot() call using either the data.frame() or the tibble() function. Give informative names and make sure the types are correct (use the as.<type>() functions). Name the result gg_students"

```{r}
set.seed(1234)
student_grade  <- rnorm(32, 7)
student_number <- round(runif(32) * 2e6 + 5e6)
programme      <- sample(c("Science", "Social Science"), 32, replace = TRUE)
```

```{r}
gg_students <- tibble(
  number = as.character(student_number), # an identifier
  grade  = student_grade,                # already the correct type.
  prog   = as.factor(programme)          # categories should be factors.
)

head(gg_students)
```

###QUESTION 3
"Plot the first homeruns_plot again, but map the Hits to the y-axis and the HmRun to the x-axis instead."

```{r}
ggplot(Hitters, aes(y=Hits, x=HmRun)) + geom_point() + labs(y = "Hits", x = "Home runs")
```

###QUESTION 4 
"Recreate the same plot once more, but now also map the variable League to the colour aesthetic and the variable Salary to the size aesthetic."

```{r}
ggplot(Hitters, aes(x=HmRun, y=Hits, colour = League, size = Salary)) + geom_point() + labs(x = "Home runs", y = "Hits", colour = "League", size = "Salary")
```
Other possible aesthetics : 
- x
- y
- alpha (transparency)
- colour
- fill
- group
- shape
- size
- stroke

## Geoms

Geoms controls the mapping representation of the plot. The most used one are for example contour lines and points. 
We can identify 2 type of geoms : 
- The one that transform the data beforehand
- The one that don't transform the data but just control the mapping

###QUESTION 5
"Look at the many different geoms on the reference website."

We can cite some geos usefull for data analysis :
- Bar charts : 'geom_bar()', 'geom_col()' or 'stat_count()'
- Box and whiskers plot : 'geom_boxplot()', 'stat_boxplot()'
- Smoothed density : 'geom_count()', 'stat_sum()'
- Contour of 2D density : 'geom_density_2d()', 'geom_density_2d_filler()', 'stat_density_2d()', 'stat_density_2d_filled()'
- Quantile regressino : 'geom_quantile()', 'stat_quantile()'

## Visual exploratory data analysis

###Histogram

####QUESTION 6 
"Use geom_histogram() to create a histogram of the grades of the students in the gg_students dataset. Play around with the binwidth argument of the geom_histogram() function."

```{r}
gg_students %>%
  ggplot(aes(x = grade)) +
  geom_histogram(binwidth = .4)
```

### Density

####QUESTION 7
"Use geom_density() to create a density plot of the grades of the students in the gg_students dataset. Add the argument fill = "light seagreen" to geom_density()"

```{r}
gg_students %>%
  ggplot(aes(x = grade)) +
  geom_density(fill = "light seagreen")
```


####QUESTION 8
"Add rug marks to the density plot through geom_rug(). You can edit the colour and size of the rug marks using those arguments within the geom_rug() function."

```{r}
gg_students %>%
  ggplot(aes(x = grade)) +
  geom_density(fill = "light seagreen") +
  geom_rug(size=0.75, colour = "light seagreen")
```

####QUESTION 9
"Increase the data to ink ratio by removing the y axis label, setting the theme to theme_minimal(), and removing the border of the density polygon. Also set the limits of the x-axis to go from 0 to 10 using the xlim() function, because those are the plausible values for a student grade."

```{r}
gg_students %>% 
  ggplot(aes(x = grade)) +
  geom_density(fill = "light seagreen", colour = NA) +
  geom_rug(size = 0.75, colour = "light seagreen") +
  theme_minimal() +
  labs(y = "") +
  xlim(0, 10)
```

### Boxplot

####QUESTION 10
"Create a boxplot of student grades per programme in the gg_students dataset you made earlier: map the programme variable to the x position and the grade to the y position. For extra visual aid, you can additionally map the programme variable to the fill aesthetic."

```{r}
gg_students %>% 
  ggplot(aes(x= prog, y = grade, fill=prog)) +
  geom_boxplot() +
  theme_minimal() 
```

####QUESTION 11
"What do each of the horizontal lines in the boxplot mean? What do the vertical lines (whiskers) mean?"

The horizontal lines represents the range of grades for each programme. The start of the line represent the lowest grade of the programme and the end(top) the highest one.
The box shows where the data is condensed. More precisely the lower horizontal line is the 25th percentile, the middle one is the median and the upper one is the 75th percentile.

###Two densities

####QUESTION 12
"Comparison of distributions across categories can also be done by adding a fill aesthetic to the density plot you made earlier. Try this out. To take care of the overlap, you might want to add some transparency in the geom_density() function using the alpha argument."

```{r}
gg_students %>% 
  ggplot(aes(x= grade, fill=prog)) +
  geom_density(alpha=0.5, colour = NA) +
  geom_rug(size = 0.75) +
  theme_minimal() +
  labs(y = "") +
  xlim(0, 10)
```

### Bar plot

####QUESTION 13
"Create a bar plot of the variable Years from the Hitters dataset."

```{r}
Hitters %>% 
  ggplot(aes(x = Years)) + 
  geom_bar() +
  theme_minimal()
```

### Line plot

####QUESTION 14
"Use geom_line() to make a line plot out of the first 200 observations of the variable Volume (the number of trades made on each day) of the Smarket dataset. You will need to create a Day variable using mutate() to map to the x-position. This variable can simply be the integers from 1 to 200. Remember, you can select the first 200 rows using Smarket[1:200, ]"

```{r}
Smarket[1:200,] %>% 
  mutate(Day = 1:200) %>% 
  ggplot(aes(x = Day, y = Volume)) +
  geom_line() +
  theme_minimal()
```

####QUESTION 15
"Give the line a nice colour and increase its size. Also add points of the same colour on top."

```{r}
Smarket[1:200,] %>% 
  mutate(Day = 1:200) %>% 
  ggplot(aes(x = Day, y = Volume)) +
  geom_line(colour = 'blue') +
  geom_point(colour = 'blue', size=1) +
  theme_minimal()
```

####QUESTION 16
"Use the function which.max() to find out which of the first 200 days has the highest trade volume and use the function max() to find out how large this volume was."

```{r}
day_max = which.max(Smarket[1:200, ]$Volume)
```
```{r}
vol_max = max(Smarket[1:200, ]$Volume)
```

The year with the maximum volume is 170, at this year the volume was 2.33

####QUESTION 17
"Use geom_label(aes(x = your_x, y = your_y, label = "Peak volume")) to add a label to this day. You can use either the values or call the functions. Place the label near the peak!"

```{r}
Smarket[1:200,] %>% 
  mutate(Day = 1:200) %>% 
  ggplot(aes(x = Day, y = Volume)) +
  geom_line(colour = "blue", size=1) +
  geom_point(colour = "blue", size=1) +
  geom_label(aes(x = day_max, y = vol_max, label = "Peak volume")) +
  theme_minimal()
```

### Faceting

####QUESTION 18
"Create a data frame called baseball based on the Hitters dataset. In this data frame, create a factor variable which splits players’ salary range into 3 categories. Tip: use the filter() function to remove the missing values, and then use the cut() function and assign nice labels to the categories. In addition, create a variable which indicates the proportion of career hits that was a home run."

```{r}
baseball <- Hitters %>% 
  filter(!is.na(Salary)) %>% 
  mutate(Salary_range = cut(Salary, breaks = 3, labels = c("Low salary", "Mid salary", "High salary")), Career_hmrun_proportion = CHmRun/CHits)
```

####QUESTION 19
"Create a scatter plot where you map CWalks to the x position and the proportion you calculated in the previous exercise to the y position. Fix the y axis limits to (0, 0.4) and the x axis to (0, 1600) using ylim() and xlim(). Add nice x and y axis titles using the labs() function. Save the plot as the variable baseball_plot."

```{r}
baseball_plot <- 
  baseball %>% 
  ggplot(aes(x = CWalks, y = Career_hmrun_proportion)) +
  geom_point() +
  ylim(0, 0.4) +
  xlim(0, 1600) + 
  theme_minimal() +
  labs(y = "Proportion of home runs",
       x = "Career number of walks")

baseball_plot
```

####QUESTION 20 
"Split up this plot into three parts based on the salary range variable you calculated. Use the facet_wrap() function for this; look at the examples in the help file for tips."

```{r}
baseball_plot + facet_wrap(facets = ~Salary_range)
```

## Final exercise

###QUESTION 21
"Create an interesting data visualisation based on the Carseats data from the ISLR package."

```{r}
summary(Carseats)
```
```{r}
carseat_income <-
  Carseats %>% 
  filter(!is.na(Income)) %>% 
  mutate(
    Income_range = cut(Income, breaks = 3, 
                       labels = c("Low income", "Mid income", "High income")),
    Career_proportion = Income/Age
  )
```


```{r}
carseat_income %>%
  ggplot(aes(x=CompPrice, y=Price, colour=Advertising)) +
  geom_point() +
  theme_minimal() + facet_wrap(~Income_range)
```

From the previous graph we can see the relation between the Price and the CompPrice, for several income categories and as a function of the advertising. First of all we can note that there is not huge difference of price between the 3 income categories, however it still seems that, in the high income category, sales are done with higher price. Also we can note that more advertisement is used in the high income category. 
