---
title: 'Practical 3 Answers '
author: "Li-Yeun Xu"
date: "02-10-2022"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
fontsize: 12pt
urlcolor: blue
mainfont: Arial
---
First, load the packages:

```{r load_packages}
library(tidyverse)
library(magrittr)
library(mice)
library(DAAG)
```

## Introduction
```{r}
head(boys)
tail(boys)
!is.unsorted(boys$age)
summary(boys)
```

1. Create a histogram of the variable age using the function geom_histogram().
```{r}
boys %>% 
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = .5)
```

2. Create a bar chart of the variable gen using the function geom_bar().
```{r}
boys %>% 
  ggplot(aes(x = gen)) +
  geom_bar(fill = "#0000FF")
```

## Assessing missing data
```{r}
md.pattern(boys)
```

3. Create a missingness indicator for the variables gen, phb and tv.
```{r}
new_boys <- boys %>% 
  mutate(mis_gen = is.na(gen), mis_phb = is.na(phb), mis_tv = is.na(tv))
```

4. Assess whether missingness in the variables gen, phb and tv is related to someones age.
```{r}
new_boys %>% 
  group_by(mis_gen) %>% 
  summarize(age = mean(age))
```

```{r}
new_boys %>% 
  group_by(mis_phb) %>% 
  summarize(age = mean(age))
```

```{r}
new_boys %>% 
  group_by(mis_tv) %>% 
  summarize(age = mean(age))
```

5. Create a histogram for the variable age, faceted by whether or not someone has a missing value on gen.
```{r}
new_boys %>% 
  ggplot(aes(x = age)) +
  geom_histogram() +
  facet_wrap(~mis_gen)
```

6. Create a scatterplot with age on the x-axis and bmi on the y-axis, using the function geom_point().
```{r}
new_boys %>% 
  ggplot(aes(x=age, y = bmi)) +
  geom_point()
```

7. Add a colour aesthetic to the previous plot using the missingness indicator of the variable gen.
```{r}
new_boys %>% 
  ggplot(aes(x=age, y = bmi, colour = mis_gen)) +
  geom_point()
```

## Visualizing the boys data

8. Visualize the relationship between reg (region) and age using a boxplot.
```{r}
new_boys %>% 
  ggplot(aes(x = reg, y = age)) +
  geom_boxplot()
```
9. Create a density plot of age, splitting the densities by gen using the fill aesthetic.
```{r}
new_boys %>% 
  ggplot(aes(x = age, fill = gen)) +
  geom_density(alpha = .5)
```

10. Create a diverging bar chart for hgt in the boys data set, that displays for every age year that year’s mean height in deviations from the overall average hgt.
```{r}
#SKIPPED
```
# Regression visualization

11. Load the data elastic1 and elastic2 and bind the data frames together using the function bind_rows() and add a grouping variable indicating whether an observation comes from elastic1 or from elastic2.

```{r}
elastic <- bind_rows("elastic1" = elastic1, "elastic2" = elastic2, .id = "Group")
```

12. Create a scatterplot mapping stretch on the x-axis and distance on the y-axis, and map the just created group indicator as the color aesthetic.

```{r}
elastic %>% 
  ggplot(aes(x=stretch, y=distance, colour=Group)) +
  geom_point()
```
13. Recreate the previous plot, but now assess whether the results of the two data sets appear consistent by adding a linear regression line.
```{r}
elastic %>% 
  ggplot(aes(x=stretch, y=distance, colour=Group)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```
14. For each of the data sets elastic1 and elastic2, fit a regression model with y = distance on x = stretch using lm(y ~ x, data).
```{r}
lm1 <- lm(distance ~ stretch, elastic1)
lm2 <- lm(distance ~ stretch, elastic2)
```

15. For both of the previously created fitted models, determine the fitted values and the standard errors of the fitted values, and the proportion explained variance R2.
```{r}
predict(lm1, se.fit = TRUE)
predict(lm2, se.fit = TRUE)
lm1 %>% summary() %$% r.squared
lm2 %>% summary() %$% r.squared
```

16. Study the residual versus leverage plots for both models.
```{r}
lm1 %>% plot(which = 5)
lm2 %>% plot(which = 5)
```

17. Use the elastic2 variable stretch to obtain predictions on the model fitted on elastic1.
```{r}
pred <- predict(lm1, elastic2)
```

18. Now make a scatterplot to investigate similarity between the predicted values and the observed values for elastic2.
```{r}
new_stretch <- data.frame(stretch = c(elastic2$stretch, elastic2$stretch), 
                          distance = c(pred,elastic2$distance),
                          group = as.factor(c(rep('P', length(pred)),rep('A', length(elastic2$stretch)))))

new_stretch %>% 
  ggplot(aes(x = stretch, y= distance, color = group)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

