---
title: "Practical 1"
author: "Ana Martins"
date: "September 2022"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
fontsize: 12pt
urlcolor: blue
mainfont: Arial
---

```{r load_packages}
library(ISLR)
library(tidyverse)
library(haven)
```

## Data Types

---

1. **Run the following code in R and inspect their data types using the `class()` function. Try to guess beforehand what their types will be!**

```{r objects}
object_1 <- 1:5
object_2 <- 1L:5L
object_3 <- "-123.456"
object_4 <- as.numeric(object_2)
object_5 <- letters[object_1]
object_6 <- as.factor(rep(object_5, 2))
object_7 <- c(1, 2, 3, "4", "5", "6")
```

```{r classes}
class(object_1)
class(object_2)
class(object_3)
class(object_4)
class(object_5)
class(object_6)
class(object_7)
```

---

2. **Convert `object_7` back to a vector of numbers using the `as.numeric()` function**

```{r object7 as numeric}
object_7 <- as.numeric(object_7)
class(object_7)
```

---

## Lists and data frames

---

3. **Make a list called objects containing object 1 to 7 using the `list()` function.**

```{r list of objects}
objects <- list(object_1, object_2, object_3, object_4, object_5, object_6, object_7)
objects[[1]]
objects[[2]]
objects[[3]]
```

---

4. **Make a data frame out of `object_1`, `object_2`, and `object_5` using the `data.frame()` function**

```{r dataframe from objects 1 2 5}
df <- data.frame(object_1, object_2, object_5)
```

---

5. **Useful functions for determining the size of a data frame are `ncol()` and `nrow()`. Try them out!**

```{r no of cols and rows}
ncol(df)
nrow(df)
```

---

## Loading, viewing, and summarising data

---

6. **Use the function `read_csv()` to import the file “data/googleplaystore.csv” and store it in a variable called `apps`.**

```{r reading apps data}
apps <- read_csv("data/googleplaystore.csv")
```

---

7. **Did any column get a variable type you did not expect?**

The price could be numeric instead of char, but having the $ on some of them makes all char.

---

8. **Use the function `head()` to look at the first few rows of the `apps` dataset**

```{r apps head}
head(apps)
```

---

9. **Repeat steps 6, 7, and 8 but now for “data/students.xlsx” (NB: You’ll need a function from the package `readxl`). Also try out the function `tail()` and `View()` (with a capital V).**

```{r students}
library(readxl)
students <- read_excel("data/students.xlsx")
head(students)
tail(students)
View(students)
```

---

10. **Create a summary of the three columns in the students dataset using the `summary()` function. What is the range of the grades achieved by the students?**

```{r students summary}
summary(students)
```

The range of grades achieved by the students is 4.844 - 9.291.

---

## Data transformation with `dplyr` verbs

### Filter

---

11. **Look at the help pages for `filter()` (especially the examples) and show the students with a grade lower than 5.5**

```{r students grade < 5.5}
filter(students, grade < 5.5)
```

---

12. **Show only the students with a grade higher than 8 from programme A**

```{r students grade > 8 programme A}
filter(students, grade > 8 & programme == "A")
```

---

### Arrange

---

13. **Sort the students dataset such that the students from programme A are on top of the data frame and within the programmes the highest grades come first.**

```{r students by programme -grade}
arrange(students, programme, -grade)
```

---

### Select

---

14. **Show only the `student_number` and `programme` columns from the students dataset**

```{r students only student_number programme}
select(students, student_number, programme)
```

---

### Mutate

---

15. **Use `mutate()` and `recode()` to change the codes in the programme column of the students dataset to their names. Store the result in a variable called `students_recoded`**

```{r students_recoded}
students_recoded <- mutate(students, programme=recode(programme, "A" = "Science", "B" = "Social Science"))
```

---

## Data processing pipelines

---

16. **Create a data processing pipeline that (a) loads the apps dataset, (b) parses the number of installs as ‘Downloads’ variable using `mutate` and `parse_number()`, (c) shows only apps with more than 500 000 000 downloads, (d) orders them by rating (best on top), and (e) shows only the relevant columns (you can choose which are relevant, but select at least the `Rating` and `Category` variables). Save the result under the name `popular_apps`.**

```{r popular apps}
popular_apps <- 
  apps %>%
  mutate(Downloads = parse_number(Installs)) %>%
  filter(Downloads > 500000000) %>%
  arrange(-Rating) %>%
  select(App, Category, Rating) %>% 
  distinct(App, .keep_all = TRUE)
popular_apps
```

---

## Grouping and Summarisation

---

17. **Show the median, minimum, and maximum for the popular apps dataset you made in the previous assignment.**

```{r popular_apps summary}
popular_apps %>%
  summarise(
    median = median(Rating),
    minimum = min(Rating),
    maximum = max(Rating)
  )
```

---

18. **Add the median absolute deviation to the summaries you made before**

```{r popular_apps summary with mad}
mad <- function(x) {
  median(abs(x - median(x)))
}
popular_apps %>%
  summarise(
    median = median(Rating),
    minimum = min(Rating),
    maximum = max(Rating),
    mad = mad(Rating)
  )
```

---

19. **Create a grouped summary of the ratings per category in the popular apps dataset.**

```{r popular_apps grouped}
popular_apps %>% 
  group_by(Category) %>% 
  summarise(
    median = median(Rating),
    minimum = min(Rating),
    maximum = max(Rating),
    mad = mad(Rating)
  )
```

---

20. **Create an interesting summary based on the Google play store apps dataset. An example could be “do games get higher ratings than communication apps?”**

With this summary, we are asking and answering the questions "Does the price affect the number of Downloads? Do free apps get more downloads?":

```{r apps summary}

apps %>% 
  mutate(Downloads = parse_number(Installs), NPrice = parse_number(Price)) %>% 
  group_by(NPrice) %>% 
  summarise(
    mean = mean(Downloads),
    maximum = max(Downloads),
    minimum = min(Downloads)
  ) %>% 
  arrange(NPrice)

```

The answer to the second question is: definitely yes! The free apps have the biggest mean and maximum for downloads.
The answer to the first question is: kind of. Apparently, as long as people are paying, be it 99 cents for 400 euros, the downloads are not inversely proportional (nor proportional) to the price.