---
title: "Supervised Learning and Visualisation"
author: "Willem van Veluw"
date: "19-9-2022"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
mainfont: Arial
fontsize: 12pt
urlcolor: blue
subtitle: Practical 3
---
For this practical we first load the necessary packages.
```{r, warning = FALSE, message = FALSE}
library(tidyverse)
library(magrittr)
library(mice)
library(DAAG)
```

### Exercise 1
```{r}
boys %>% ggplot(aes(x = age)) + geom_histogram() + theme_minimal()
```

### Exercise 2
```{r}
boys %>% ggplot(aes(x = gen)) + 
  geom_bar() + 
  labs(x = "Genital Tanner Stage") +
  theme_minimal()
```

### Exercise 3
```{r}
boysMiss <-
  boys %>% 
  mutate(genMiss = is.na(gen), phbMiss = is.na(phb), tvMiss = is.na(tv))
```

### Exercise 4
We see that the boys with missing data for either `gen`, `phb` or `tv` have a higher age than the boys with no missing data.
```{r}
boysMiss %>% 
  group_by(genMiss) %>% 
  summarize(age = mean(age))

boysMiss %>% 
  group_by(phbMiss) %>% 
  summarize(age = mean(age))

boysMiss %>% 
  group_by(tvMiss) %>% 
  summarize(age = mean(age))
```

### Exercise 5
```{r}
boysMiss %>% 
  ggplot(aes(x = age, fill = genMiss)) +
  geom_histogram() +
  guides(fill = "none") +
  facet_wrap(vars(genMiss)) +
  labs(title = "Histogram of age, splitted on wether or not gen is missing.",
       subtitle = "gen = Genital Tanner Stage") +
  theme_minimal()
```

### Exercise 6
```{r}
boysMiss %>% 
  ggplot(aes(x = age, y = bmi)) +
  geom_point() +
  theme_minimal()
```

### Exercise 7
```{r}
boysMiss %>% 
  ggplot(aes(x = age, y = bmi, colour = genMiss)) +
  geom_point() +
  labs(colour = "gen is missing") +
  theme_minimal()
```

### Exercise 8
```{r}
boys %>% 
  ggplot(aes(x = reg,y = age, fill = reg)) +
  geom_boxplot() +
  labs(x = "Region") +
  theme(legend.position = "none")
```

### Exercise 9
```{r}
boys %>% 
  ggplot(aes(x = age, fill = gen)) +
  geom_density(alpha = 0.5) +
  theme_minimal()
```

### Exercise 10
```{r}
boysAge <- boys %>% 
  mutate(Age = cut(age, breaks = c(0:22), labels = paste0(0:21))) %>% 
  mutate(Height = hgt - mean(hgt, na.rm = TRUE)) %>% 
  group_by(Age) %>% 
  summarise(Height = mean(Height, na.rm = TRUE)) %>% 
  mutate(hgtAvg = ifelse(Height > 0, "Above average", "Below average"))

boysAge %>% ggplot(aes(x = Age, y = Height, fill = hgtAvg)) +
  geom_bar(stat = "identity") +
  labs(fill = "") +
  scale_fill_manual(values = c("darkgreen", "darkred")) +
  theme_minimal()
```

### Exercise 11
```{r}
elastic1$id <- rep("1", nrow(elastic1))
elastic2$id <- rep("2", nrow(elastic2))
elastic <- bind_rows(elastic1, elastic2) %>% 
  mutate(id = as.factor(id))
```

### Exercise 12
```{r}
elastic %>% ggplot(aes(x = stretch, y = distance, colour = id)) +
  geom_point() +
  labs(colour = "From dataset") +
  theme_minimal()
```

### Exercise 13

```{r}
elastic %>% ggplot(aes(x = stretch, y = distance, colour = id)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(colour = "From dataset") +
  theme_minimal()
```

### Exercise 14
```{r}
model1 <- elastic1 %>% lm(distance ~ stretch, .)
model2 <- elastic2 %>% lm(distance ~ stretch, .)
```

### Exercise 15
The $R^2$ can be read off from the output of `summary()`.  
The $R^2$ for the model on `elastic1` is 0.7992.  
The $R^2$ for the model on `elastic2` is 0.9808.
```{r}
#Predictions for elastic1
preds1 <- predict(model1, se.fit = TRUE)
preds1

#Predictions for elastic2
preds2 <- predict(model2, se.fit = TRUE)
preds2
```

### Exercise 16
In the model for `elastic1`, we see that observation 2 has the largest effect on estimation. In the model for `elastic2` the observation with largest effect is observation 1.
```{r}
plot(model1, which = 5)
plot(model2, which = 5)
```

### Exercise 17
```{r}
preds <- predict(model1, newdata = elastic2)
preds
```

### Exercise 18
```{r}
elastic2$State <- rep("Observed", nrow(elastic2))
elastic2 %>% select(stretch, distance, State)
predic2 <- data.frame(
  stretch = elastic2$stretch,
  distance = preds,
  State = rep("Predicted", length(preds))
)

data <- bind_rows(elastic2, predic2)
data %>% ggplot(aes(x = stretch, y = distance, colour = State)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  labs(colour = "") +
  theme_minimal()
```

