---
title: "Practical 1: Data Wrangling"
author: "Nina van Gerwen"
date: "13-09-2018"
mainfont: Arial
fontsize: 12pt
urlcolor: blue
output: 
  pdf_document:
    latex_engine: xelatex
---

First, load the packages:

```{r load_packages, warning=FALSE}
library(ISLR)
library(tidyverse)
library(haven)
library(readxl)
```

### 1. Objects and classes

```{r objects}
object_1 <- 1:5 ## integers
object_2 <- 1L:5L ## integers
object_3 <- "-123.456" ## character
object_4 <- as.numeric(object_2) ## numeric
object_5 <- letters[object_1] ## characters
object_6 <- as.factor(rep(object_5, 2)) ## factors
object_7 <- c(1, 2, 3, "4", "5", "6") ## characters
```

### 2. Converting

```{r converting}
object_7 <- as.numeric(object_7)
```

### 3. Making a list

```{r listing}
objects <- list(object_1, object_2, object_3, object_4, object_5,
                  object_6, object_7)
```

### 4. Making a data frame

```{r framing}
object_frame <- data.frame(object_1, object_2, object_5)
```

### 5. Determining the size of a data frame

```{r}
ncol(object_frame) ## 3
nrow(object_frame) ## 5
```

## Loading, viewing and summarising data

### 6. Importing google data

```{r importing data, warning=FALSE}
apps <- read_csv("Data/googleplaystore.csv")
```

### 7. Unexpected results

```{r}
str(apps)
```
The variables 'Category', 'Size', 'Installs', 'Price', 'Content Rating',
'Genres' and 'Android Ver' should probably be a Factor instead of a Character.

### 8. The head

```{r}
head(apps)
```

### 9. Students data

```{r}
## Loading data
students <- read_xlsx("Data/students.xlsx")

## Seeing the class of every variable
str(students)

head(students)
```

Again, the variable 'Programme' should probably be a factor. 

### 10. Summarising 

```{r}
summary(students)
```
The grades achieved by students ranges from 4.84 to 9.29 with an average
grade of 7.15.

## Data transformation

### 11. Filtering

```{r}
students %>% filter(., grade < 5.5)
```

### 12. More filtering

```{r}
students %>% filter(., grade > 8 & programme == "A")
```

### 13. Arranging

```{r}
students %>% arrange(., programme, desc(grade))
```
We use desc(grade) because we want the order to be descending and not
ascending.

### 14. Selecting

```{r}
students %>% select(., student_number, programme)
```
### 15. Mutating

```{r}
students_recoded <- students %>% mutate(., programme = recode(programme, 
                                        A = "Science", B = "Social Science"))
```

So the above code states: students_recoded will be students, where
we mutated the column programme such that A = Sciencce, B = Social Science.

## Data processing pipelines

### 16. Pipelining

```{r}
popular_apps <-
  read_csv("Data/googleplaystore.csv") %>%
  mutate(Downloads = parse_number(Installs)) %>%
  filter(Downloads > 500000000) %>%
  arrange(desc(Rating)) %>%
  select(Rating, Category, Price, App) %>%
  distinct(App, .keep_all = TRUE)

popular_apps
```

## Grouping and summarisation

### 17 & 18. Summarising

The median of which variable exactly? I'm guessing Rating.

```{r}
mad <- function(x){
  median(abs(x - median(x)))
}
popular_apps %>%
  summarise(
    median = median(Rating),
    min = min(Rating),
    max = max(Rating),
    mad = mad(Rating)
  )
```

### 19. Grouped summary

```{r}
popular_apps %>%
  group_by(Category) %>%
  summarise(
    median = median(Rating),
    min = min(Rating),
    max = max(Rating),
    mad = mad(Rating)
  )
```

## Final Exercise

Summary of how many Downloads per category:

```{r}
own_summary <-
  read_csv("Data/googleplaystore.csv") %>%
  mutate(Downloads = parse_number(Installs)) %>%
  group_by(Category) %>%
  summarise(
    mean = mean(Downloads),
    min = min(Downloads),
    max = max(Downloads),
    var = var(Downloads)
  ) %>%
  arrange(desc(mean))

own_summary
```
From this summary, we can see that on average 'Communication' apps get the most downloads. However, we do not know whether the difference between the other apps categories and Communication is significant (though I would bet good money on it).

