---
title: "P1_AmaliaTsakali"
author: "Amalia Tsakali"
date: "18-9-2022"

---


First, load the packages:

```{r load_packages}
library(ISLR)
library(tidyverse)
library(haven)
library(readxl)
```

1. Objects and classes

```{r objects}
object_1 <- 1:5
object_2 <- 1L:5L
object_3 <- "-123.456"
object_4 <- as.numeric(object_2)
object_5 <- letters[object_1]
object_6 <- as.factor(rep(object_5, 2))
object_7 <- c(1, 2, 3, "4", "5", "6")
```
Inspect the data types of the objects
```{r classes}
class(object_1)
# integer
class(object_2)
# integer
class(object_3)
# character
class(object_4)
# numeric
class(object_5)
# character
class(object_6)
# factor
class(object_7)
# character

```

2. converting object_7 to numeric
```{r 2}
object_7 <- as.numeric(object_7)
```

3.Create a list containing the 7 objects
```{r 3}
objects <- list(object_1, object_2, object_3, object_4, object_5, object_6, object_7)
```

4. create a dataframe from objects 1,2 and 5
```{r 4}
df1<-data.frame(object_1,object_2,object_5)
```
5.finding the size of the dataframe we created
```{r 5}
ncol(df1)
nrow(df1)
```
6. Import dataset

```{r 6}
apps <- read_csv("data/googleplaystore.csv")
```
7. check the variable types of the columns
```{r 7}
# We can see from the warning message that several columns that we would expect to have numeric values have character data type (such as size,price etc)
```
8.Viewing the first few rows of the dataset
```{r 8}
head(apps)
```
9. Repeat the previous steps for the dataset "data/students.xlsx"
```{r 9}
students <- read_xlsx("data/students.xlsx")
nrow(students)
ncol(students)
head(students)
tail(students)
View(students)
```
10. summary for the columns
```{r 10}
summary(students)
#the range of the grades= 4.844 - 9.291
```
11. show students with a grade less than 5.5
```{r 11}
filter(students,grade<5.5)
```
12. Filter students with higher than 8 for programme A
```{r 12}
filter(students, grade>8 & programme=="A")

```

13. arrange the dataset so that students from programme A are first, and within that goup students with the highest grades are first.
```{r 13}
arrange(students,programme,desc(grade))
```
14. show only the student number and the programme columns
```{r 14}
select(students, student_number, programme)
```
14b. create a logical column indicating whether a student passed or failed
```{r 14b}
students<-mutate(students, pass=grade>5.5)
students
```
15. change the codes of the programme columns with recode and mutate
```{r 15}
students_recoded <-mutate(students,programme = recode(programme, "A" = "Science", "B" = "Social Science"))
```
16. Create a data processing pipeline that (a) loads the apps dataset, (b) parses the number of installs as 'Downloads' variable using `mutate` and `parse_number()`, (c) shows only apps with more than 500 000 000 downloads, (d) orders them by rating (best on top), and (e) shows only the relevant columns (you can choose which are relevant, but select at least the `Rating` and `Category` variables). Save the result under the name `popular_apps`
```{r 16}
popular_apps <- read_csv("data/googleplaystore.csv") %>%
  mutate(Downloads=parse_number(Installs)) %>%
  filter(Downloads > 500000000)%>%
  arrange(desc(Rating)) %>%
  select(App,Category,Rating,Downloads)%>% 
  distinct(App, .keep_all = TRUE)
popular_apps
```
17.Show the median, minimum, and maximum for the popular apps dataset you made in the previous assignment
```{r 17}
popular_apps %>%
  summarise(
    med = median(Rating),
    min = min(Rating), 
    max = max(Rating)
  )
```
18.Add the median absolute deviation to the summaries you made before
```{r 18}
mad <- function(x) {
  median(abs(x - median(x)))
}
popular_apps %>%
  summarise(
    med = median(Rating),
    min = min(Rating), 
    max = max(Rating),
    mad = mad(Rating)
  )
```
19. Create a grouped summary of the ratings per category in the popular apps dataset
```{r 19}
popular_apps %>% 
  group_by(Category) %>% 
  summarise(
    med = median(Rating),
    min = min(Rating), 
    max = max(Rating),
    mad = mad(Rating)
  )
```
20. Create an interesting summary based on the Google play store apps dataset. An example could be "do games get higher ratings than communication apps?"
```{r 20}
apps%>% 
  filter(Category=="GAME"|Category=="COMMUNICATION") %>% 
  group_by(Category) %>%
  distinct(App, .keep_all = TRUE) %>%
  summarise(
    med = median(Rating,na.rm = TRUE),
    min = min(Rating,na.rm = TRUE), 
    max = max(Rating,na.rm=TRUE),
    mean(Rating,na.rm=TRUE)
  )
```

