---
title: "Assignment1 - RHAT"
output: html_document
date: "September 2022"
---

# NYC Rat Sightings

## Data Description

The NYC Rat Sightings dataset describes mostly the location where there are rat sightings, the dates when it happened and which type of building it happened on. In the original dataset there are also many empty columns or columns with only one value, which just means the data was organised poorly, and we will not use these.

We will analyse if sightings depend on the month they happen. Additionally, we will analyse a dataset that describes borough populations so we can analyse if population density compares to number of rat sightings and which boroughs have more density of rat sightings. 

The dataset we will be using can be found on [kaggle](https://www.kaggle.com/datasets/new-york-city/nyc-rat-sightings).

```{r}
library(tidyverse)
```

## Data cleaning

First, we create a function to use for splitting up the different dates.

```{r}
date.split <- function(txtdate, x){
  date <- parse_date(txtdate, format = "%m/%d/%Y %I:%M:%S %p")
  time <- parse_time(txtdate, format = "%m/%d/%Y %I:%M:%S %p")
  year <- as.factor(format(date, format = "%Y"))
  month <- as.factor(format(date, format = "%m"))
  day <- as.numeric(format(date, format = "%d"))
  
  if (x == 1) {value <- year}
  else if (x == 2) {value <- month}
  else if (x == 3) {value <- day}
  else if (x == 4) {value <- time}
  
  return(value)
}
```


Secondly, we create a function to use when splitting a specific number from a data point, as used for both the address and the community boards.

```{r}
#Second attempt at a function that could split "Incident_Address" and "Community Board":
get.number <- function(address){
  if(!identical(address, NA)){
    nums <- str_extract(address, "[0-9]+")
    return(nums)
  }
}
```

The dataset initially looks like this:

```{r}
rat_dataset <-
  read_csv("Rat_Sightings.csv", show_col_types = FALSE)
head(rat_dataset)
```

with many empty features or with one single value. In this next step, we will be taking these features out and separating some interesting data in more features:

```{r}
rat_dataset <-
  read_csv("Rat_Sightings.csv", show_col_types = FALSE) %>% 
  #select("Unique Key", "Created Date", "Closed Date", "Location Type", "Incident Zip", "Incident Address", "Street Name", "Cross Street 1", "Cross Street 2", "Intersection Street 1", "Intersection Street 2", "Address Type", "Status", "Due Date", "Resolution Action Updated Date", "Community Board", "Borough", "Latitude", "Longitude") %>% 
  rename(Unique_Key = "Unique Key", Created_Date = "Created Date", Closed_Date = "Closed Date", Location_Type = "Location Type", Incident_Zip = "Incident Zip", Incident_Address = "Incident Address", Street_Name = "Street Name", Cross_Street_1 = "Cross Street 1", Cross_Street_2 = "Cross Street 2", Intersection_Street_1 = "Intersection Street 1", Intersection_Street_2 = "Intersection Street 2", Address_Type = "Address Type", Status = "Status", Due_Date = "Due Date", RAU_Date = "Resolution Action Updated Date", Community_Board = "Community Board", Borough = "Borough", Latitude = "Latitude", Longitude = "Longitude") %>% 
  mutate(Created_Year = date.split(Created_Date, 1), Created_Month = date.split(Created_Date, 2), Created_Day = date.split(Created_Date, 3)) %>% 
  mutate(Closed_Year = date.split(Closed_Date, 1), Closed_Month = date.split(Closed_Date, 2), Closed_Day = date.split(Closed_Date, 3)) %>%
  mutate(Due_Year = date.split(Due_Date, 1), Due_Month = date.split(Due_Date, 2), Due_Day = date.split(Due_Date, 3), Due_Time = date.split(Due_Date, 4)) %>%
  mutate(RAU_Year = date.split(RAU_Date, 1), RAU_Month = date.split(RAU_Date, 2), RAU_Day = date.split(RAU_Date, 3), RAU_Time = date.split(RAU_Date, 4)) %>% 
  mutate(Address_Number = get.number(Incident_Address)) %>% 
  mutate(Community_Board_Number = get.number(Community_Board)) %>% 
  select(Unique_Key, Created_Year, Created_Month, Created_Day, Closed_Year, Closed_Month, Closed_Day, Location_Type, Incident_Zip, Address_Number, Street_Name, Cross_Street_1, Cross_Street_2, Intersection_Street_1, Intersection_Street_2, Address_Type, Status, Due_Year, Due_Month, Due_Day, Due_Time, RAU_Year, RAU_Month, RAU_Day, RAU_Time, Community_Board_Number, Borough, Latitude, Longitude)
```

We also want to make sure that all of the features have the proper value type.

```{r}
rat_dataset$Location_Type <- 
  as.factor(rat_dataset$Location_Type)
rat_dataset$Street_Name <-
  as.factor(rat_dataset$Street_Name)
rat_dataset$Cross_Street_1 <-
  as.factor(rat_dataset$Cross_Street_1)
rat_dataset$Cross_Street_2 <-
  as.factor(rat_dataset$Cross_Street_2)
rat_dataset$Address_Type <-
  as.factor(rat_dataset$Address_Type)
rat_dataset$Status <-
  as.factor(rat_dataset$Status)
rat_dataset$Borough <-
  as.factor(rat_dataset$Borough)
```


So, our starting point data is:

```{r}
head(rat_dataset)
summary(rat_dataset)
```

## Where are the sightings more concentrated?

To answer this question, we will import an extension of the `ggplot2` library, the `ggmap` library. If you don't have the `ggmap` package installed, please do so by running `install.packages("ggmap")` on your R console. 

```{r}
library("ggmap")
```

```{r}
KEY = 
register_google(key = KEY)
```

Firstly, we take a look at how the signthings are distributed:

```{r}
ggplot(rat_dataset, mapping = aes(y = Longitude)) +
  geom_boxplot()
ggplot(rat_dataset, mapping = aes(y = Latitude)) +
  geom_boxplot()
```

Translating this into a map, we get:

```{r}
data4lat <-
  rat_dataset %>% 
  select(Latitude) %>% 
  filter(!is.na(Latitude))

max_lat <- max(data4lat)
min_lat <- min(data4lat)

data4lon <-
  rat_dataset %>% 
  select(Longitude) %>% 
  filter(!is.na(Longitude))

max_lon <- max(data4lon)
min_lon <- min(data4lon)

newyork.map <- get_map(location= 'New York')
sightings_map <-
  ggmap(newyork.map) +
  ylim(min_lat, max_lat) +
  xlim(min_lon, max_lon)
sightings_map
```

From the distribution on the boxplots, we can assume that most of the sightings are concentrated in the middle of the map, which we can confirm by plotting the sightings:

```{r}
sightings_map <-
  ggmap(newyork.map) +
  geom_point(data = rat_dataset, mapping = aes(x = Longitude, y = Latitude), alpha = 0.1) +
  stat_density2d(data = rat_dataset, mapping = aes(x = Longitude, y = Latitude)) +
  ylim(min_lat, max_lat) +
  xlim(min_lon, max_lon)
sightings_map
```

And what type of location do these happen in, and how are they distributed on the map?

```{r}
location_map <-
  ggmap(newyork.map) +
  geom_point(data = rat_dataset, mapping = aes(x = Longitude, y = Latitude, color = Location_Type), alpha = 0.3) +
  facet_wrap(~ Location_Type) +
  ylim(min_lat, max_lat) +
  xlim(min_lon, max_lon)
location_map
```

That does not seem like a great visualization (as you really can't see anything), let's divide our facets into multiple plots so we can take a closer look at each of them.

```{r}
location_subset1 <- subset(rat_dataset, Location_Type %in% c("1-2 Family Dwelling", "1-2 Family Mixed Use Building", "3+ Family Apt. Building"))
location_map1 <-
  ggmap(newyork.map) +
  geom_point(data = location_subset1, mapping = aes(x = Longitude, y = Latitude, color = Location_Type), alpha = 0.3) +
  facet_wrap(~ Location_Type) +
  ylim(min_lat, max_lat) +
  xlim(min_lon, max_lon) +
  scale_color_brewer(palette = "Dark2")
location_map1

location_subset2 <- subset(rat_dataset, Location_Type %in% c("3+ Family Mixed Use Building", "Catch Basin/Sewer", "Commercial Building"))
location_map2 <-
  ggmap(newyork.map) +
  geom_point(data = location_subset2, mapping = aes(x = Longitude, y = Latitude, color = Location_Type), alpha = 0.3) +
  facet_wrap(~ Location_Type) +
  ylim(min_lat, max_lat) +
  xlim(min_lon, max_lon) +
  scale_color_brewer(palette = "Paired")
location_map2

location_subset3 <- subset(rat_dataset, Location_Type %in% c("Construction Site", "Day Care/Nursery", "Government Building"))
location_map3 <-
  ggmap(newyork.map) +
  geom_point(data = location_subset3, mapping = aes(x = Longitude, y = Latitude, color = Location_Type), alpha = 0.3) +
  facet_wrap(~ Location_Type) +
  ylim(min_lat, max_lat) +
  xlim(min_lon, max_lon) +
  scale_color_brewer(palette = "Set2")
location_map3

location_subset4 <- subset(rat_dataset, Location_Type %in% c("Hospital", "Office Building", "Other (Explain Below)"))
location_map4 <-
  ggmap(newyork.map) +
  geom_point(data = location_subset4, mapping = aes(x = Longitude, y = Latitude, color = Location_Type), alpha = 0.3) +
  facet_wrap(~ Location_Type) +
  ylim(min_lat, max_lat) +
  xlim(min_lon, max_lon) +
  scale_color_brewer(palette = "Accent")
location_map4

location_subset5 <- subset(rat_dataset, Location_Type %in% c("Parking Lot/Garage", "Public Garden", "Public Stairs"))
location_map5 <-
  ggmap(newyork.map) +
  geom_point(data = location_subset5, mapping = aes(x = Longitude, y = Latitude, color = Location_Type), alpha = 0.3) +
  facet_wrap(~ Location_Type) +
  ylim(min_lat, max_lat) +
  xlim(min_lon, max_lon) +
  scale_color_brewer(palette = "Pastel1")
location_map5

location_subset6 <- subset(rat_dataset, Location_Type %in% c("School/Pre-School", "Single Room Occupancy (SRO)", "Summer Camp"))
location_map6 <-
  ggmap(newyork.map) +
  geom_point(data = location_subset6, mapping = aes(x = Longitude, y = Latitude, color = Location_Type), alpha = 0.3) +
  facet_wrap(~ Location_Type) +
  ylim(min_lat, max_lat) +
  xlim(min_lon, max_lon) +
  scale_color_brewer(palette = "Set1")
location_map6

location_subset7 <- subset(rat_dataset, Location_Type %in% c("Vacant Building", "Vacant Lot", "NA"))
location_map7 <-
  ggmap(newyork.map) +
  geom_point(data = location_subset7, mapping = aes(x = Longitude, y = Latitude, color = Location_Type), alpha = 0.3) +
  facet_wrap(~ Location_Type) +
  ylim(min_lat, max_lat) +
  xlim(min_lon, max_lon) +
  scale_color_brewer(palette = "Set3")
location_map7
```

```{r}

#since brooklyn is the borough with more rats, it might be useful to take a closer look at the map

data4lat_brooklyn <-
  rat_dataset %>% 
  filter(!is.na(Latitude)) %>% 
  filter(Borough == "BROOKLYN") %>% 
  select(Latitude)

max_brooklyn_lat <- max(data4lat_brooklyn)
min_brooklyn_lat <- min(data4lat_brooklyn)

data4lon_brooklyn <-
  rat_dataset %>% 
  filter(!is.na(Longitude)) %>% 
  filter(Borough == "BROOKLYN") %>% 
  select(Longitude)

max_brooklyn_lon <- max(data4lon_brooklyn)
min_brooklyn_lon <- min(data4lon_brooklyn)

brooklyn_rats <-
  rat_dataset %>% 
  filter(Borough == "BROOKLYN")

brooklyn_map <-
  ggmap(newyork.map) +
  geom_point(data = brooklyn_rats, mapping = aes(x = Longitude, y = Latitude), color = "Red", size = 0.1) +
  ylim(min_brooklyn_lat, max_brooklyn_lat) +
  xlim(min_brooklyn_lon, max_brooklyn_lon)
brooklyn_map
```

## Location Type and Rat number

**Check how many location types it is.**
```{r}
unique(rat_dataset$Location_Type)
# 20 location types
```

There are 20 location types.

**Compute the percentage of rat sighting in each location type**
```{r}
loc_per <- rat_dataset %>%
  group_by(Location_Type) %>%
  summarise(Percentage = round(100*(n() / nrow(.)), 1))

loc_per
```

**Plot the bar chart that reveals the occupation of rat sighting in each location type**
```{r, fig.width= 10, fig.height=7}

library(ggcharts)

loc_per %>%
  bar_chart(Location_Type, Percentage,
            bar_color = c("#882E72", "#B178A6", "#114477", "#1965B0", 
                          "#5289C7", "#7BAFDE", "#4EB265", "#90C987", 
                          "#CAE0AB", "#F7EE55", "#F6C141", "#F1932D", 
                          "#E8601C", "#DC050C", "#D6C1DE", "#4477AA", 
                          "#77AADD", "#117755", "#44AA88", "#99CCBB", 
                          "#777711")) +
  geom_text(aes(label = paste(Percentage , "%")),
            hjust = -0.2) +
  ylim(NA, 45) +
  theme(plot.title = element_text(hjust=0.5)) +
  labs(x = "", y = "", title = "Percentage of Rat Sighting in Different Location Type",
       fill = "Location Type")
  
```

Following the implications of the bar chart, the most frequent cases of rat sightings occurred in apartment buildings with 3 or more family members, accounting for 40.3%. This is followed by residences with 1 to 2 family members at 19.3%. Notably, mixed-use buildings with 3 or more family members also accounted for 7.8%. Overall, the types of locations associated with families held the most complaints of rat sightings.

## Time and Rat number

**Plot the number of rat sightings each month**
```{r}
rat_dataset %>%
  ggplot(aes(x = Created_Month, fill = Created_Month)) +
  geom_bar() +
  geom_text(stat='count', aes(label=..count..), vjust=-1) +
  scale_fill_manual(values=c("#882E72", "#B178A6", "#114477", "#1965B0", 
                             "#5289C7", "#7BAFDE", "#4EB265", "#90C987", 
                             "#CAE0AB", "#F7EE55", "#F6C141", "#F1932D", 
                             "#E8601C", "#DC050C", "#D6C1DE", "#4477AA", 
                             "#77AADD", "#117755", "#44AA88", "#99CCBB", 
                             "#777711")) +
  ylim(0, 15000) +
  labs(x = "Month", y = "Total Number of Rat Sighting", 
       title = "Number of Rat Sighting Each Month", fill = "Month") +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5))
```

In July, the highest number of rat sightings was recorded, with 11,982 cases, followed by August and June. In general, the number of rat sightings can be identified in relation to fluctuations in temperature, as in winter months, i.e. November, December, January and February, the number of rat sightings are the lowest. With rising temperatures, however, the number of rat sightings increased markedly.

**Plot the Percentage of rat sightings by month**
```{r}
# compute the percentage of each month on rat sighting
month_per <- rat_dataset %>%
  group_by(Created_Month) %>%
  summarise(Percentage = round(100*(n() / nrow(.)),1))

month_per
```

```{r}
month_per %>%
  mutate(Month = Created_Month,
         Month = recode_factor(Month, '01' = "January", '02' = "Febuary",
                        '03' = "March", '04' = "April", '05' = "May", '06' = "June",
                        '07' = "July", '08' = "August", '09' = "September", 
                        '10' = "October", '11' = "November", '12' = "December")) %>%
  bar_chart(Month, Percentage,
            bar_color = c("#882E72", "#B178A6", "#114477", "#1965B0", 
                          "#5289C7", "#7BAFDE", "#4EB265", "#90C987", 
                          "#CAE0AB", "#F7EE55", "#F6C141", "#F1932D")) +
  geom_text(aes(label = paste(Percentage , "%")),
            hjust = -0.2) +
  ylim(NA, 15) +
  theme(plot.title = element_text(hjust=0.5)) +
  labs(x = "", y = "Percentage", title = "Percentage of Rat Sighting in Each Month",
       fill = "Month")
```

A monthly percentage of rat sightings clearly reveals that rat sightings can be roughly divided by season. Rat sightings are highest in the summer months, followed by spring and fall, and lowest in the winter months.

## Determine whether the cases of rat sightings evolve with the year

**Plot the number of rat sightings versus year** 
```{r}
rat_dataset %>%
  ggplot(aes(x = Created_Year, fill = Created_Year)) +
  geom_bar() +
  geom_text(stat='count', aes(label=..count..), vjust=-1) +
  scale_fill_manual(values=c("#882E72", "#B178A6", "#5289C7", "#7BAFDE", 
                             "#CAE0AB", "#F7EE55", "#E8601C", "#DC050C"))  +
  ylim(0, 20000) +
  labs(x = "Year", y = "Total Number of Rat Sighting", 
       title = "Number of Rat Sighting 2010 - 2017", fill = "Year") +
  theme_classic() +
  theme(plot.title = element_text(hjust=0.5))
```

Since 2010 to 2013, the number of rat sighting cases did not fluctuate measurably. Yet, the number of rat sightings escalated yearly after 2013, from 2014 to 2016, with the number of rat sightings growing from 10,534 cases in 2010 to 17,230 cases in 2016. Possible reasons behind this are that New York residents tend to report rat sightings to the administration year by year or that the population of New York rises over the period of the years, thus causing the rat population to rise with it. 

Another thing that needs to be mentioned here is that data collection was interrupted in September 2017, hence the drop in rat sightings in 2017.

**Plot the line chart of rat sightings versus year to more easily observe changes**
```{r}
# Year with rat sighting (count)
year_count <- rat_dataset %>%
  group_by(Created_Year) %>%
  summarise(Number = n())

year_count
```

```{r}
# Year with rat sighting (count)
year_count %>%
  ggplot(aes(x = Created_Year, y = Number, group = 1)) +
  geom_point(size = 1.5) +
  geom_line(color = "dark blue", size = 1) +
  labs(x = "Year", y = "Number of Rat Sighting", 
       title = "Number of Rat Sighting 2010 - 2017", fill = "Year") +
  theme_classic() +
  theme(plot.title = element_text(hjust=0.5))
```

The line graph offers a better experience in observing the trend of rat sightings by year. The line graph reaffirms the above discussion, with an alarming increase in rat sightings from 2013 to 2016.

**Mapping the number of rat sightings for each month of each year**
```{r}
# Number of Rat Sighting in each Months of Each Year
year_month_count <- rat_dataset %>%
  group_by(Created_Year) %>%
  count(Created_Month)
```

```{r}
# Number of Rat Sighting in each Months of Each Year
year_month_count %>%
  ggplot(aes(x = Created_Month, y = n, color = Created_Year, group = Created_Year)) +
  geom_point() +
  geom_line() +
  scale_color_manual(values=c("#882E72", "black", "#EEA236", "#5CB85C", 
                              "#46B8DA", "#9632B8", "dark green", "#DC050C")) +
  labs(x = "Month", y = "Number of Rat Sighting", 
       title = "Number of Rat Sighting", color = "Year") +
  theme_classic() +
  theme(plot.title = element_text(hjust=0.5))
```






```{r}
# combine the year, month, and day of started day
starttime <- as.Date(with(rat_dataset, paste(Created_Year, Created_Month, 
                                             Created_Day, sep = "-")), "%Y-%m-%d")

# combine the year, month, and day of closed day
endtime <- as.Date(with(rat_dataset, paste(Closed_Year, Closed_Month, 
                                             Closed_Day, sep = "-")), "%Y-%m-%d")

# compute for the duration taken to solve
solved_duration <- data.frame(difftime(endtime,starttime,units = "days"))
colnames(solved_duration)[1] <- "duration"
solved_duration$duration <- as.numeric(solved_duration$duration)

# merge it to the original dataset
rat_dataset <- cbind(rat_dataset, solved_duration)

# merge year created year with solved time, in order to see if the government's efficient improve
solved_duration <- cbind(rat_dataset$Created_Year, solved_duration)
colnames(solved_duration)[1] <- "Sighting_Year"

# remove the negative number, as it cannot be negative time to solve the issue
solved_duration <- solved_duration %>%
  drop_na() %>%
  filter(duration > 0) %>%
  mutate(month_cat = cut(duration, breaks = c(0, 30, 60, 90, 120, 150, 180, 1000), 
                          labels = c("less than 1 month", "less than 2 month",
                               "less than 3 month", "less than 4 month",
                               "less than 5 month", "less than half year",
                               "more than half year")))
```

```{r}
# plot the duration taken to solved the issue
solved_duration %>%
  ggplot(aes(x = month_cat, fill = month_cat)) +
  geom_bar() +
  geom_text(stat='count', aes(label=..count..), vjust=-1) +
  scale_fill_manual(values=c("#882E72", "#5289C7", "#CAE0AB", 
                             "#E8601C", "#DC050C", "#D6C1DE", "#4477AA")) +
  ylim(0, 70000) +
  labs(x = "Time to Solve", y = "Cases Count", 
       title = "Time to Solve Rat Sighting", fill = "Month") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, 
                                   hjust = 1, vjust = 0),
        plot.title = element_text(hjust=0.5))
```

```{r, fig.width=10, fig.height=7}
# plot the duration taken to solved the issue of each year
solved_duration %>%
  filter(Sighting_Year == c(2010, 2011, 2012, 2013)) %>%
  ggplot(aes(x = month_cat, fill = month_cat)) +
  geom_bar(position=position_dodge(preserve = "single"), stat='count') +
  geom_text(stat='count', aes(label=..count..), vjust=-1) +
  scale_fill_manual(values=c("#882E72", "#5289C7", "#CAE0AB", 
                             "#E8601C", "#DC050C", "#D6C1DE", "#4477AA")) +
  ylim(0, 2300) +
  labs(x = "Time to Solve", y = "Cases Count", 
       title = "Time to Solve Rat Sighting", fill = "Month") +
  facet_grid(.~Sighting_Year) +
  theme(axis.text.x = element_text(angle = 90, 
                                   hjust = 1, vjust = 0),
        plot.title = element_text(hjust=0.5))
```

```{r, fig.width=10, fig.height=7}
# plot the duration taken to solved the issue of each year
solved_duration %>%
  filter(Sighting_Year == c(2014, 2015, 2016, 2017)) %>%
  ggplot(aes(x = month_cat, fill = month_cat)) +
  geom_bar(position=position_dodge(preserve = "single"), stat='count') +
  geom_text(stat='count', aes(label=..count..), vjust=-1) +
  scale_fill_manual(values = c("#882E72", "#5289C7", "#CAE0AB", 
                             "#E8601C", "#DC050C", "#D6C1DE", "#4477AA"),
                    labels = c("less than 1 month", "less than 2 month",
                               "less than 3 month", "less than 4 month",
                               "less than 5 month", "less than half year",
                               "more than half year")) +
  ylim(0, 3000) +
  labs(x = "Time to Solve", y = "Cases Count", 
       title = "Time to Solve Rat Sighting", fill = "Month") +
  facet_grid(.~Sighting_Year) +
  theme(axis.text.x = element_text(angle = 90, 
                                   hjust = 1, vjust = 0),
        plot.title = element_text(hjust=0.5))
```


**by percentage**
```{r}
solved_duration.1 <- solved_duration %>%
  group_by(Sighting_Year) %>%
  count(month_cat) %>%
  mutate(mon_solve_per = round(100*(n / sum(n)),1))

df <- data.frame(Sighting_Year = as.factor(2017), 
                 month_cat = as.factor("more than half year"), 
                 n = 0, mon_solve_per = 0)

solved_dur <- data.frame(Sighting_Year = solved_duration.1$Sighting_Year,
                         month_cat = solved_duration.1$month_cat,
                         n = solved_duration.1$n,
                         mon_solve_per = solved_duration.1$mon_solve_per)

solved_dur <- rbind(solved_dur, df)
```

```{r}
library(scales)
ggplot(solved_duration, aes(x= month_cat,  group=Sighting_Year)) + 
    geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
    geom_text(aes(label = scales::percent(..prop..),
                   y= ..prop.. ), stat= "count", vjust = -.5) +
    labs(y = "Percent", fill="test2") +
    facet_grid(~Sighting_Year) +
    scale_y_continuous(labels=percent)



solved_duration.1 %>%
  filter(Sighting_Year == c(2014, 2015, 2016, 2017)) %>%
  ggplot(aes(x = month_cat, y = mon_solve_per)) +
  geom_bar(stat="identity") +
  facet_grid(. ~ Sighting_Year) +
  scale_y_continuous(labels=percent, limits=c(0,0.9)) + 
  geom_text(aes(label=paste0((mon_solve_per),"%")), size=4)












ggplot(test.pct, aes(x=test2, y=pct, colour=test2, fill=test2)) +
  geom_bar(stat="identity") +
  facet_grid(. ~ test1) +
  scale_y_continuous(labels=percent, limits=c(0,0.27)) + 
  geom_text(data=test.pct, aes(label=paste0(round(pct*100,1),"%"),
                               y=pct+0.012), size=4)




solved_dur %>%
  filter(Sighting_Year == c(2014, 2015, 2016, 2017)) %>%
  ggplot(aes(x = month_cat, y = mon_solve_per, fill = month_cat)) +
  geom_bar(position=position_dodge(preserve = "single"), stat='identity') +
  geom_text(aes(y = mon_solve_per,label = paste0((mon_solve_per),"%")),
            vjust=-1) +
  scale_fill_manual(values = c("#882E72", "#5289C7", "#CAE0AB", 
                               "#E8601C", "#DC050C", "#D6C1DE", "#4477AA"),
                    labels = c("less than 1 month", "less than 2 month",
                               "less than 3 month", "less than 4 month",
                               "less than 5 month", "less than half year",
                               "more than half year")) +
  labs(x = "Time to Solve", y = "Cases Count", 
       title = "Time to Solve Rat Sighting", fill = "Month") +
  facet_grid(.~Sighting_Year) +
  ylim(0, 100) +
  theme(axis.text.x = element_text(angle = 90, 
                                   hjust = 1, vjust = 0),
        plot.title = element_text(hjust=0.5))
```








## Is the number of rat sightings in a given borough proportional to its population?

Quick side note: 2008 and 2009 have been ignored because they only have 18 and 1 rat sightings respectively. Also, when looking up the data; some borough have different names as counties, but exactly the same area, so one should look up the county names instead. 



```{r}
Borough_Rats <- 
  rat_dataset %>% 
  group_by(Borough, Due_Year) %>% 
  tally(sort = TRUE) %>% 
  pivot_wider(names_from = Due_Year, values_from = n) %>% 
  select(Borough, "2010", "2011", "2012", "2013", "2014", "2015", "2016", "2017") %>% 
  drop_na()

head(Borough_Rats)


Borough_Population <- 
  read_csv("New_York_City_Population_by_Borough__1950_-_2040.csv", show_col_types = FALSE) %>% 
  select("Borough", "Area", "2010", "2011", "2012", "2013", "2014", "2015", "2016", "2017") %>% 
  mutate(Borough = toupper(Borough)) %>% 
  inner_join(Borough_Rats, by = "Borough") %>% 
  rename(Borough = "Borough", Area = "Area", Population_2010 = "2010.x", Rat_Sightings_2010 = "2010.y", Population_2011 = "2011.x", Rat_Sightings_2011 = "2011.y", Population_2012 = "2012.x", Rat_Sightings_2012 = "2012.y", Population_2013 = "2013.x", Rat_Sightings_2013 = "2013.y", Population_2014 = "2014.x", Rat_Sightings_2014 = "2014.y", Population_2015 = "2015.x", Rat_Sightings_2015 = "2015.y", Population_2016 = "2016.x", Rat_Sightings_2016 = "2016.y", Population_2017 = "2017.x", Rat_Sightings_2017 = "2017.y") %>%
  relocate(Borough, Area, Population_2010, Rat_Sightings_2010, Population_2011, Rat_Sightings_2011, Population_2012, Rat_Sightings_2012, Population_2013, Rat_Sightings_2013, Population_2014, Rat_Sightings_2014, Population_2015, Rat_Sightings_2015, Population_2016, Rat_Sightings_2016, Population_2017, Rat_Sightings_2017) #%>% 
  #mutate(Sightings_Rate = round(((Rat_Sightings / Population) * 100000), digits = 2))
  

head(Borough_Population)
```

##Creating two new databases; one for rats per population and one for rats per area

```{r}
Rats_Per_Pop <-
  Borough_Population %>% 
  mutate('2010' = round(((Rat_Sightings_2010 / Population_2010) * 100000), digits = 2), '2011' = round(((Rat_Sightings_2011 / Population_2011) * 100000), digits = 2), '2012' = round(((Rat_Sightings_2012 / Population_2012) * 100000), digits = 2), '2013' = round(((Rat_Sightings_2013 / Population_2013) * 100000), digits = 2), '2014' = round(((Rat_Sightings_2014 / Population_2014) * 100000), digits = 2), '2015' = round(((Rat_Sightings_2015 / Population_2015) * 100000), digits = 2), '2016' = round(((Rat_Sightings_2016 / Population_2016) * 100000), digits = 2), '2017' = round(((Rat_Sightings_2017 / Population_2017) * 100000), digits = 2)) %>% 
  select(Borough, '2010', '2011', '2012', '2013', '2014', '2015', '2016', '2017')

head(Rats_Per_Pop)

Rats_Per_Km <-
  Borough_Population %>% 
  mutate('2010' = round((Rat_Sightings_2010 / Area), digits = 2), '2011' = round((Rat_Sightings_2011 / Area), digits = 2), '2012' = round((Rat_Sightings_2012 / Area), digits = 2), '2013' = round((Rat_Sightings_2013 / Area), digits = 2), '2014' = round((Rat_Sightings_2014 / Area), digits = 2), '2015' = round((Rat_Sightings_2015 / Area), digits = 2), '2016' = round((Rat_Sightings_2016 / Area), digits = 2), '2017' = round((Rat_Sightings_2017 / Area), digits = 2)) %>% 
  select(Borough, '2010', '2011', '2012', '2013', '2014', '2015', '2016', '2017')

head(Rats_Per_Km)
```

##Summary statistics for above

```{r}

```

##Visualization for above

```{r}
P_Rates <-
  Rats_Per_Pop %>% 
  pivot_longer(c('2010', '2011', '2012', '2013', '2014', '2015', '2016', '2017'), names_to = "Year", values_to = "Rate") 

as.factor(P_Rates$Year)

head(P_Rates, n = 20)

P_Rates %>% 
  ggplot(aes(x=Year, y=Rate, color = Borough, group = Borough)) +
  geom_point() +
  geom_line() +
  labs(x = "Year", y = "Sightings per 100.000 population", title = "Rate of sightings by Borough population", color = "Borough") +
  theme_classic() +
  theme(plot.title = element_text(hjust=0.5))


P_Rates %>% 
  ggplot(aes(x=Year, y=Rate, group = Borough)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ Borough) + 
  labs(x = "Year", y = "Sightings per 100.000 population", title = "Rate of sightings by Borough population") +
  theme_classic() +
  theme(plot.title = element_text(hjust=0.5))

A_Rates <-
  Rats_Per_Km %>% 
  pivot_longer(c('2010', '2011', '2012', '2013', '2014', '2015', '2016', '2017'), names_to = "Year", values_to = "Rate") 

as.factor(A_Rates$Year)

head(A_Rates, n = 20)

A_Rates %>% 
  ggplot(aes(x=Year, y=Rate, color = Borough, group = Borough)) +
  geom_point() +
  geom_line() +
  labs(x = "Year", y = "Sightings per square kilometre", title = "Rate of sightings by Borough area", color = "Borough") +
  theme_classic() +
  theme(plot.title = element_text(hjust=0.5))

A_Rates %>% 
  ggplot(aes(x=Year, y=Rate, group = Borough)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ Borough) +
  labs(x = "Year", y = "Sightings per square kilometre", title = "Rate of sightings by Borough area", color = "Borough") +
  theme_classic() +
  theme(plot.title = element_text(hjust=0.5))

```


