---
title: "SLV_practical8"
author: "Amalia Tsakali"
date: "2022-11-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(magrittr)
library(psych)
library(caret)
library(gbm)
library(xgboost)
library(data.table)
library(ggforce)
```
```{r}
set.seed(45)
df <- readRDS("data/train_disease.RDS")
```
1. Get an impression of the data by looking at the structure of the data and creating some descriptive statistics.
```{r 1}
head(df)
summary(df)
df %>%
  select(-c(Gender,Disease)) %>%
  describeBy(df$Disease,fast=TRUE)
```
2. To further explore the data we work with, create some interesting data visualizations that show whether there are interesting patterns in the data.
```{r 2}
head(df)
df %>% ggplot(aes(x=Gender, fill=Disease))+geom_bar(stat = "count")
#our dataset has more male patients, but the proportion of women that are patients is higher

df %>%
  select(-Gender) %>%
  pivot_longer(where(is.numeric)) %>%
  ggplot(aes(x = value, col = Disease, fill = Disease)) +
  geom_density(alpha = 0.8) +
  facet_wrap(~name, scales = "free") +
  theme_minimal()
#there seem to be some slight differences in the distributions of th contiuous variables between healthy and diseased individuals 
```
3. Shortly reflect on the difference between bagging, random forests, and boosting.
```{r 3}
#bagging
#random forests
#boosting

# we can first of all divide these in two groups, with boosting being the odd one out. 
#boosting is done in steps, while bagging and random forests are done on parallel. 
#Boosting also uses all the data , while bagging and random forests take a bootstrap sample 
#(the out of bag used for validation)

#The main difference betweeen bagging and random forests is that random forests have a "handicap" on
#how many features the trees are trained on
```

4. Apply bagging to the training data, to predict the outcome Disease, using the caret library.

```{r 4}
cvcontrol <- trainControl(method = "repeatedcv", 
                          number = 10,
                          allowParallel = TRUE)

bag_train <- train(Disease ~ .,
                   data = df, 
                   method = 'treebag',
                   trControl = cvcontrol,
                   importance = TRUE)
```
5. Interpret the variable importance measure using the varImp function on the trained model object.
```{r 5}
plot(varImp(bag_train))

#we see the features that have the biggest importance for the model on the top
```
6. Create training set predictions based on the bagged model, and use the confusionMatrix() function from the caret package to assess it’s performance.`
```{r 6}
confusionMatrix(predict(bag_train, type = "raw"),
                df$Disease)
  
#the confusion matrix shows that our model predicts perfectly on our training data
```
7. Now ask for the output of the bagged model. Explain why the under both approaches differ.
```{r 7}
bag_train
```
8. Fit a random forest to the training data to predict the outcome Disease, using the caret library.
```{r 8}
rf_train <- train(Disease ~ .,
                   data = df, 
                   method = 'rf',
                   trControl = cvcontrol,
                   importance = TRUE)
```
9. Again, interpret the variable importance measure using the varImp function on the trained model object. Do you draw the same conclusions as under the bagged model?
```{r 9}
plot(varImp(rf_train))

#this model has different importance for each feature 
```
10. Output the model output from the random forest. Are we doing better than with the bagged model?
```{r 10}
rf_train

#we see a slightly higher accuracy
```
11. Now, fit a boosting model using the caret library to predict disease status.`
```{r 11}
boosting_train<-train(Disease ~ .,
                   data = df, 
                   method = 'gbm',
                   trControl = cvcontrol,
                   verbose=F)
```
12. Again, interpret the variable importance measure. You will have to call for summary() on the model object you just created. Compare the output to the previously obtained variable importance measures.
```{r 12}
plot(varImp(boosting_train))
summary(boosting_train)

#again different importances
```
13. Output the model output from our gradient boosting procedure. Are we doing better than with the bagged and random forest model?
```{r 13}
boosting_train
#the best model has a slighly bettter accuracy
```
14. Download the file shap.R from this Github repository.
```{r 14}
library(devtools)
source_url("https://github.com/pablo14/shap-values/blob/master/shap.R?raw=TRUE")
```
15. Specify your model as follows, and use it to create predictions on the training data.
```{r 15}
train_x <- model.matrix(Disease ~ ., df)[,-1]
train_y <- as.numeric(df$Disease) - 1
xgboost_train <- xgboost(data = train_x,
                         label = train_y, 
                         max.depth = 10,
                         eta = 1,
                         nthread = 4,
                         nrounds = 4,
                         objective = "binary:logistic",
                         verbose = 2)



pred <- tibble(Disease = predict(xgboost_train, newdata = train_x)) %>%
  mutate(Disease = factor(ifelse(Disease < 0.5, 1, 2),
                          labels = c("Healthy", "Disease")))

table(pred$Disease, df$Disease)
```
16. First, calculate the SHAP rank scores for all variables in the data, and create a variable importance plot using these values. Interpret the plot.
```{r 16}
shap_results <- shap.score.rank(xgboost_train,
                                X_train = train_x,
                                shap_approx = F)

plot(var_importance(shap_results))

```
17. Plot the SHAP values for every individual for every feature and interpret them.
```{r 17}
shap_long <- shap.prep(shap = shap_results,
                       X_train = train_x)

plot.shap.summary(shap_long)

xgb.plot.shap(train_x, features = colnames(train_x), model = xgboost_train, n_col = 3)

#1st plot
#High value for Direct Birubulin decreases the probability of being diseased. 
#same for age

#2nd plot gives the same information in greater detail for every variable
```
18. Verify which of the models you created in this practical performs best on the test data.
```{r 18}
test <- readRDS("data/test_disease.RDS")

bag_test <- predict(bag_train, newdata = test)
rf_test  <- predict(rf_train, newdata = test)
boosting_test <- predict(boosting_train, newdata = test)
xgboost_test <- predict(xgboost_train, newdata = model.matrix(Disease ~ ., test)[,-1]) %>%
  factor(x = ifelse(. < 0.5, 1, 2), levels = c(1,2), labels = c("Healthy", "Disease"))

confusionMatrix(bag_test, test$Disease)
confusionMatrix(rf_test, test$Disease)
confusionMatrix(boosting_test, test$Disease)
confusionMatrix(xgboost_test, test$Disease)

#the random forest has the best performance
```

