---
title: "Practical 7"
author: "Judith Neve"
date: '2022-10-28'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS)
library(splines)
library(ISLR)
library(tidyverse)

set.seed(45)
```

# Prediction plot

```{r}
Boston %>% 
  ggplot(aes(x = lstat, y = medv)) +
  geom_point() +
  theme_minimal()
```

## 1. Create a function called pred_plot() that takes as input an lm object, which outputs the above plot but with a prediction line generated from the model object using the predict() method.

```{r}
pred_plot <- function(model) {
  x_pred <- seq(min(Boston$lstat),
                max(Boston$lstat),
                length.out = 500)
  y_pred <- predict(model,
                    newdata = tibble("lstat" = x_pred))
  p <- Boston %>% 
    ggplot(aes(x = lstat, y = medv)) +
    geom_point() +
    geom_line(data = data.frame("lstat" = x_pred,
                                "medv" = y_pred)) +
    theme_minimal()
  
  print(p)
}
```

## 2. Create a linear regression object called lin_mod which models medv as a function of lstat. Check if your prediction plot works by running pred_plot(lin_mod). Do you see anything out of the ordinary with the predictions?

```{r}
lin_mod <- lm(medv ~ lstat, data = Boston)

pred_plot(lin_mod)
# much steeper decrease for lower values
```

# Polynomial regression

## 3. Create another linear model pn3_mod, where you add the second and third-degree polynomial terms I(lstat^2) and I(lstat^3) to the formula. Create a pred_plot() with this model.

```{r}
pn3_mod <- lm(medv ~ lstat + I(lstat^2) + I(lstat^3), data = Boston)

pred_plot(pn3_mod)
```

## 4. Play around with the poly() function. What output does it generate with the arguments degree = 3 and raw = TRUE?

```{r}
head(poly(Boston$lstat, degree = 3))
```

```{r}
head(poly(Boston$lstat, degree = 3, raw = TRUE))
```

## 5. Use the poly() function directly in the model formula to create a 3rd-degree polynomial regression predicting medv using lstat. Compare the prediction plot to the previous prediction plot you made. What happens if you change the poly() function to raw = FALSE?

```{r}
poly3_mod <- lm(medv ~ poly(lstat, degree = 3), data = Boston)

pred_plot(poly3_mod)
```

# Piecewise regression

## 6. Create a model called pw2_mod with one predictor: I(lstat <= median(lstat)). Create a pred_plot with this model. Use the coefficients in coef(pw2_mod) to find out what the predicted value for a low-lstat neighbourhood is.

```{r}
pw2_mod <- lm(medv ~ I(lstat <= median(lstat)),
              data = Boston)

pred_plot(pw2_mod)
```

```{r}
coef(pw2_mod)
```

## 7. Use the cut() function in the formula to generate a piecewise regression model called pw5_mod that contains 5 equally spaced sections. Again, plot the result using pred_plot.

```{r}
pw5_mod <- lm(medv ~ I(cut(lstat, 5)),
              data = Boston)

pred_plot(pw5_mod)
```

## 8. Optional: Create a piecewise regression model pwq_mod where the sections are not equally spaced, but have equal amounts of training data. Hint: use the quantile() function.

```{r}
lims <- quantile(Boston$lstat)

pwq_mod <- lm(medv ~ I(cut(lstat, lims)),
              data = Boston)

pred_plot(pwq_mod)
```

# Piecewise polynomial regression

## 9. This function does not have comments. Copy - paste the function and add comments to each line. To figure out what each line does, you can first create “fake” vec and knots variables, for example vec <- 1:20 and knots <- 2 and try out the lines separately.

```{r}
piecewise_cubic_basis <- function(vec, knots = 1) { # input is a vector and a number of knots
  if (knots == 0) return(poly(vec, degree = 3, raw = TRUE)) # if there are no knots we return a third degree polynomial of the vector
  
  cut_vec <- cut(vec, breaks = knots + 1) # we cut the vector into multiple pieces, for the number of knots + 1
  
  out <- matrix(nrow = length(vec), ncol = 0) # create a matrix which will be filled with the design matrix
  
  for (lvl in levels(cut_vec)) { # for each section of the cut vector
    tmp <- vec
    tmp[cut_vec != lvl] <- 0 # we assign 0 to the parts of the vector that are not the given section
    out <- cbind(out, poly(tmp, degree = 3, raw = TRUE)) # we add a set of columns to the design matrix, which are a polynomial of the section of interest
  }
  
  out
}
```

## 10. Create piecewise cubic models with 1, 2, and 3 knots (pc1_mod - pc3_mod) using this piecewise cubic basis function. Compare them using the pred_plot() function.

```{r}
pc1_mod <- lm(medv ~ piecewise_cubic_basis(lstat, 1), data = Boston)

pred_plot(pc1_mod)
```

```{r}
pc2_mod <- lm(medv ~ piecewise_cubic_basis(lstat, 2), data = Boston)

pred_plot(pc2_mod)
```

```{r}
pc3_mod <- lm(medv ~ piecewise_cubic_basis(lstat, 3), data = Boston)

pred_plot(pc3_mod)
```

# Splines

## 11. Create a data frame called boston_tpb with the columns medv and lstat from the Boston dataset.

```{r}
boston_tbp <- data.frame(medv = Boston$medv,
                         lstat = Boston$lstat)
```

## 12. Now use mutate to add squared and cubed versions of the lstat variable to this dataset.

```{r}
boston_tbp <- boston_tbp %>% 
  mutate(lstat_sq = lstat^2,
         lstat_cb = lstat^3)
```

## 13. Use mutate to add a column lstat_tpb to this dataset which is 0 below the median and has value (lstat - median(lstat))^3 above the median. Tip: you may want to use ifelse() within your mutate() call.

```{r}
boston_tbp <- boston_tbp %>% 
  mutate(lstat_tpb = ifelse(lstat < median(lstat),
                            0,
                            (lstat - median(lstat))^3))
```

## 14. Create a linear model tpb_mod using the lm() function. How many predictors are in the model? How many degrees of freedom does this model have?

```{r}
tpb_mod <- lm(medv ~ ., data = boston_tbp)
summary(tpb_mod)
```

## 15. Create a cubic spline model bs1_mod with a knot at the median using the bs() function. Compare its predictions to those of the tpb_mod using the predict() function on both models.

```{r}
bs1_mod <- lm(medv ~ bs(lstat, knots = median(lstat)),
              data = Boston)
```

```{r}
predict(tpb_mod) - predict(bs1_mod)
```

```{r}
pred_plot(bs1_mod)
```

## 17. Create a natural cubic spline model (ns3_mod) with 3 degrees of freedom using the ns() function. Plot it, and compare it to the bs1_mod.

```{r}
ns3_mod <- lm(medv ~ ns(lstat, df = 3),
              data = Boston)

pred_plot(ns3_mod)
```

## 18. Plot lin_mod, pn3_mod, pw5_mod, pc3_mod, bs1_mod, and ns3_mod and give them nice titles by adding + ggtitle("My title") to the plot. You may use the function plot_grid() from the package cowplot to put your plots in a grid.

```{r}
library(cowplot)

p1 <- pred_plot(lin_mod) + ggtitle("Linear model")
p2 <- pred_plot(pn3_mod) + ggtitle("Degree 3 polynomial model")
p3 <- pred_plot(pw5_mod) + ggtitle("Piecewise model with 5 pieces")
p4 <- pred_plot(pc3_mod) + ggtitle("Piecewise cubic model with 3 pieces")
p5 <- pred_plot(bs1_mod) + ggtitle("Basic spline model with 1 knot")
p6 <- pred_plot(ns3_mod) + ggtitle("Natural spline model with three DF")

plot_grid(plotlist = list(p1, p2, p3, p4, p5, p6))
```

