---
title: "Pratical 2"
subtitle: "Supervised Learning and Visualization"
author: "Ana Martins"
date: "September 2022"
output: html_document
---

**Libraries**:

```{r}
library(ISLR)
library(tidyverse)
```

## What is `ggplot`?

1. **Name the aesthetics, geoms, scales, and facets of the above visualisation. Also name any statistical transformations or special coordinate systems.**

Specifications of the "Cool density and scatter plot of baseball data" plot:

- Aesthetics: none defined by the user, the plot uses the default
- Geoms: the points geoms is used, meaning the data is displayed in a scatter plot, as well as the 2d density, which displays the data according to density curves
- Scales: the scales are defined automatically
- Facets: there are no facets, the whole data is displayed in one place
- Other: the labels for the axis and the title are changed, as well as the theme, which is set to minimal

## Aesthetics and data preparation

2. **Run the code below to generate data. There will be three vectors in your environment. Put them in a data frame for entering it in a `ggplot()` call using either the `data.frame()` or the `tibble()` function. Give informative names and make sure the types are correct (use the `as.<type>()` functions). Name the result `gg_students`**

```{r}
set.seed(1234)
student_grade  <- rnorm(32, 7)
student_number <- round(runif(32) * 2e6 + 5e6)
programme      <- sample(c("Science", "Social Science"), 32, replace = TRUE)
```

```{r}
gg_students <- data.frame(student_grade, student_number, programme)
gg_students$programme <-  as.factor(programme)
gg_students
```

The type of `programme` was assumed to be `character`. We know that there is a finite number of values in `programme`, so we can set it as `factor`.

3. **Plot the first `homeruns_plot` again, but map the `Hits` to the y-axis and the `HmRun` to the x-axis instead.**

```{r}
homeruns_plot <- 
  ggplot(Hitters, aes(x = HmRun, y = Hits)) +
  geom_point() +
  labs(x = "Home runs", y = "Hits")
homeruns_plot
```

4. **Recreate the same plot once more, but now also map the variable `League` to the `colour` aesthetic and the variable `Salary` to the size aesthetic.**

```{r}
homeruns_plot <- 
  ggplot(Hitters, aes(x = HmRun, y = Hits, colour = League, size = Salary)) +
  geom_point() +
  labs(x = "Home runs", y = "Hits")
homeruns_plot
```

## Visual exploratory data analysis

### Histogram

6. **Use `geom_histogram()` to create a histogram of the grades of the students in the `gg_students` dataset. Play around with the `binwidth` argument of the `geom_histogram()` function.**

```{r}
histogram <-
  ggplot(gg_students, aes(x = student_grade)) +
  geom_histogram()
histogram
histogram <-
  ggplot(gg_students, aes(x = student_grade)) +
  geom_histogram(binwidth = 1)
histogram
histogram <-
  ggplot(gg_students, aes(x = student_grade)) +
  geom_histogram(binwidth = 0.5, boundary = 0)
histogram
```

### Density

7. **Use `geom_density()` to create a density plot of the grades of the students in the `gg_students` dataset. Add the argument `fill = "light seagreen"` to `geom_density()`.**

```{r}
density <-
  ggplot(gg_students, aes(student_grade)) +
  geom_density(fill = "light seagreen")
density
```

8. **Add rug marks to the density plot through `geom_rug()`. You can edit the `colour` and `size` of the rug marks using those arguments within the `geom_rug()` function.**

```{r}
density <- 
  density +
  geom_rug(color = "Coral", size = 1)
density
```

9. **Increase the data to ink ratio by removing the y axis label, setting the theme to `theme_minimal()`, and removing the border of the density polygon. Also set the limits of the x-axis to go from 0 to 10 using the `xlim()` function, because those are the plausible values for a student grade.**

Here, we have to redefine density as geoms overlay, so we would never be able to "take out" the border of the density polygon.

```{r}
density <-
  ggplot(gg_students, aes(student_grade)) +
  geom_density(fill = "light seagreen", color = NA) +
  geom_rug(color = "Coral", size = 1) +
  ylab(element_blank()) +
  theme_minimal() +
  xlim(0, 10)
density
```

### Boxplot

10. **Create a boxplot of student grades per programme in the `gg_students` dataset you made earlier: map the programme variable to the x position and the grade to the y position. For extra visual aid, you can additionally map the programme variable to the fill aesthetic.**

```{r}
bxplt <- 
  ggplot(gg_students, aes(x = programme, y = student_grade, fill = programme)) +
  geom_boxplot()
bxplt
```

11. **What do each of the horizontal lines in the boxplot mean? What do the vertical lines (whiskers) mean?**

The lower and upper horizontal lines (hinges) correspond to the first and third quartiles and the central one corresponds to the median. The vertical lines (whiskers) go from the minimum to the maximum values (excluding outliers, which are shown as points).

### Two densities

12 **Comparison of distributions across categories can also be done by adding a fill aesthetic to the density plot you made earlier. Try this out. To take care of the overlap, you might want to add some transparency in the geom_density() function using the alpha argument.**

```{r}
density <-
  ggplot(gg_students, aes(student_grade, fill = programme)) +
  geom_density(color = NA, alpha = 0.5) +
  geom_rug() +
  ylab(element_blank()) +
  theme_minimal() +
  xlim(0, 10)
density
```

### Bar plot

13. **Create a bar plot of the variable Years from the Hitters dataset.**

```{r}
brplt <- 
  ggplot(Hitters, aes(Years)) +
  geom_bar()
brplt
```

### Line plot

14. **Use `geom_line()` to make a line plot out of the first 200 observations of the variable `Volume` (the number of trades made on each day) of the `Smarket` dataset. You will need to create a `Day` variable using `mutate()` to map to the x-position. This variable can simply be the integers from 1 to 200. Remember, you can select the first 200 rows using `Smarket[1:200, ]`.**

```{r}
nSmarket <- 
  Smarket %>% 
  mutate(Day = c(1:1250))
ln <- 
  ggplot(nSmarket[1:200, ], aes(x = Day, y = Volume)) +
  geom_line()
ln
```

15. **Give the line a nice `colour` and increase its `size`. Also add points of the same colour on top.**

```{r}
ln <- 
  ggplot(nSmarket[1:200, ], aes(x = Day, y = Volume)) +
  geom_line(colour = "Antique White 4", size = 1) + 
  geom_point(colour = "Antique White 4")
ln
```

16. **Use the function `which.max()` to find out which of the first 200 days has the highest trade volume and use the function `max()` to find out how large this volume was**

```{r}
day <- which.max(nSmarket[1:200, ]$Volume)
volume <- max(nSmarket[1:200, ]$Volume)
```

17. **Use `geom_label(aes(x = your_x, y = your_y, label = "Peak volume"))` to add a label to this day. You can use either the values or call the functions. Place the label near the peak!**

```{r}
ln <-
  ln +
  geom_label(aes(x = day, y = volume, label = "Peak volume"))
ln
```

## Faceting

18. **Create a data frame called `baseball` based on the `Hitters` dataset. In this data frame, create a factor variable which splits players’ salary range into 3 categories. Tip: use the `filter()` function to remove the missing values, and then use the `cut()` function and assign nice `labels` to the categories. In addition, create a variable which indicates the proportion of career hits that was a home run.**

```{r}
baseball <-
  Hitters %>%
  filter(!is.na(Hitters$Salary))

baseball <-
  baseball %>% 
  mutate(SalaryTiers = cut(baseball$Salary, breaks=c(0, 425, 750, 2460), labels = c("Rich", "Very Rich", "Extremely Rich"))) %>% 
  mutate(HmRunProportion = (baseball$HmRun / baseball$Hits))

baseball
```

19. **Create a scatter plot where you map `CWalks` to the x position and the proportion you calculated in the previous exercise to the y position. Fix the y axis limits to (0, 0.4) and the x axis to (0, 1600) using `ylim()` and `xlim()`. Add nice x and y axis titles using the `labs() function`. Save the plot as the variable `baseball_plot`.**

```{r}
baseball_plot <-
  baseball %>% 
  ggplot(aes(x = CWalks, y = HmRunProportion)) +
  geom_point() +
  ylim(0, 0.4) + 
  xlim(0, 1600) +
  labs(x = "CWalks", y = "Home Runs Proportion")
baseball_plot
```

20. **Split up this plot into three parts based on the salary range variable you calculated. Use the `facet_wrap()` function for this; look at the examples in the help file for tips.**

```{r}
baseball_plot <-
  baseball_plot +
  facet_wrap(baseball$SalaryTiers)
baseball_plot
```

## Final Exercise

21. **Create an interesting data visualisation based on the `Carseats` data from the `ISLR` package.**

```{r}
library(ISLR)
```

Is the Advertising related to the Price? Or is it more related to the Population?

```{r}
adv <-
  Carseats %>% 
  ggplot(aes(x = Advertising, y = Price, color = Population)) +
  geom_point()
adv
```

It does seem that the ones with most Advertising are in places with higher Population, and are sold for a higher Price. However, the rest of the plot is not very interesting.

Let us compare the Company's Prices with the Competitor's Prices:

```{r}
prices <-
  Carseats %>% 
  ggplot(aes(x = CompPrice, y = Price)) +
  geom_point()
prices
```

It does seem that there is a loosely linear relation between the two.

Lastly, something that might be interesting to look at is how the Education level affects how much the company uses Advertising in a certain location. We would expect that the higher the Education, the lower the Advertising, so we are expecting an inversely proportional relation.

```{r}
edu_adv <- 
  Carseats %>%
  ggplot(aes(x = Advertising, y = Education)) +
  geom_point()
edu_adv
```

The data seems to have almost no correlation, unlike what we expected, so the Carseat company should maybe think about changing their strategy...