---
title: "Pratical 2"
subtitle: "Supervised Learning and Visualization"
author: "Ana Martins"
date: "September 2022"
output: html_document
---

**Libraries**:

```{r}
library(ISLR)
library(tidyverse)
```

## What is `ggplot`?

1. **Name the aesthetics, geoms, scales, and facets of the above visualisation. Also name any statistical transformations or special coordinate systems.**

Specifications of the "Cool density and scatter plot of baseball data" plot:

- Aesthetics: none defined by the user, the plot uses the default
- Geoms: the points geoms is used, meaning the data is displayed in a scatter plot, as well as the 2d density, which displays the data according to density curves
- Scales: the scales are defined automatically
- Facets: there are no facets, the whole data is displayed in one place
- Other: the labels for the axis and the title are changed, as well as the theme, which is set to minimal

## Aesthetics and data preparation

2. **Run the code below to generate data. There will be three vectors in your environment. Put them in a data frame for entering it in a `ggplot()` call using either the `data.frame()` or the `tibble()` function. Give informative names and make sure the types are correct (use the `as.<type>()` functions). Name the result `gg_students`**

```{r}
set.seed(1234)
student_grade  <- rnorm(32, 7)
student_number <- round(runif(32) * 2e6 + 5e6)
programme      <- sample(c("Science", "Social Science"), 32, replace = TRUE)
```

```{r}
gg_students <- data.frame(student_grade, student_number, programme)
gg_students$programme <-  as.factor(programme)
gg_students
```

The type of `programme` was assumed to be `character`. We know that there is a finite number of values in `programme`, so we can set it as `factor`.

3. **Plot the first `homeruns_plot` again, but map the `Hits` to the y-axis and the `HmRun` to the x-axis instead.**

```{r}
homeruns_plot <- 
  ggplot(Hitters, aes(x = HmRun, y = Hits)) +
  geom_point() +
  labs(x = "Home runs", y = "Hits")
homeruns_plot
```

4. **Recreate the same plot once more, but now also map the variable `League` to the `colour` aesthetic and the variable `Salary` to the size aesthetic.**

```{r}
homeruns_plot <- 
  ggplot(Hitters, aes(x = HmRun, y = Hits, colour = League, size = Salary)) +
  geom_point() +
  labs(x = "Home runs", y = "Hits")
homeruns_plot
```

## Histogram

6. **Use `geom_histogram()` to create a histogram of the grades of the students in the `gg_students` dataset. Play around with the `binwidth` argument of the `geom_histogram()` function.**

```{r}
histogram <-
  ggplot(gg_students, aes(x = student_grade)) +
  geom_histogram()
histogram
histogram <-
  ggplot(gg_students, aes(x = student_grade)) +
  geom_histogram(binwidth = 1)
histogram
histogram <-
  ggplot(gg_students, aes(x = student_grade)) +
  geom_histogram(binwidth = 0.5, boundary = 0)
histogram
```

## Density

7. **Use `geom_density()` to create a density plot of the grades of the students in the `gg_students` dataset. Add the argument `fill = "light seagreen"` to `geom_density()`.**

```{r}
?geom_density
density <-
  ggplot(gg_students, aes(student_grade)) +
  geom_density(fill = "light seagreen")
density
```

8. **Add rug marks to the density plot through `geom_rug()`. You can edit the `colour` and `size` of the rug marks using those arguments within the `geom_rug()` function.**

```{r}
density <- density +
  geom_rug()
density
```

