---
title: "Classification 1"
author: "Ana Martins"
date: "October 2022"
output: html_document
---

## Introduction

```{r}
library(MASS)
library(class)
library(ISLR)
library(tidyverse)
```

```{r}
set.seed(45)
```

## Default dataset

1. **Create a scatterplot of the `Default` dataset, where `balance` is mapped to the x position, `income` is mapped to the y position, and `default` is mapped to the colour. Can you see any interesting patterns already?**

```{r}
default.plot <-
  Default %>% 
  ggplot(mapping = aes(x = balance, y = income, color = default)) +
  geom_point()
default.plot
```

The default Nos clearly appear more for lower balances and the default Yeses on the higher balances, but there seems to be no or low correlation to the income.

2. **Add facet_grid(cols = vars(student)) to the plot. What do you see?**

```{r}
default.plot <-
  default.plot +
  facet_grid(cols = vars(student))
default.plot
```

As expected, the students have a lower income but they also have about the same balance.

3. **Transform “student” into a dummy variable using `ifelse()` (0 = not a student, 1 = student). Then, randomly split the Default dataset into a training set `default_train` (80%) and a test set `default_test` (20%)**

```{r}
default.data <-
  Default %>% 
  mutate(student = ifelse(student == "Yes", 1, 0)) %>% 
  mutate(split = sample(rep(c("train", "test"), times = c(8000, 2000))))
default.data

default.train <-
  default.data %>% 
  filter(split == "train")
```


## K-Nearest Neighbours

## Confusion matrix

## Logistic regression

## Linear discriminant analysis

## Final assignment