---
title: "Supervised Learning and Visualization Practical 3"
author: "Daniel Anadria"
date: 04 October 2022
output:
  html_document:
    css: Daniel_03_EDA.css
    
---
<p style="text-align: center;">**Practical 3: Exploratory Data Analysis**</p>

```{r, results='hide', message=FALSE, warning=FALSE}
library(tidyverse)
library(magrittr)
library(mice)
library(DAAG)
```

```{r}
boys %>% head() 
```

```{r}
boys %>% tail()
```

```{r}
!is.unsorted(boys$age)
```

```{r}
summary(boys)
```

1. Create a histogram of the variable age using the function geom_histogram().

```{r}
ggplot(boys, aes(x = age))+
  geom_histogram(bins = 100, color = "dark blue")+
  labs(title = "Histogram of Boys Age", x = "Age", y = "Count")+
  theme_minimal()
```
2. Create a bar chart of the variable gen using the function geom_bar().

```{r}
ggplot(boys, aes(x = gen))+
  geom_bar(color = 'dark blue')+
  labs(title = "Bar Chart of Gen", x = "Gen", y = "Count")+
  geom_text(stat ='count', aes(label = ..count.., vjust = -0.3))+
  theme_minimal()
```
```{r}
md.pattern(boys)
```
Now we now that there is a substantial amount of missing information in the boys data, we can more closely focus on the missing data patterns. A first step could be to test whether missingness in the variables gen, phb and tv depends on someones age.

3. Create a missingness indicator for the variables gen, phb and tv.

```{r}
boys_mis <- boys %>%
  mutate(gen_mis = is.na(gen),
         phb_mis = is.na(phb),
         tv_mis  = is.na(tv))
```

4. Assess whether missingness in the variables gen, phb and tv is related to someones age.

Is gen missingness related to age? 

```{r}
boys_mis %>%
  group_by(gen_mis) %>%
  summarize(age = mean(age))
```

Is phb missingness related to age?

```{r}
boys_mis %>%
  group_by(phb_mis) %>%
  summarize(age = mean(age))
```

Is tv missingness related to age?

```{r}
boys_mis %>%
  group_by(tv_mis) %>%
  summarize(age = mean(age))
```

For all three variables, we notice that missingness is related to age, with the mean age of boys with missing data being between 6 and 8 years, and the mean age of boys with non-missing data being about 14 years. Therefore, we miss information about younger boys.

5. Create a histogram for the variable age, faceted by whether or not someone has a missing value on gen.

```{r}
ggplot(boys_mis, aes(x = age))+
  geom_histogram(bins = 100, color = "dark blue")+
  facet_grid(~gen_mis)+
  labs(title = "Missingness in Gen as a Function of Age", x = "Age", y = "Count")+
  theme_minimal()
```
We see that boys with an observed value on gen are all are at least seven years old, while those with a missing value on gen are far more often between 0 and 5 years old.

6. Create a scatterplot with age on the x-axis and bmi on the y-axis, using the function geom_point().

```{r}
ggplot(boys_mis, aes(x = age, y = bmi))+
  geom_point()+
  labs(x = "Age", y = "BMI", title = "Scatterplot of Age vs BMI")+
  theme_minimal()
```
Although we already know how missingness in the variable gen is related to age, we do not know whether gen is also related to bmi.

7. Add a colour aesthetic to the previous plot using the missingness indicator of the variable gen.

```{r}
ggplot(boys_mis, aes(x = age, y = bmi, color = gen_mis))+
  geom_point()+
  labs(x = "Age", y = "BMI", title = "Scatterplot of Age vs BMI with Gen Missingness")+
  theme_minimal()
```

We clearly see that younger boys generally have a missing value on gen, while there does not seem to be much of a relation between missingness on gen and bmi

8. Visualize the relationship between reg (region) and age using a boxplot.

```{r}
ggplot(boys, aes(x = reg, y = age))+
  geom_boxplot(fill = 'gray')+
  labs(title = "Boxplot of Age by Region.")+
  theme_minimal()
```
9. Create a density plot of age, splitting the densities by gen using the fill aesthetic.

```{r}
ggplot(boys, aes(x = age, fill = gen))+
  geom_density(alpha = 0.7)+
  labs(title = "Density of age by genital Tanner stage")+
  theme_minimal()
```
We see a clear relation between gen and age, which is not really surprising, because physical development is usually driven by aging.

10. Create a diverging bar chart for hgt in the boys data set, that displays for every age year that year’s mean height in deviations from the overall average hgt. 

```{r}
boys %>%
  mutate(Age = cut(age, 0:22, labels = paste0(0:21, " years")),
         Height = hgt - mean(hgt, na.rm = TRUE)) %>%
  group_by(Age) %>%
  summarize(Height = mean(Height, na.rm = TRUE)) %>%
  mutate(color = ifelse(Height > 0, "Above average", "Below average")) %>%
  ggplot(aes(x = Height, y = Age, fill = color)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
  theme(legend.title = element_blank())

```
11. Load the data elastic1 and elastic2 and bind the data frames together using the function bind_rows() and add a grouping variable indicating whether an observation comes from elastic1 or from elastic2.

```{r}
elastic <- bind_rows("Elastic1" = elastic1,
                     "Elastic2" = elastic2,
                     .id = "Set")
```

12. Create a scatterplot mapping stretch on the x-axis and distance on the y-axis, and map the just created group indicator as the color aesthetic.

```{r}
elastic %>%
  ggplot(aes(x = stretch, y = distance, col = Set)) +
  geom_point() +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  labs(title = "Elastic bands data")
```
13. Recreate the previous plot, but now assess whether the results of the two data sets appear consistent by adding a linear regression line.

```{r}
elastic %>%
  ggplot(aes(x = stretch, y = distance, col = Set)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  labs(title = "Elastic bands data")
```

The results seem very consistent: elastic2 has more observations over a larger range, but both sets result in roughly the same regression line. Data set elastic1 seems to have an outlier.

14. For each of the data sets elastic1 and elastic2, fit a regression model with y = distance on x = stretch using lm(y ~ x, data).

```{r}
fit1 <- lm(distance ~ stretch, elastic1)
fit2 <- lm(distance ~ stretch, elastic2)
```

15. For both of the previously created fitted models, determine the fitted values and the standard errors of the fitted values, and the proportion explained variance R2.

```{r}
fit1 %>% predict(se.fit = TRUE)
paste0("R-squared: ", round(summary(fit1)$r.squared,3))
```
```{r}
fit2 %>% predict(se.fit = TRUE)
paste0("R-squared: ", round(summary(fit2)$r.squared,3))
```

Fit1 (based on elastic1) has a larger residual standard deviation (i.e., $residual.scale).
The model based on elastic2 has smaller standard errors and a much larger R2. This is due to the larger range of values in elastic2 and the absence of outliers.

16. Study the residual versus leverage plots for both models.

A residuals vs. leverage plot is a type of diagnostic plot that allows us to identify influential observations in a regression model.

```{r}
fit1 %>% plot(which = 5)
```
We see that case 2 has the highest influence on elastic1.

```{r}
fit1$residuals
```

However, case 2 does not have the largest residual, case 7 does.

```{r}
fit2 %>% plot(which = 5)
```
Fit2 doesn't have any influential points. 
```{r}
fit2$residuals
```
The largest residual belongs to case 2.

17. Use the elastic2 variable stretch to obtain predictions on the model fitted on elastic1.

```{r}
pred <- predict(fit1, newdata = elastic2)
pred
```

18. Now make a scatterplot to investigate similarity between the predicted values and the observed values for elastic2.

```{r}
predicted_vs_observed <- cbind("predicted" = fit2$fitted.values, "observed" = elastic2$distance, stretch = elastic2$stretch) %>% as.data.frame()

ggplot(predicted_vs_observed,aes(x = stretch, y = predicted, color = "blue"))+
  geom_point(fill = "blue", size = 3)+
  geom_smooth(method = "lm", color = "blue")+ 
  labs(x = "Stretch", y = "Distance", title = "Difference Between Predicted and Observed Distance")+
  #observed values
  geom_point(aes(x = stretch, y = observed), fill = "orange", color = "dark orange", size = 3)+
  theme_minimal()+
  scale_color_manual(name = "Origin", values = c("Predicted" = "blue", "Observed" = "dark orange"))
```

The end.
